### 2024-11-18 Manuscript_figures.Rmd
Initial commit

---
title: "Manuscript figures"
author: "Rasmus Rydbirk"
date: "2024-11-11"
output: 
  html_document:
    toc: True
    toc_float: True
---

# Setup

```{r setup, message=FALSE}
library(conos)
library(qs)
library(dplyr)
library(magrittr)
library(pagoda2)
library(scHelper) # github.com/rrydbirk/scHelper
library(cacoa)
library(ggplot2)
library(rstatix)
library(ggpubr)
library(ComplexHeatmap)
library(destiny)
library(Seurat)
library(harmony)
library(slingshot)
library(circlize)
library(cowplot)
library(sccore)

meta <- qread("meta.qs")

tt <- Sys.time()
```

# Figure 1

## Load data

```{r f1load}
con.major <- qread("con_major.qs", nthreads = 10)
anno.major <- qread("anno_major.qs")
```

## Figure 1G

```{r f1g}
con.major$plotGraph(groups = anno.major, 
                    plot.na = FALSE, 
                    size = 0.1,
                    alpha = 0.1, 
                    embedding = "UMAP_refined7",
                    show.labels = TRUE, 
                    font.size = 5) + 
  labs(x = "UMAP1", y = "UMAP2") + 
  theme(line = element_blank()) + 
  scale_color_manual(values = RColorBrewer::brewer.pal(9, "Set1"))
```

## Figure 1H

```{r f1h}
cm.merged <- con.major$getJointCountMatrix() %>% 
  .[rownames(.) %in% names(anno.major), ]

c("ADIPOQ", "PLIN1", # Adipocytes
  "PDGFRA", "DCN", # ASPCs
  "MECOM", "ADGRL4", # ECs
  "PKHD1L1", "PROX1", # LECs
  "MYOCD", "MYH11", # SMCs
  "STEAP4", "GIPR", # Pericytes
  "IL7R", "CD3E", # Lymphoid
  "KIT", "CPA3", # Mast
  "MRC1", "F13A1") %>% # Myeloid
  sccore::dotPlot(., 
                  cm.merged, 
                  factor(anno.major[rownames(cm.merged)], 
                         levels = c("Adipocytes", 
                                    "ASPCs", 
                                    "ECs", 
                                    "Lymphatic ECs", 
                                    "SMCs", 
                                    "Pericytes", 
                                    "Lymphoid immune cells", 
                                    "Mast cells", 
                                    "Myeloid immune cells")), 
                  gene.order = ., 
                  cols = c("white","grey20"))
```

## Figure 1I

```{r f1i, fig.height=5, fig.width=9}
# Use Cacoa to calculate proportions
cao <- Cacoa$new(data.object = con.major, 
                 cell.groups = anno.major, 
                 ref.level = "vis2", 
                 target.level = "vis3", 
                 sample.groups = con.major$samples %>% 
                   names() %>% 
                   strsplit("_") %>% 
                   sapply('[[', 3) %>% 
                   gsub(1, 2, .) %>% 
                   `names<-`(con.major$samples %>% 
                               names()),
                 sample.groups.palette = ggsci::pal_jama()(7))

cao$sample.groups <- con.major$samples %>% 
  names() %>% 
  strsplit("_") %>% 
  sapply('[[', 3) %>% 
  `names<-`(con.major$samples %>% 
              names()) %>% 
  gsub("vis", "Visit ", .)

p <- cao$plotCellGroupSizes(show.significance = TRUE, 
                            filter.empty.cell.types = FALSE)
```

```{r f1i2, fig.height=5, fig.width=9}
plot.dat <- p$data %>% 
  mutate(sex = rep(meta$sex, 9)) %>% 
  mutate(sample = stringi::stri_dup(LETTERS[seq(14)], 3) %>% 
           paste(collapse = "") %>% 
           strsplit("") %>% 
           unlist() %>% 
           rep(9)) %>% 
  mutate(sex_visit = paste(sex, group, sep = " ") %>% 
           gsub("Vis", "vis", .),
         facet = ifelse(variable %in% c("SMCs", "Mast cells", "Lymphatic ECs", "Pericytes", "Lymphoid immune cells"), "low", "high"),
         variable = factor(variable, levels = c("Adipocytes", "ASPCs", "ECs", "Myeloid immune cells", "Lymphatic ECs", "SMCs", "Pericytes", "Lymphoid immune cells", "Mast cells")) %>% 
           renameAnnotation("Myeloid immune cells", "   Myeloid immune cells")) %>% 
  mutate(sex_visit_anno = paste0(sex_visit, "_", variable))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$variable)) %>%
  lapply(lapply, \(x) kruskal.test(x$value, g = x$group)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(ct = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(ct = factor(ct, levels = unique(plot.dat$variable))) %>% 
  arrange(ct) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T),
         facet = ifelse(ct %in% c("SMCs", "Mast cells", "Lymphatic ECs", "Pericytes", "Lymphoid immune cells"), "low", "high"))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(var, sex) %>%
  wilcox_test(value~sex_visit, paired = T) %>%
  mutate(., p.adj.signif = apply(., 1, \(x) if(is.na(x[10])) "" else if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p), !var %in% c("Adipocytes", "Lymphoid immune cells")) %>% # Remove those not significant for KW
  add_xy_position(x = "var", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05) %>% # Adjust line position
  arrange(desc(y.position)) %>% 
  mutate(facet = ifelse(var %in% c("SMCs","Mast cells","Lymphatic ECs","Pericytes","Lymphoid immune cells"), "low", "high"))

# Plot
p.high <- plot.dat %>% 
  filter(facet == "high") %>% 
  ggplot(aes(variable, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "% cells per sample") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1)) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat %>% filter(facet == "high"), aes(ct, y = 65, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test %>% filter(facet == "high"), label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) + # Wilcoxon
  guides(fill = "none")

x.cor <- 4

p.low <- plot.dat %>% 
  filter(facet == "low") %>%
  ggplot(aes(variable, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "", fill = "") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1)) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = FALSE, data = sex.stat %>% filter(facet == "low"), aes(ct, y = 15, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test %>% filter(facet == "low") %>% mutate(x = x-x.cor, xmin = xmin-x.cor, xmax = xmax-x.cor), label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 2, hide.ns = FALSE) # Wilcoxon

pp <- plot_grid(plotlist = list(p.high, p.low), rel_widths = c(1,1.5))

pp
```

# Figure 2

## Load data

```{r f2load}
con <- qread("con_vascular.qs", nthreads = 10)
anno.vascular <- qread("anno_vascular.qs")
```

## Figure 2A

```{r f2a}
con$plotGraph(groups = anno.vascular, 
              plot.na = FALSE, 
              size = 0.2,
              alpha = 1, 
              embedding = "UMAP", 
              shuffle.colors = TRUE, 
              show.labels = TRUE, 
              font.size = 5) + 
  labs(x = "UMAP1", y = "UMAP2") + 
  theme(line = element_blank()) +
  scale_colour_manual(values = RColorBrewer::brewer.pal(9, "Purples")[-c(1,2)][c(1,4,7,2,6,3,5)])
```

## Figure 2B

```{r f2b}
cm.merged <- con$getJointCountMatrix() %>% 
  .[rownames(.) %in% names(anno.vascular), ]

c("PCSK5", "NEBL", "FN1", "CADM2", "BTNL9", "CEACAM1", "IL1R1", "VCAN", "ACKR1", "PKHD1L1", "MMRN1", "RELN", "MYH11", "ACTA2", "MYOCD", "COL25A1", "STEAP4", "NCKAP5") %>%  
  sccore::dotPlot(., 
                  cm.merged, 
                  anno.vascular %>% factor(levels = c("arEC", "cEC", "cvEC", "vEC", "lEC", "SMCs", "Pericytes")), 
                  gene.order = ., 
                  cols = c("white","purple3"))
```

## Figure 2D

Here, we're plotting proportions of total cells, so we load the Conos object and the annotation for all cells first.

```{r f2d}
con.major <- qread("con_major.qs", nthreads = 10)
anno.major <- qread("anno_major.qs")

# Use Cacoa to calculate proportions
cao <- Cacoa$new(data.object = con.major, 
                 cell.groups = anno.major, 
                 ref.level = "vis2", 
                 target.level = "vis3",
                 sample.groups = con.major$samples %>% 
                   names() %>% 
                   strsplit("_") %>% 
                   sapply('[[', 3) %>% 
                   gsub(1, 2, .) %>% 
                   `names<-`(con.major$samples %>% 
                               names()),
                 sample.groups.palette = ggsci::pal_jama()(7))

cao$sample.groups <- con.major$samples %>% 
  names() %>% 
  strsplit("_") %>% 
  sapply('[[', 3) %>% 
  `names<-`(con.major$samples %>% 
              names()) %>% 
  gsub("vis", "Visit ", .)

anno.comb <- anno.major[!names(anno.major) %in% names(anno.vascular)] %>% 
  {factor(c(., anno.vascular))}

p <- cao$plotCellGroupSizes(cell.groups = anno.comb,
                            show.significance = FALSE, 
                            filter.empty.cell.types = FALSE)
```

```{r f2d2}
plot.dat <- p$data %>% 
  filter(variable %in% levels(anno.vascular)) %>% 
  mutate(sex = rep(meta$sex, anno.vascular %>% levels() %>% length())) %>% 
  mutate(sex_visit = paste(sex, group, sep = " ") %>% gsub("Vis", "vis", .),
         facet = ifelse(variable %in% c("cvEC", "lEC"), "low", "high"),
         variable = factor(variable, levels = c("arEC", "cEC", "vEC", "SMCs", "Pericytes", "cvEC", "lEC")) %>% 
           renameAnnotation("lEC", "         lEC"))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$variable)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$group)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(ct = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(ct = factor(ct, levels = unique(plot.dat$variable))) %>% 
  arrange(ct) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T),
         facet = ifelse(ct %in% c("cvEC", "         lEC"), "low", "high"))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(var, sex) %>%
  wilcox_test(value~sex_visit, paired = T) %>%
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "var", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05) %>% # Adjust line position
  arrange(desc(y.position)) %>% 
  mutate(y.position = c(y.position[1]+1, y.position[2], y.position[3]+1, y.position[4], y.position[5]+1, y.position[6:7]),  # Manually adjust overlapping lines
         facet = ifelse(var %in% c("cvEC", "         lEC"), "low", "high"))

# Plot
p.high <- ggplot(plot.dat %>% filter(facet == "high"), aes(variable, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "% cells per sample") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat %>% filter(facet == "high"), aes(ct, y = 22.5, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test %>% filter(facet == "high"), label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) + # Wilcoxon
  guides(fill = "none")

p.low <- ggplot(plot.dat %>% filter(facet == "low"), aes(variable, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "", fill = "") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat %>% filter(facet == "low"), aes(ct, y = 4.5, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test %>% filter(facet == "low"), label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) # Wilcoxon

plot_grid(plotlist = list(p.high, p.low), rel_widths = c(1,0.8))
```

# Figure 3

## Load data

```{r f3load}
con <- qread("con_myeloid.qs", nthreads = 10)
anno.immune <- qread("anno_immune.qs")

anno.myeloid <- anno.immune %>% 
  .[!. %in% c("NKT","NK","Tem","Th1","Th","Treg","B-cells")] %>% 
  factor(levels = c("ATM", "mono/mac", "Early LAM", "LAM", "cDC1", "cDC2", "mDC"))
```

## Figure 3A

```{r f3a}
con$plotGraph(groups = anno.immune[!anno.immune == "B-cells"], 
              plot.na = FALSE, 
              size = 0.2,
              alpha = 1, 
              embedding = "largeVis", 
              show.labels = TRUE, 
              font.size = 5) + 
  labs(x = "largeVis1", y = "largeVis2") + 
  theme(line = element_blank()) +
  scale_colour_manual(values = c("#174D86", "#9EC3B5", "#356790", "#ADD0BA", "#628EA0", "#80A9AA", "#719CA5"))
```

## Figure 3B

```{r f3b}
cm.merged <- con$getJointCountMatrix(raw = FALSE) %>% 
  .[rownames(.) %in% names(anno.myeloid), ]

c("LYVE1", "SELENOP", # ATM
  "FCN1","VCAN", # mono/mac
  "CYP27A1", "APOE", # Early LAM
  "LPL", "TREM2", # LAM
  "CLEC9A", "CADM1", # cDC1
  "CD1C", "CLEC10A", # cDC2
  "CCR7", "LAMP3") %>% # mDC
  sccore::dotPlot(., 
                  cm.merged, 
                  anno.myeloid, 
                  gene.order = ., 
                  cols = c("white","#377EB8"))
```

## Figure 3C

Here, we're plotting proportions of total cells, so we load the Conos object and the annotation for all cells first.
```{r f3c}
con.major <- qread("con_major.qs", nthreads = 10)
anno.major <- qread("anno_major.qs")

# Use Cacoa to calculate proportions
cao <- Cacoa$new(data.object = con.major, 
                 cell.groups = anno.major, 
                 ref.level = "vis2", 
                 target.level = "vis3", 
                 sample.groups = con.major$samples %>% 
                   names() %>% 
                   strsplit("_") %>% 
                   sapply('[[', 3) %>% 
                   gsub(1, 2, .) %>% 
                   `names<-`(con.major$samples %>% 
                               names()),
                 sample.groups.palette = ggsci::pal_jama()(7))

cao$sample.groups <- con.major$samples %>% 
  names() %>% 
  strsplit("_") %>% 
  sapply('[[', 3) %>% 
  `names<-`(con.major$samples %>% 
              names()) %>% 
  gsub("vis", "Visit ", .)

anno.comb <- anno.major[!names(anno.major) %in% names(anno.myeloid)] %>% 
  {factor(c(., anno.myeloid))}

p <- cao$plotCellGroupSizes(cell.groups = anno.comb,
                            show.significance = FALSE, 
                            filter.empty.cell.types = FALSE)
```

```{r f3c2}
plot.dat <- p$data %>% 
  filter(variable %in% levels(anno.myeloid)) %>% 
  mutate(sex = rep(meta$sex, anno.myeloid %>% levels() %>% length()),
         variable = factor(variable)) %>% 
  mutate(sex_visit = paste(sex, group, sep = " ") %>% gsub("Vis", "vis", .),
         facet = ifelse(variable != "ATM", "low", "high"),
         variable = factor(variable, levels = c("ATM", "mono/mac", "Early LAM", "LAM", "cDC1", "cDC2", "mDC")) %>% 
           renameAnnotation("ATM", "         ATM"))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$variable)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$group)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(ct = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(ct = factor(ct, levels = unique(plot.dat$variable))) %>% 
  arrange(ct) %>% 
  mutate(sig = gsub(".", "", sig, fixed = TRUE),
         facet = ifelse(ct != "         ATM", "low", "high"))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(var, sex) %>%
  wilcox_test(value~sex_visit, paired = TRUE) %>%
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "var", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05) %>% # Adjust line position
  arrange(desc(y.position)) %>% 
  mutate(facet = ifelse(var != "         ATM", "low", "high"))

# Plot
p.high <- ggplot(plot.dat %>% filter(facet == "high"), aes(variable, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "% cells per sample") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1)) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat %>% filter(facet == "high"), aes(ct, y = 50, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test %>% filter(facet == "high"), label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) + # Wilcoxon
  guides(fill = "none")

x.cor = 1

p.low <- ggplot(plot.dat %>% filter(facet == "low"), aes(variable, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "", fill = "") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1)) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = FALSE, data = sex.stat %>% filter(facet == "low"), aes(ct, y = 10, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test %>% filter(facet == "low") %>% mutate(x = x-x.cor, xmin = xmin-x.cor, xmax = xmax-x.cor), label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 1, hide.ns = FALSE) # Wilcoxon

plot_grid(plotlist = list(p.high, p.low), rel_widths = c(1,4.5))
```

## Figure 3E

Preparations

```{r f3e}
cm.merged <- con$getJointCountMatrix(raw = TRUE) %>% 
  Matrix::t() %>% 
  .[, colnames(.) %in% names(anno.myeloid)]

# Calculate DEGs using Cacoa
con.vis13 <- con$samples %>% 
  .[!grepl("vis2", names(.))] %>% 
  Conos$new()

sample.groups <- con.vis13$samples %>% 
  names() %>% 
  `names<-`(ifelse(grepl("vis1", .), "visit1", "visit3"), .)

cao <- Cacoa$new(con.vis13, 
                 sample.groups, 
                 anno.myeloid, 
                 ref.level = "visit1", 
                 target.level = "visit3", 
                 n.cores = 10)

cao$estimateDEPerCellType(min.cell.frac = 0.1, 
                          min.cell.count = 5)

de <- cao$test.results$de %>%
  lapply('[[', "res")

# Create sample-wise annotation
anno.donor <- con$getDatasetPerCell()[colnames(cm.merged)]
anno.subtype <- anno.myeloid %>% 
  .[!is.na(.)]

idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())

anno.donor %<>% .[idx]
anno.subtype %<>% .[idx] %>% 
  {`names<-`(as.character(.), names(.))}

anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
  `names<-`(anno.donor %>% names()) %>% 
  .[grepl("ATM|Early LAM|LAM|mono/mac", .)]

# Create pseudo CM
cm.pseudo <- sccore::collapseCellsByType(cm.merged %>% Matrix::t(), 
                                         groups = anno.final, min.cell.count = 1) %>% 
  t() %>%
  apply(2, as, "integer")

cm.pseudo %<>% 
  DESeq2::DESeqDataSetFromMatrix(., 
                                 colnames(.) %>% 
                                   strsplit("_|!!") %>% 
                                   sget(3) %>% 
                                   data.frame() %>% 
                                   `dimnames<-`(list(colnames(cm.pseudo), "group")), 
                                 design = ~ group) %>% 
  DESeq2::estimateSizeFactors() %>% 
  DESeq2::counts(normalized = T)

genes <- de[c("ATM","Early LAM","LAM","mono/mac")] %>%
  {`names<-`(lapply(names(.), \(nn) mutate(.[[nn]], ct = nn)), names(.))} %>% 
  lapply(filter, padj <= 0.05, abs(log2FoldChange) > 1) %>%
  lapply(pull, Gene) %>% 
  {sapply(names(.), \(nn) .[[nn]] %>% `names<-`(., rep(nn, length(.))))} %>% 
  Reduce(c, .) %>% 
  .[match(unique(.), .)]

idx <- cm.pseudo %>% 
  colnames() %>% 
  data.frame(id = .) %>% 
  mutate(vis = strsplit(id, "_|!!") %>% sget(3),
         ct = strsplit(id, "!!") %>% sget(2)) %>% 
  mutate(ord = order(vis, ct))

x <- cm.pseudo %>% 
  .[match(genes, rownames(.)), match(colnames(na.omit(.)), colnames(.))] %>%
  .[, idx$ord] %>% 
  Matrix::t() %>%
  scale() %>%
  Matrix::t()

# Ordering
spc <- con$getDatasetPerCell()

sepc <- meta$sex[match(spc, meta$sample)] %>% 
  setNames(names(spc))

sex.df <- sepc %>% 
  {data.frame(nn = names(.), sex = unname(.))} %>% 
  mutate(nn = strsplit(nn, "!!") %>% sget(1)) %>% 
  .[match(unique(.$nn), .$nn), ] %>% 
  filter(!grepl("pool", nn))

ord <- data.frame(Visit = colnames(x) %>%
                    setNames(colnames(x)) %>% 
                    strsplit("_|!!|vis") %>%
                    sget(4)) %>% 
  mutate(., 
         donor = rownames(.) %>% strsplit("_vis") %>% sget(1)) %>% 
  mutate(., 
         Sex = sex.df$sex[match(.$donor, sex.df$nn)],
         ct = rownames(.) %>% strsplit("!!") %>% sget(2)) %>% 
  arrange(ct, Visit, Sex) %>% 
  rownames()

x <- x[, ord]

groups <- colnames(x) %>%
  `names<-`(colnames(x))

# Limit scale
x[x > 2] <- 2
x[x < -1.5] <- -1.5
```

Please note, the order of the clusters is not always reproducible.

```{r f3e2}
# Create top annotation
tannot <- data.frame(Visit = groups[colnames(x)] %>%
                       strsplit("_|!!|vis") %>%
                       sget(4)) %>% 
  mutate(., donor = rownames(.) %>% strsplit("_vis") %>% sget(1)) %>% 
  mutate(., Sex = sex.df$sex[match(.$donor, sex.df$nn)]) %>% 
  select(-donor) %>% 
  {HeatmapAnnotation(df=.,
                     border=T,
                     col=list(Visit = ggsci::pal_jama()(7)[c(1:3)] %>% `names<-`(c(1:3)),
                              Sex = c("darkred","blue3") %>% setNames(c("Female","Male"))),
                     show_legend=T)}

# Create color palette
pal <- colorRampPalette(c('navy','grey95','firebrick'))(1024)
labeled.gene.subset <- NULL

# Plot
set.seed(1337)

ha <- ComplexHeatmap::Heatmap(x, 
                              name='Expression', 
                              col=pal, 
                              cluster_columns=FALSE, 
                              show_row_names=FALSE, 
                              show_column_names=FALSE, 
                              top_annotation=tannot,
                              left_annotation=NULL,
                              border=TRUE,
                              show_column_dend = FALSE, 
                              show_row_dend = FALSE,
                              row_km = 5,
                              row_km_repeats = 500,
                              column_split = groups[colnames(x)] %>% strsplit("!!") %>% sapply('[[', 2) %>% unname() %>% factor(levels = c("ATM", "mono/mac", "Early LAM", "LAM")))

ht = draw(ha); row_order(ht) -> cls
```

Calculate GO enrichment

```{r f3e3}
go <- cls %>% 
  lapply(\(y) clusterProfiler::enrichGO(rownames(x)[y],
                                        OrgDb = "org.Hs.eg.db", 
                                        keyType = "SYMBOL", 
                                        ont = "BP", 
                                        universe = rownames(cm.merged))) %>% 
  setNames(names(cls))
```

## Figures 3F,G

Preparations

```{r f3fg, fig.width=6, fig.height=4}
cts <- c("Early LAM", "LAM", "mono/mac", "ATM")
cm.merged <- con$getJointCountMatrix(raw = TRUE)

# Get markers
markers <- con$getDifferentialGenes(groups = anno.myeloid %>% .[. %in% cts], 
                                    z.threshold = 1, 
                                    upregulated.only = TRUE, 
                                    append.specificity.metrics = TRUE, 
                                    append.auc = TRUE)

genes <- markers$LAM %>% 
  filter(AUC > 0.8) %>% 
  pull(Gene)

# Create sample-wise annotation
anno.donor <- con$getDatasetPerCell()[rownames(cm.merged)]
anno.subtype <- anno.myeloid[anno.myeloid %in% cts] %>% 
  .[!is.na(.)] %>% 
  factor()

idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())

anno.donor %<>% .[idx]
anno.subtype %<>% .[idx] %>% 
  {`names<-`(as.character(.), names(.))}

anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
  `names<-`(anno.donor %>% names())

cm.pseudo.tmp <- sccore::collapseCellsByType(cm.merged,
                                             groups = anno.final, min.cell.count = 1) %>% 
  t() %>%
  apply(2, as, "integer")

cm.pseudo <- cm.pseudo.tmp %>% 
  DESeq2::DESeqDataSetFromMatrix(., 
                                 colnames(.) %>% 
                                   strsplit("_|!!") %>% 
                                   sget(3) %>% 
                                   data.frame() %>% 
                                   `dimnames<-`(list(colnames(cm.pseudo.tmp), "group")), 
                                 design = ~ group) %>% 
  DESeq2::estimateSizeFactors() %>% 
  DESeq2::counts(normalized = TRUE) %>% 
  t()
```

### Figure 3F

```{r f3f}
plot.dat <- cm.pseudo %>% 
  .[, colnames(.) %in% c("LPL")] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("ATM", "mono/mac", "Early LAM", "LAM")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>%
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", title = "LPL", fill = "") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 500, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) # Wilcoxon
```

### Figure 3G

```{r f3g}
plot.dat <- cm.pseudo %>%
  .[, colnames(.) %in% genes] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("ATM", "mono/mac", "Early LAM", "LAM")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " ")) %>% 
  group_by(visit, sample, anno, donor, sex, sex_visit) %>%
  summarize(value = mean(value)) %>% 
  filter(visit != "visit 2") %>% 
  mutate(visit = factor(visit),
         donor_anno = paste(strsplit(donor, "_") %>% sapply(\(x) paste(x[1], x[2], sep = "_")), anno, sep = "_")) %>% 
  arrange(visit, sample) %$% 
  split(., visit) %>% 
  {lapply(., filter, donor_anno %in% .[[2]]$donor_anno)} %>% 
  {data.frame(.[[1]], diff = .[[2]]$value - .[[1]]$value)}

plot.dat %>% 
  ggplot(aes(anno, diff)) +
  geom_boxplot(aes(fill = sex)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_point(aes(fill = sex), position = position_jitterdodge(jitter.width = 0.1),
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Mean normalized pseudobulk difference\nin top gene marker expression", title = "LAM top markers") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c("darkred", "blue3")) +
  stat_compare_means(mapping = aes(fill = sex), method = "wilcox", label = "p.signif")
```

```{r f3g2}
plot.dat %>% 
  mutate(sex_visit_anno = paste(sex_visit, anno, sep = "_")) %$% 
  split(diff, sex_visit_anno) %>% 
  lapply(t.test)
```

# Figure 4

## Load data

```{r f4load}
con <- qread("con_aspc.qs", nthreads = 10)
anno.aspc <- qread("anno_aspc.qs")
```

## Figure 4A

```{r f4a}
con$plotGraph(groups = anno.aspc, 
              plot.na = FALSE, 
              size = 0.2,
              alpha = 1, 
              embedding = "largeVis",
              show.labels = TRUE, 
              font.size = 5) + 
  labs(x = "largeVis1", y = "largeVis2") + 
  theme(line = element_blank()) +
  scale_colour_manual(values = RColorBrewer::brewer.pal(5, "Greens")[-1][c(2,1,3,4)])
```

## Figure 4B

```{r f4b}
cm.merged <- con$getJointCountMatrix()

c("DPP4", "SEMA3C", "FBN1", # DPP4
  "CXCL14", "CFD", "C3", # CXCL14
  "PPARG", "ACACB", "SEMA3A", # PPARG
  "EPHA3", "MEOX2", "IGFBP7" # EPHA3
) %>% 
  sccore::dotPlot(., 
                  cm.merged, 
                  anno.aspc %>% factor(levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")), 
                  gene.order = ., 
                  cols = c("white","green4"))
```

## Figure 4C

Here, we're plotting proportions of total cells, so we load the Conos object and the annotation for all cells first.
```{r f4c}
con.major <- qread("con_major.qs", nthreads = 10)
anno.major <- qread("anno_major.qs")

# Use Cacoa to calculate proportions
cao <- Cacoa$new(data.object = con.major, 
                 cell.groups = anno.major, 
                 ref.level = "vis2", 
                 target.level = "vis3", 
                 sample.groups = con.major$samples %>% 
                   names() %>% 
                   strsplit("_") %>% 
                   sapply('[[', 3) %>% 
                   gsub(1, 2, .) %>% 
                   `names<-`(con.major$samples %>% 
                               names()),
                 sample.groups.palette = ggsci::pal_jama()(7))

cao$sample.groups <- con.major$samples %>% 
  names() %>% 
  strsplit("_") %>% 
  sapply('[[', 3) %>% 
  `names<-`(con.major$samples %>% 
              names()) %>% 
  gsub("vis", "Visit ", .)

anno.comb <- anno.major[!names(anno.major) %in% names(anno.aspc)] %>% 
  {factor(c(., anno.aspc))}

p <- cao$plotCellGroupSizes(cell.groups = anno.comb,
                            show.significance = FALSE, 
                            filter.empty.cell.types = FALSE)
```

```{r f4c2}
plot.dat <- p$data %>% 
  filter(variable %in% levels(anno.aspc)) %>% 
  mutate(sex = rep(meta$sex, anno.aspc %>% levels() %>% length()),
         variable = factor(variable)) %>% 
  mutate(sex_visit = paste(sex, group, sep = " ") %>% gsub("Vis", "vis", .),
         variable = factor(variable, levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$variable)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$group)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(ct = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(ct = factor(ct, levels = unique(plot.dat$variable))) %>% 
  arrange(ct) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(var, sex) %>%
  wilcox_test(value~sex_visit, paired = T)

stat.test[1, "p.adj.signif"] <- "."
stat.test[15, "p.adj.signif"] <- "."

stat.test %<>% 
  filter(p.adj.signif != "ns", !is.na(p), !var %in% c("ASPC_CXCL14")) %>% # Remove those not significant for KW
  add_xy_position(x = "var", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05) %>% # Adjust line position
  arrange(desc(y.position))

# Plot
ggplot(plot.dat, aes(variable, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1), 
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "% cells per sample", fill = "") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1)) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(ct, y = 20, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 2, hide.ns = F) # Wilcoxon
```

## Figure 4D

Preparations

```{r f4d}
cm.merged <- con$getJointCountMatrix(raw = TRUE) %>% 
  Matrix::t() %>% 
  .[, colnames(.) %in% names(anno.aspc)]

# Calculate DEGs using Cacoa
con.vis13 <- con$samples %>% 
  .[!grepl("vis2", names(.))] %>% 
  Conos$new()

sample.groups <- con.vis13$samples %>% 
  names() %>% 
  `names<-`(ifelse(grepl("vis1", .), "visit1", "visit3"), .)

cao <- Cacoa$new(con.vis13, 
                 sample.groups, 
                 anno.aspc, 
                 ref.level = "visit1", 
                 target.level = "visit3", 
                 n.cores = 10)

cao$estimateDEPerCellType()

de <- cao$test.results$de %>%
  lapply('[[', "res")

# Create sample-wise annotation
anno.donor <- con$getDatasetPerCell()[colnames(cm.merged)]
anno.subtype <- anno.aspc %>% 
  .[!is.na(.)]

idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())

anno.donor %<>% .[idx]
anno.subtype %<>% .[idx] %>% 
  {`names<-`(as.character(.), names(.))}

anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
  `names<-`(anno.donor %>% names())

# Create pseudo CM
cm.pseudo <- sccore::collapseCellsByType(cm.merged %>% Matrix::t(), 
                                         groups = anno.final, min.cell.count = 1) %>% 
  t() %>%
  apply(2, as, "integer")

cm.pseudo %<>% 
  DESeq2::DESeqDataSetFromMatrix(., 
                                 colnames(.) %>% 
                                   strsplit("_|!!") %>% 
                                   sget(3) %>% 
                                   data.frame() %>% 
                                   `dimnames<-`(list(colnames(cm.pseudo), "group")), 
                                 design = ~ group) %>% 
  DESeq2::estimateSizeFactors() %>% 
  DESeq2::counts(normalized = T)

genes <- de %>%
  {`names<-`(lapply(names(.), \(nn) mutate(.[[nn]], ct = nn)), names(.))} %>% 
  lapply(filter, padj <= 0.05, abs(log2FoldChange) > 1) %>%
  lapply(pull, Gene) %>% 
  {sapply(names(.), \(nn) .[[nn]] %>% `names<-`(., rep(nn, length(.))))} %>% 
  Reduce(c, .) %>% 
  .[match(unique(.), .)]

idx <- cm.pseudo %>% 
  colnames() %>% 
  data.frame(id = .) %>% 
  mutate(vis = strsplit(id, "_|!!") %>% sget(3),
         ct = strsplit(id, "!!") %>% sget(2)) %>% 
  mutate(ord = order(vis, ct))

x <- cm.pseudo %>% 
  .[match(genes, rownames(.)), match(colnames(na.omit(.)), colnames(.))] %>%
  .[, idx$ord] %>% 
  Matrix::t() %>%
  scale() %>%
  Matrix::t()

# Ordering
spc <- con$getDatasetPerCell()

sepc <- meta$sex[match(spc, meta$sample)] %>% 
  setNames(names(spc))

sex.df <- sepc %>% 
  {data.frame(nn = names(.), sex = unname(.))} %>% 
  mutate(nn = strsplit(nn, "!!") %>% sget(1)) %>% 
  .[match(unique(.$nn), .$nn), ] %>% 
  filter(!grepl("pool", nn))

ord <- data.frame(Visit = colnames(x) %>%
                    setNames(colnames(x)) %>% 
                    strsplit("_|!!|vis") %>%
                    sget(4)) %>% 
  mutate(., 
         donor = rownames(.) %>% strsplit("_vis") %>% sget(1)) %>% 
  mutate(., 
         Sex = sex.df$sex[match(.$donor, sex.df$nn)],
         ct = rownames(.) %>% strsplit("!!") %>% sget(2)) %>% 
  arrange(ct, Visit, Sex) %>% 
  rownames()

x <- x[, ord]

groups <- colnames(x) %>%
  `names<-`(colnames(x))

# Limit scale
x[x > 2] <- 2
x[x < -1] <- -1
```

Please note, the order of the clusters is not always reproducible.

```{r f4d2}
# Create top annotation
tannot <- data.frame(Visit = groups[colnames(x)] %>%
                       strsplit("_|!!|vis") %>%
                       sget(4)) %>% 
  mutate(., donor = rownames(.) %>% strsplit("_vis") %>% sget(1)) %>% 
  mutate(., Sex = sex.df$sex[match(.$donor, sex.df$nn)]) %>% 
  select(-donor) %>% 
  {HeatmapAnnotation(df=.,
                     border=TRUE,
                     col=list(Visit = ggsci::pal_jama()(7)[c(1:3)] %>% `names<-`(c(1:3)),
                              Sex = c("darkred","blue3") %>% setNames(c("Female","Male"))),
                     show_legend=TRUE)}

# Create color palette
pal <- colorRampPalette(c('navy','grey95','firebrick'))(1024)
labeled.gene.subset <- NULL

# Plot
set.seed(1337)

ha <- ComplexHeatmap::Heatmap(x, 
                              name='Expression', 
                              col=pal, 
                              cluster_columns=FALSE, 
                              show_row_names=FALSE, 
                              show_column_names=FALSE, 
                              top_annotation=tannot,
                              left_annotation=NULL,
                              border=TRUE,
                              show_column_dend = FALSE, 
                              show_row_dend = FALSE,
                              row_km = 3,
                              row_km_repeats = 100,
                              column_split = groups[colnames(x)] %>% strsplit("!!") %>% sapply('[[', 2) %>% unname() %>% factor(levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")),
                              row_dend_reorder = TRUE)

ht = draw(ha); row_order(ht) -> cls
```

Calculate GO enrichment

```{r f4d3}
go <- cls %>% 
  lapply(\(y) clusterProfiler::enrichGO(rownames(x)[y],
                                        OrgDb = "org.Hs.eg.db", 
                                        keyType = "SYMBOL", 
                                        ont = "BP", 
                                        universe = rownames(cm.merged))) %>% 
  setNames(names(cls))
```

## Figure 4E

Please note, cluster number may be different.

```{r f4e}
genes.go <- clusterProfiler::bitr("GO:0097193", 
                                  fromType="GOALL", 
                                  toType="SYMBOL", 
                                  OrgDb='org.Hs.eg.db')$SYMBOL %>% 
  unique()

genes.cls <- x %>% 
  rownames() %>% 
  .[cls[[1]]] # Change if needed

genes <- intersect(genes.go, genes.cls)
```

```{r f4e2, fig.width=6, fig.height=4}
cm.merged <- con$getJointCountMatrix(raw = T)

# Create sample-wise annotation
anno.donor <- con$getDatasetPerCell()[rownames(cm.merged)]
anno.subtype <- anno.aspc %>% 
  .[!is.na(.)] %>% 
  factor()

idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())

anno.donor %<>% .[idx]
anno.subtype %<>% .[idx] %>% 
  {`names<-`(as.character(.), names(.))}

anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
  `names<-`(anno.donor %>% names())

cm.pseudo.tmp <- sccore::collapseCellsByType(cm.merged,
                                             groups = anno.final, min.cell.count = 1) %>% 
  t() %>%
  apply(2, as, "integer")

cm.pseudo <- cm.pseudo.tmp %>% 
  DESeq2::DESeqDataSetFromMatrix(., 
                                 colnames(.) %>% 
                                   strsplit("_|!!") %>% 
                                   sget(3) %>% 
                                   data.frame() %>% 
                                   `dimnames<-`(list(colnames(cm.pseudo.tmp), "group")), 
                                 design = ~ group) %>% 
  DESeq2::estimateSizeFactors() %>% 
  DESeq2::counts(normalized = T) %>% 
  t()
```

```{r f4e3, fig.width=5, fig.height=4}
plot.dat <- cm.pseudo %>%
  .[, colnames(.) %in% genes] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " ")) %>% 
  group_by(visit, sex, anno, sex_visit, sample) %>%
  summarize(value = mean(value))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>% 
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Mean normalized pseudobulk expression", title = "Apoptosis ontology gene expression") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 250, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) + # Wilcoxon
  guides(fill = "none")
```

## Figure 4G

See `Scenic.ipynb` for `mat` object calculation.

```{r f4g, fig.height = 16, fig.width = 16}
mat <- fastMatMR::fmm_to_mat("aspc_adi_auc_mat.mtx") %>% 
  `dimnames<-`(list(
    read.table("aspc_adi_auc_mat.rownames", header = T)[, 1],
    read.table("aspc_adi_auc_mat.colnames", header = T)[, 1]
  ))

con.major <- qread("con_major.qs", nthreads = 10)
spc <- con.major$getDatasetPerCell()
vpc <- spc %>% 
  as.character() %>% 
  strsplit("_|!!") %>% 
  sget(3) %>% 
  setNames(names(spc))
```

```{r f4g2, fig.height = 4, fig.width = 5}
anno.adipocytes <- qread("anno_major.qs") %>% 
  .[. == "Adipocytes"] %>% 
  factor()
anno.comb <- factor(c(anno.adipocytes, anno.aspc))

plot.dat <- c("JUN(+)", "ZEB1(+)") %>% # ZEB1 is just a random pick to make it easier to manipulate data, omitted later
  {mat[, colnames(mat) %in% .]} %>% 
  as.data.frame() %>%
  mutate(., 
         visit = unname(vpc)[match(rownames(.), 
                                   names(vpc) %>% 
                                     strsplit("!!") %>% 
                                     sapply(\(x) paste0(x[2],"!!",x[3]))
         )
         ],
         anno = unname(anno.comb)[match(rownames(.),
                                        names(anno.comb) %>% 
                                          strsplit("!!") %>% 
                                          sapply(\(x) paste0(x[2],"!!",x[3]))
         )
         ],
         sample = unname(spc)[match(rownames(.), 
                                    names(spc) %>% 
                                      strsplit("!!") %>% 
                                      sapply(\(x) paste0(x[2],"!!",x[3]))
         )
         ]
  ) %>% 
  filter(!is.na(anno)) %>% 
  reshape2::melt(id.vars = c("visit", "anno", "sample")) %>% 
  mutate(anno_vis = paste0(anno,"_",visit),
         anno_var = paste0(anno,"_",variable)) %>% 
  group_by(visit, anno, variable, sample) %>% 
  summarize(mm = median(value)) %>% 
  ungroup() %>% 
  filter(variable == "JUN(+)",
         !anno == "Adipocytes") %>% 
  mutate(anno = factor(anno, levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3", "Adipocytes")),
         visit = factor(visit, levels = c("vis1", "vis2", "vis3"), labels = c("Visit 1", "Visit 2", "Visit 3")),
         sex = meta$sex[match(.$sample, meta$sample)],
         variable = factor(variable)) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " ") %>% gsub("Visit", "visit", .),
         anno = factor(anno, levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")))
```

```{r f4g3, fig.height=4, fig.width=8}
# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$mm, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(ct = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(ct = factor(ct, levels = unique(plot.dat$anno))) %>% 
  arrange(ct) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(anno, sex) %>%
  wilcox_test(mm~sex_visit, paired = TRUE) %>%
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05) %>% # Adjust line position
  arrange(desc(y.position))

# Plot
ggplot(plot.dat, aes(anno, mm)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1), 
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Mean AUC", fill = "") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(ct, y = 6, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 2, hide.ns = F) # Wilcoxon
```

# Figure 5

## Load data

```{r f5load}
con <- qread("con_adipocytes.qs", nthreads = 10)
anno.adipocytes <- qread("anno_adipocytes_archetypes.qs")

missing.cells <- setdiff(rownames(con$embedding), 
                         names(anno.adipocytes)) %>% 
  {setNames(rep("", length(.)), .)} %>% 
  factor()

anno.adipocytes.ext <- factor(c(anno.adipocytes, missing.cells))
```

## Figure 5A

```{r f5a}
con$plotGraph(groups = anno.adipocytes.ext, 
              show.labels = TRUE, 
              embedding = "largeVis") +
  labs(x = "largeVis1", y = "largeVis2") +
  theme(line = element_blank()) + 
  scale_colour_manual(values = c(RColorBrewer::brewer.pal(4, "Reds")[-1], "gray60"))
```

## Figure 5B

```{r f5b}
cm.merged <- con$getJointCountMatrix()

c("PLXDC2","CBLB","TCF4",
  "GPAM","ACSL1","LPL","DGAT2",
  "PRSS23","THSD7A","AL139317.5") %>% 
  sccore::dotPlot(., 
                  cm.merged, 
                  anno.adipocytes, 
                  gene.order = ., 
                  cols = c("white","red3"))
```

## Figure 5C

Preparations

```{r f5c}
cm.merged <- con$getJointCountMatrix(raw = TRUE) %>% 
  Matrix::t() %>% 
  .[, colnames(.) %in% names(anno.adipocytes)]

# Calculate DEGs using Cacoa
con.vis13 <- con$samples %>% 
  .[!grepl("vis2", names(.))] %>% 
  Conos$new()

sample.groups <- con.vis13$samples %>% 
  names() %>% 
  `names<-`(ifelse(grepl("vis1", .), "visit1", "visit3"), .)

cao <- Cacoa$new(con.vis13, 
                 sample.groups, 
                 anno.adipocytes, 
                 ref.level = "visit1", 
                 target.level = "visit3", 
                 n.cores = 10)

cao$estimateDEPerCellType()

de <- cao$test.results$de %>%
  lapply('[[', "res")

# Create sample-wise annotation
anno.donor <- con$getDatasetPerCell()[colnames(cm.merged)]
anno.subtype <- anno.adipocytes %>% 
  .[!is.na(.)]

idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())

anno.donor %<>% .[idx]
anno.subtype %<>% .[idx] %>% 
  {`names<-`(as.character(.), names(.))}

anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
  `names<-`(anno.donor %>% names())

# Create pseudo CM
cm.pseudo <- sccore::collapseCellsByType(cm.merged %>% Matrix::t(), 
                                         groups = anno.final, min.cell.count = 1) %>% 
  t() %>%
  apply(2, as, "integer")

cm.pseudo %<>% 
  DESeq2::DESeqDataSetFromMatrix(., 
                                 colnames(.) %>% 
                                   strsplit("_|!!") %>% 
                                   sget(3) %>% 
                                   data.frame() %>% 
                                   `dimnames<-`(list(colnames(cm.pseudo), "group")), 
                                 design = ~ group) %>% 
  DESeq2::estimateSizeFactors() %>% 
  DESeq2::counts(normalized = T)

genes <- de %>%
  {`names<-`(lapply(names(.), \(nn) mutate(.[[nn]], ct = nn)), names(.))} %>% 
  lapply(filter, padj <= 0.05, abs(log2FoldChange) > 1) %>%
  lapply(pull, Gene) %>% 
  {sapply(names(.), \(nn) .[[nn]] %>% `names<-`(., rep(nn, length(.))))} %>% 
  Reduce(c, .) %>% 
  .[match(unique(.), .)]

idx <- cm.pseudo %>% 
  colnames() %>% 
  data.frame(id = .) %>% 
  mutate(vis = strsplit(id, "_|!!") %>% sget(3),
         ct = strsplit(id, "!!") %>% sget(2)) %>% 
  mutate(ord = order(vis, ct))

x <- cm.pseudo %>% 
  .[match(genes, rownames(.)), match(colnames(na.omit(.)), colnames(.))] %>%
  .[, idx$ord] %>% 
  Matrix::t() %>%
  scale() %>%
  Matrix::t()

# Ordering
spc <- con$getDatasetPerCell()

sepc <- meta$sex[match(spc, meta$sample)] %>% 
  setNames(names(spc))

sex.df <- sepc %>% 
  {data.frame(nn = names(.), sex = unname(.))} %>% 
  mutate(nn = strsplit(nn, "!!") %>% sget(1)) %>% 
  .[match(unique(.$nn), .$nn), ] %>% 
  filter(!grepl("pool", nn))

ord <- data.frame(Visit = colnames(x) %>%
                    setNames(colnames(x)) %>% 
                    strsplit("_|!!|vis") %>%
                    sget(4)) %>% 
  mutate(., 
         donor = rownames(.) %>% strsplit("_vis") %>% sget(1)) %>% 
  mutate(., 
         Sex = sex.df$sex[match(.$donor, sex.df$nn)],
         ct = rownames(.) %>% strsplit("!!") %>% sget(2)) %>% 
  arrange(ct, Visit, Sex) %>% 
  rownames()

x <- x[, ord]

groups <- colnames(x) %>%
  `names<-`(colnames(x))

# Limit scale
x[x > 2] <- 2
x[x < -1] <- -1
```

Please note, the order of the clusters is not always reproducible, and the clusters them selves may also change slightly.

```{r f5c2}
# Create top annotation
tannot <- data.frame(Visit = groups[colnames(x)] %>%
                       strsplit("_|!!|vis") %>%
                       sget(4)) %>% 
  mutate(., donor = rownames(.) %>% strsplit("_vis") %>% sget(1)) %>% 
  mutate(., Sex = sex.df$sex[match(.$donor, sex.df$nn)]) %>% 
  select(-donor) %>% 
  {HeatmapAnnotation(df=.,
                     border=TRUE,
                     col=list(Visit = ggsci::pal_jama()(7)[c(1:3)] %>% `names<-`(c(1:3)),
                              Sex = c("darkred","blue3") %>% setNames(c("Female","Male"))),
                     show_legend=TRUE)}

# Create color palette
pal <- colorRampPalette(c('navy','grey95','firebrick'))(1024)
labeled.gene.subset <- NULL

# Plot
set.seed(1337)

ha <- ComplexHeatmap::Heatmap(x, 
                              name='Expression', 
                              col=pal, 
                              cluster_columns=FALSE, 
                              show_row_names=FALSE, 
                              show_column_names=FALSE, 
                              top_annotation=tannot,
                              left_annotation=NULL,
                              border=TRUE,
                              show_column_dend = FALSE, 
                              show_row_dend = FALSE,
                              row_km = 4,
                              row_km_repeats = 200, # Here, this parameter greatly influences cluster composition
                              column_split = groups[colnames(x)] %>% strsplit("!!") %>% sapply('[[', 2) %>% unname() %>% factor(levels = c("CLSTN2_archetype", "DGAT2_archetype", "PRSS23_archetype")),
                              row_dend_reorder = TRUE)

ht = draw(ha); row_order(ht) -> cls
```

Calculate GO enrichment

```{r f5c3}
go <- cls %>% 
  lapply(\(y) clusterProfiler::enrichGO(rownames(x)[y],
                                        OrgDb = "org.Hs.eg.db", 
                                        keyType = "SYMBOL", 
                                        ont = "BP", 
                                        universe = rownames(cm.merged))) %>% 
  setNames(names(cls))
```

## Figure 5D

```{r f5d, fig.width=6, fig.height=4}
cm.merged <- con$getJointCountMatrix(raw = T)

# Create sample-wise annotation
anno.donor <- con$getDatasetPerCell()[rownames(cm.merged)]
anno.subtype <- anno.adipocytes %>% 
  .[!is.na(.)] %>% 
  factor()

idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())

anno.donor %<>% .[idx]
anno.subtype %<>% .[idx] %>% 
  {`names<-`(as.character(.), names(.))}

anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
  `names<-`(anno.donor %>% names())

cm.pseudo.tmp <- sccore::collapseCellsByType(cm.merged,
                                             groups = anno.final, min.cell.count = 1) %>% 
  t() %>%
  apply(2, as, "integer")

cm.pseudo <- cm.pseudo.tmp %>% 
  DESeq2::DESeqDataSetFromMatrix(., 
                                 colnames(.) %>% 
                                   strsplit("_|!!") %>% 
                                   sget(3) %>% 
                                   data.frame() %>% 
                                   `dimnames<-`(list(colnames(cm.pseudo.tmp), "group")), 
                                 design = ~ group) %>% 
  DESeq2::estimateSizeFactors() %>% 
  DESeq2::counts(normalized = T) %>% 
  t()
```

```{r f5d2, fig.width=6, fig.height=4}
gene <- "DECR1"

plot.dat <- cm.pseudo %>% 
  .[, colnames(.) %in% gene] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("DGAT2_archetype", "CLSTN2_archetype", "PRSS23_archetype")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>%
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", title = gene, fill = "") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 200, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F)# Wilcoxon
```

```{r f5d3, fig.width=6, fig.height=4}
gene <- "LIPA"

plot.dat <- cm.pseudo %>% 
  .[, colnames(.) %in% gene] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("DGAT2_archetype", "CLSTN2_archetype", "PRSS23_archetype")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>%
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", title = gene, fill = "") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 30, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F)# Wilcoxon
```

```{r f5d4, fig.width=6, fig.height=4}
gene <- "PLIN5"

plot.dat <- cm.pseudo %>% 
  .[, colnames(.) %in% gene] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("DGAT2_archetype", "CLSTN2_archetype", "PRSS23_archetype")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>% 
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", title = gene, fill = "") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 200, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F)# Wilcoxon
```

# Figure 6

## Figure 6A

Calculate diffusion embeddings. Please note, this takes some time. We saved the embeddings in `adipocytes_aspc_embeddings.qs`.

```{r f6a, eval = F}
## Load Conos object and annotation
data = qread("con_adipocytes_faps.qs", nthreads = 10)
anno.adipocytes <- qread("anno_major.qs") %>% 
  .[. == "Adipocytes"] %>% 
  factor()
anno.aspc <- qread("anno_aspc.qs")
anno <- factor(c(anno.adipocytes, anno.aspc))

## Convert raw counts to Seurat
for (name in names(data$samples)) {
  seu <- CreateSeuratObject(CreateAssayObject(counts = t(data$samples[[name]]$misc$rawCounts)))
  seu$orig.ident <- substr(name, 0, regexpr("vis", name)-2)
  seu$visit <- substr(name, regexpr("vis", name), nchar(name))
  if (name == names(data$samples)[1]) {
    final <- seu
  } else {
    final <- merge(final, seu)
  }
}

## Merge annotation
final <- subset(final, cells = names(anno))
final <- NormalizeData(final)
anno.final <- anno[ match(colnames(final), names(anno))]
final$fine <- anno.final
final$coarse <- substr(anno.final, 0, regexpr("_", anno.final)-1)
final$dataset <- paste(final$orig.ident, final$visit, sep="_")
final <- subset(final, cells = names(which(!is.na(final$fine))))
full <- final

embeddings <- list()
for (nfeats in c(25,50,100,250)) { 
  # Recreate original object
  final <- full
  # ASPC
  ASPC <- subset(final, coarse == "ASPC")
  VariableFeatures(ASPC) <- NULL
  obj.list <- SplitObject(ASPC, split.by = "dataset")
  feats_ASPC <- suppressWarnings(SelectIntegrationFeatures(obj.list, nfeatures = nfeats, verbose = FALSE))
  # Adipocytes
  Ad <- subset(final, coarse == "Adipocytes")
  VariableFeatures(Ad) <- NULL
  obj.list <- SplitObject(Ad, split.by = "dataset")
  feats_Ad <- suppressWarnings(SelectIntegrationFeatures(obj.list, nfeatures = nfeats, verbose = FALSE))
  # Combine
  feats <- unique(c(feats_ASPC, feats_Ad))
  ## Embed
  # PCA
  VariableFeatures(final) <- feats
  final <- ScaleData(final)
  final <- RunPCA(final)
  for (useHarmony in c(TRUE, FALSE)) {
    if (useHarmony) {
      final <- suppressWarnings(RunHarmony(final, group.by.vars = "dataset", verbose  = FALSE))
      pcs <- final@reductions$harmony@cell.embeddings
    } else {
      pcs <- final@reductions$pca@cell.embeddings
    }
    for (dims in c(5, 10, 15, 20)) {
      dm <- destiny::DiffusionMap(pcs[,1:dims], verbose = FALSE, n_pcs = NA)
      dm_ev <- dm@eigenvectors
      colnames(dm_ev) <- paste("DC", c(1:20), sep="_")
      embeddings[[(length(embeddings) + 1)]] <- dm_ev
      names(embeddings)[length(embeddings)] <- paste(nfeats, useHarmony, dims, sep="_")
    }
  }
}
```

Prepare data. We chose to continue with embedding no. 2. The `sds_obj` object is saved for Figure 6E.

```{r f6a2}
# Load and prepare data
embeddings <- qread("adipocytes_aspc_embeddings.qs")
anno.adipocytes <- qread("anno_major.qs") %>% 
  .[. == "Adipocytes"] %>% 
  factor()
anno.aspc <- qread("anno_aspc.qs")
anno <- factor(c(anno.adipocytes, anno.aspc))

# Calculate slingshot trajectory
anno.sort <- anno[rownames(embeddings[[2]])] %>%
  as.character() %>%
  unname()

sds_obj <- slingshot(embeddings[[2]][, 1:2] %>% 
                       `colnames<-`(c("UMAP1","UMAP2")), 
                     anno.sort, 
                     end.clus = "Adipocytes", 
                     start.clus = "ASPC_DPP4")

qsave(sds_obj, "sds_obj.qs")

sds <- as.SlingshotDataSet(sds_obj)
```

Final plot

```{r f6a3}
plot.df <- embeddings[[2]][,1:2] %>% 
  as.data.frame() %>% 
  `colnames<-`(c("UMAP1","UMAP2")) %>% 
  mutate(., annotation = anno[rownames(.)] %>% as.factor()) %>% 
  .[complete.cases(.),]

line.df <- lapply(1:length(sds@curves), \(lineage) {
  sds@curves[[lineage]]$s[,1:2] %>% 
    data.frame() %>% 
    mutate(lineage = lineage %>% as.factor())
}) %>% 
  setNames(sds@curves %>% names()) %>% 
  bind_rows() %>% 
  mutate(lineage = factor(lineage, labels = c("Lineage 1", "Lineage 2")))

ggplot(plot.df, aes(UMAP1, UMAP2, col = annotation)) + 
  geom_point(size = 0.3) +
  geom_path(data = line.df, aes(UMAP1, UMAP2, group = lineage, col = lineage), inherit.aes = F, linewidth = 1) + 
  theme_bw() + 
  theme(legend.position = "right",
        line = element_blank()) + 
  labs(col = "", x = "DC1", y = "DC2") +
  scale_colour_manual(values = c("#E41A1C", RColorBrewer::brewer.pal(5, "Greens")[-1][c(2,1,3,4)], "black", "grey40")) +
  dotSize(3) +
  ylim(c(-0.03, 0.013)) +
  xlim(c(-0.006, 0.0052))
```

## Figure 6B

We need the `sds_obj` object from Figure 6A.

```{r f6b, fig.width=5, fig.height=2}
sds_obj <- qread("sds_obj.qs")

pseudotime_all <- sds_obj@assays@data@listData$pseudotime %>% 
  as.data.frame() %>% 
  filter(!is.na(Lineage1)) %>% 
  {setNames(pull(., Lineage1), rownames(.))}

plot.df %>% 
  .[names(pseudotime_all), ] %>%
  mutate(pseudotime = pseudotime_all) %>%
  ggplot(aes(UMAP1, UMAP2, col = pseudotime)) +
  geom_point(size = 0.3) +
  geom_path(data = line.df %>% filter(lineage == "Lineage 1"), aes(UMAP1, UMAP2), inherit.aes = F, color = "black", linewidth = 1, alpha = 0.4) +
  theme_bw() +
  theme(line = element_blank()) +
  labs(col = "Pseudotime", x = "DC1", y = "DC2") +
  scale_color_continuous(low = "blue3", high = "orange") +
  ylim(c(-0.005, 0.013)) +
  xlim(c(-0.006, 0.0052))
```

## Figure 6C

We need the `sds_obj` object from Figure 6A.

```{r f6c, fig.width=6, fig.height=2}
sds_obj <- qread("sds_obj.qs")

pseudotime_all <- sds_obj@assays@data@listData$pseudotime %>% 
  as.data.frame() %>% 
  filter(!is.na(Lineage1)) %>% 
  {setNames(pull(., Lineage1), rownames(.))}

plot.df %>% 
  .[names(pseudotime_all), ] %>%
  mutate(pseudo = pseudotime_all) %>%
  mutate(bin = cut(pseudo, breaks = c(0, 0.005, 0.01, 0.0165, 0.023))) %>% 
  na.omit() %>% 
  mutate(bin = as.character(bin) %>% 
           strsplit(",") %>% 
           sget(2) %>% 
           gsub("]", "", ., fixed = TRUE)) %>%
  arrange(bin) %>% 
  mutate(bin = factor(bin, labels = c("Early preadipocytes", "Mid preadipocytes", "Late preadipocytes", "Adipocyte"))) %>% 
  ggplot(aes(UMAP1, UMAP2, col = bin)) +
  geom_point(size = 0.3) +
  geom_path(data = line.df %>% filter(lineage == "Lineage 1"), aes(UMAP1, UMAP2), inherit.aes = FALSE, color = "black", linewidth = 1, alpha = 0.4) +
  theme_bw() +
  theme(line = element_blank()) +
  labs(col = "Binned pseudotime", x = "DC1", y = "DC2") +
  scale_color_manual(values = colorRampPalette(c("pink","darkred"))(4)) +
  ylim(c(-0.005, 0.013)) +
  xlim(c(-0.006, 0.0052)) +
  dotSize(3)
```

## Figure 6D

See `Scenic.ipynb` for `mat` object calculation.

```{r f6d, fig.height = 16, fig.width = 16}
mat <- fastMatMR::fmm_to_mat("aspc_adi_auc_mat.mtx") %>%
  `dimnames<-`(list(
    read.table("aspc_adi_auc_mat.rownames", header = T)[, 1],
    read.table("aspc_adi_auc_mat.colnames", header = T)[, 1]
  ))

con.major <- qread("con_major.qs", nthreads = 10)
spc <- con.major$getDatasetPerCell()
vpc <- spc %>% 
  as.character() %>% 
  strsplit("_|!!") %>% 
  sget(3) %>% 
  setNames(names(spc))

anno.adipocytes <- qread("anno_major.qs") %>% 
  .[. == "Adipocytes"] %>% 
  factor()
anno.aspc <- qread("anno_aspc.qs")
anno <- factor(c(anno.adipocytes, anno.aspc))
```

We need the `sds_obj` object from Figure 6A.

```{r f6d2, fig.height = 4, fig.width = 5}
sds_obj <- qread("sds_obj.qs")

mat.df <- c("ZEB1", "EBF3", "PPARG") %>%
  paste0("(+)") %>%
  {mat[, colnames(mat) %in% .]} %>% 
  as.data.frame() %>% 
  mutate(., 
         visit = unname(vpc)[match(rownames(.), 
                                   names(vpc) %>% 
                                     strsplit("!!") %>% 
                                     sapply(\(x) paste0(x[2],"!!",x[3]))
         )
         ] %>% 
           factor(labels = c("Visit 1", "Visit 2", "Visit 3")),
         anno = unname(anno)[match(rownames(.),
                                   names(anno) %>% 
                                     strsplit("!!") %>% 
                                     sapply(\(x) paste0(x[2],"!!",x[3]))
         )
         ],
         sample = unname(spc)[match(rownames(.), 
                                    names(spc) %>% 
                                      strsplit("!!") %>% 
                                      sapply(\(x) paste0(x[2],"!!",x[3]))
         )
         ]
  ) %>% 
  filter(!is.na(anno))

pseudotime_all <- sds_obj@assays@data@listData$pseudotime %>% 
  as.data.frame() %>% 
  filter(!is.na(Lineage1)) %>% 
  mutate(., 
         cid = rownames(.) %>% 
           strsplit("!!") %>% 
           sapply(\(x) paste0(x[2],"!!",x[3])))

mat.df %<>% .[match(pseudotime_all$cid, rownames(.), nomatch = F), ]

pseudotime_all %<>% .[match(rownames(mat.df), .$cid, nomatch = F), ]

mat.df2 <- mat.df %>% 
  mutate(pseudotime = pseudotime_all$Lineage1) %>%
  mutate(bin = cut(pseudotime, breaks = c(0, 0.005, 0.01, 0.0165, 0.023))) %>% 
  na.omit() %>% 
  mutate(bin = as.character(bin) %>% 
           strsplit(",") %>% 
           sget(2) %>% 
           gsub("]", "", ., fixed = T)) %>%
  select(-anno, -pseudotime) %>% 
  reshape2::melt(id.vars = c("visit", "bin", "sample"), variable.name = "regulon") %>%
  group_by(visit, regulon, bin, sample) %>%
  mutate(value = as.numeric(value)) %>% 
  summarize(mm = median(value))
```

```{r f6d3, fig.width=18, fig.height=4}
mat.df2 %>% 
  pull(regulon) %>% 
  unique() %>% 
  as.character() %>% 
  lapply(\(vv) {
    plot.dat <- mat.df2 %>% 
      filter(regulon == vv) %>%
      ungroup() %>%
      mutate(sex = meta$sex[match(.$sample, meta$sample)]) %>% 
      mutate(sex_visit = paste(sex, visit, sep = " ") %>% gsub("Visit", "visit", .))

    # Kruskal-Wallis
    sex.stat <- plot.dat %$% 
      split(., sex) %>% 
      lapply(\(x) split(x, x$bin)) %>% 
      lapply(lapply, \(x) kruskal.test(x$mm, g = x$visit)$p.value) %>% 
      lapply(sapply, gtools::stars.pval) %>% 
      {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
      `names<-`(c("Female","Male")) %>% 
      lapply(\(x) data.frame(ct = names(x), sig = unname(x))) %>% 
      {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
      bind_rows() %>% 
      mutate(ct = factor(ct, levels = unique(plot.dat$bin))) %>% 
      arrange(ct) %>% 
      mutate(sig = gsub(".", "", sig, fixed = T))

    # Wilcoxon
    stat.test <- plot.dat %>%
      group_by(bin, sex) %>%
      wilcox_test(mm~sex_visit, paired = F)

    stat.test %<>% 
      filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
      add_xy_position(x = "bin", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05) %>% # Adjust line position
      arrange(desc(y.position))

    # Plot
    pp <- ggplot(plot.dat, aes(bin, mm)) + 
      geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
      geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1), 
                 color = "black", size = 0.5, alpha = 0.2) +
      theme_bw() +
      labs(x = "", y = "Mean AUC", fill = "", title = vv) +
      theme(line = element_blank(),
            axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
            axis.text = element_text(color = "black")) +
      scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
      geom_text(inherit.aes = F, data = sex.stat, aes(ct, y = 2.2, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 0.2, hide.ns = F) # Wilcoxon

    pp
  }) %>% 
  cowplot::plot_grid(plotlist = ., ncol = 3)
```

## Figure 6E

### Prepare data

We downsampled to 15k nuclei due to calculation times. We provide `sds_obj_reduced.qs` for slingshot pseudotime estimations for the downsampled dataset.

Here, we calculate for visit 1 and visit 3.

```{r f6e, eval = F}
sds_obj <- qread("sds_obj_reduced.qs")

pseudotime <- sds_obj@assays@data@listData$pseudotime[, 1] %>% 
  .[!is.na(.)]

con <- qread("con_adipocytes_aspc.qs", nthreads = 10)

mat <- con$getJointCountMatrix() %>%
  .[match(names(pseudotime), rownames(.)), colSums(.) > 2e2]

mt <- mat %>%
  as.matrix() %>%
  as.data.frame() %>%
  tibble::rownames_to_column() %>%
  {message("Mutating"); mutate(.,
                               pseudotime = pseudotime[rowname],
                               visit = con$getDatasetPerCell()[rowname] %>%
                                 as.character() %>%
                                 strsplit("_") %>%
                                 sget(3),
                               donor = strsplit(rowname, "!!") %>%
                                 sget(1))} %>%
  .[complete.cases(.), ] %>%
  {message("Melting"); reshape2::melt(., id.vars = c("rowname", "pseudotime", "visit", "donor"))} %$%
  {message("Splitting"); split(., variable)}
```

The calculations take app. 11 h with 14k genes and 10k cells. We've added pre-calculated results as `pseudotime_res_novis2.qs`.

```{r f6e2, eval = F}
res <- mt %>% 
  sccore::plapply(\(y) {
    x = y %>% 
      filter(!visit == "vis2")

    fit_full <- gamm4(data = x, formula = value ~ s(pseudotime, by = factor(visit)), random = ~(1 | donor), REML = F)
    fit_reduced <- gamm4(data = x, formula = value ~ s(pseudotime), random = ~(1 | donor), REML = F)
    fit_none <- gamm4(data = x, formula = value ~ 1, random = ~(1 | donor), REML = F)

    ann <- anova(fit_full$mer, fit_reduced$mer, fit_none$mer)

    if (any(ann$`Pr(>Chisq)` <= 0.05, na.rm = T)) {
      residuals <- predict(fit_full$gam, se.fit = T)$fit %>% 
        unname()

      r.sq <- c(summary(fit_full$gam)$r.sq, summary(fit_reduced$gam)$r.sq) %>% 
        setNames(c("full", "reduced"))

      out <- list(annova = ann,
                  residuals = residuals,
                  r.sq = r.sq)

      return(out)
    }
  }, n.cores = 5, mc.preschedule = T, mc.cleanup = T, progress = F) # 10 cores fails, 5 is max tested that finishes
```

### Plot

Load data. We need the `sds_obj` from Figure 6A. Also, for ease of use we provide slingshot pseudotime W/O visit 2 as `pseudotime_novis2.qs`.

```{r f6e3}
res <- qread("pseudotime_res_novis2.qs", nthreads = 10) %>% 
  .[!sapply(., is.null)]

sds_obj <- qread("sds_obj.qs")

pseudotime <- qread("pseudotime_novis2.qs")

con <- qread("con_adipocytes_aspc.qs", nthreads = 10)
```

```{r f6e4}
both <- res %>%
  sapply(\(gene) {
    ann <- gene$annova
    (ann[2, "Pr(>Chisq)"] <= 0.05 & ann[3, "Pr(>Chisq)"] <= 0.05)
  })

res.both <- res[both]

p.both <- res.both %>% 
  sapply(\(gene) gene$annova[3, "Pr(>Chisq)"])

tmp <- res.both %>% 
  lget("annova") %>% 
  lapply(dplyr::slice, 2:3) %>% 
  lapply(pull, AIC) %>%
  .[sapply(., \(x) abs(x[2]) > abs(x[1]))] %>% 
  bind_rows() %>% 
  t() %>% 
  as.data.frame() %>% 
  setNames(c("reduced", "full")) %>% 
  mutate(diff = abs(full) - abs(reduced)) %>% 
  mutate(diff.frac = diff / abs(reduced))

res.filter <- res[
  tmp %>% 
    filter(diff.frac >= 0.05) %>% 
    rownames()
]

cm.merged <- con$getJointCountMatrix() %>% 
  .[match(names(pseudotime), rownames(.)), colnames(.) %in% names(res.filter)]

visit = con$getDatasetPerCell()[rownames(cm.merged)] %>%
  as.character() %>%
  strsplit("_") %>%
  sget(3)
```

```{r f6e5}
# Vis3
cm.merged.vis <- cm.merged[visit == "vis3", ]
pseudotime.vis <- pseudotime[rownames(cm.merged.vis)]
weights.vis <- sds_obj@assays@data@listData$weights %>% .[match(names(pseudotime.vis), rownames(.)), 1] + 1E-7

scFit <- cm.merged.vis %>% 
  Matrix::t() %>% 
  tradeSeq::fitGAM(pseudotime = pseudotime.vis, cellWeights = weights.vis, verbose = TRUE)

Smooth3 <- tradeSeq::predictSmooth(scFit, gene = colnames(cm.merged.vis), tidy = FALSE, n=100)

# Vis1
cm.merged.vis <- cm.merged[visit == "vis1", ]
pseudotime.vis <- pseudotime[rownames(cm.merged.vis)]
weights.vis <- sds_obj@assays@data@listData$weights %>% .[match(names(pseudotime.vis), rownames(.)), 1] + 1E-7

scFit <- cm.merged.vis %>% 
  Matrix::t() %>% 
  tradeSeq::fitGAM(pseudotime = pseudotime.vis, cellWeights = weights.vis, verbose = TRUE)

Smooth1 <- tradeSeq::predictSmooth(scFit, gene = colnames(cm.merged.vis), tidy = FALSE, n=100)

# Combine smooth
Smooth <- cbind(Smooth3, Smooth1)

# Average across replicates and scale
Smooth <- t(scale(t(Smooth)))

# Split smooth
Smooth33 <- Smooth[, 1:100]
Smooth11 <- Smooth[, 101:200]

# Vis3
# Seriate the results

Smooth33 <- Smooth33[ seriation::get_order(seriation::seriate(Smooth33, method="PCA_angle")), ]

col_fun = circlize::colorRamp2(c(-4, 0, 4), c("navy", "white", "firebrick"))

p3 <- Heatmap(Smooth33, 
              col = col_fun,
              cluster_columns=F, 
              cluster_rows=F, 
              show_column_names = F, 
              row_names_gp = grid::gpar(fontsize = 5)
)

# Vis1
Smooth11 %<>% .[p3@matrix %>% rownames(), ]

p1 <- Heatmap(Smooth11, 
              col = col_fun,
              cluster_columns=F, 
              cluster_rows=F, 
              show_column_names = F, 
              row_names_gp = grid::gpar(fontsize = 5)
)

p3
p1
```

# Figure 7

## Figure 7D

For calculation of data, see `Liana.ipynb`.

```{r f7d}
dat.raw <- read.delim("/work/02_data/00_other/liana_res.csv",
                      sep = ",",
                      header = T) %>%
  mutate(visit = strsplit(sample, "_") %>%
           sget(3))
```

Function

```{r f7d2}
lianaCircos <- function(df,
                        top.interactions = 30, 
                        text.size = 1, 
                        pal = RColorBrewer::brewer.pal(9, "Set1"), 
                        cell.types = c("Adipocytes", 
                                       "Myeloid immune cells", 
                                       "ASPCs", 
                                       "EC", 
                                       "SMCs", 
                                       "Lymphoid immune cells", 
                                       "Mast cells", 
                                       "Lymphatic ECs", 
                                       "Pericytes"),
                        big.gap = 5,
                        small.gap = 2,
                        arrow.width = 3,
                        link.ramp.rel = T,
                        link.sort = F,
                        scale = F,
                        arrow.head.width = 0.3,
                        arrow.head.length = 0.3,
                        link.ramp.col = c("navy", "grey", "firebrick")) {
  input_df <- df %>% 
    slice_max(order_by = score, n = top.interactions) %>% 
    mutate(source = paste0(source, " ")) %>% 
    mutate(source_lig = paste0(source, "|", ligand), 
           target_rec = paste0(target, "|", receptor))

  if (link.ramp.rel) {
    arr_wd <- rep(arrow.width, nrow(input_df))
  } else {
    arr_wd <- (((input_df$score - min(input_df$score))/(max(input_df$score) - min(input_df$score))) * (arrow.width)) + 1
  }

  # Colors and segments
  anno.col <- setNames(pal, 
                       cell.types) %>% 
    c(., "ASPCs " = unname(.["ASPCs"]))

  cell_cols <- anno.col[unique(c(unique(input_df$source), unique(input_df$target), "ASPCs "))]

  link_cols <- c()

  if (!link.ramp.rel) {
    for (i in input_df$source_lig) {
      link_cols <- c(link_cols, cell_cols[stringr::str_extract(i, 
                                                               "[^|]+")])
    }
  } else {
    input_df %<>% 
      arrange(score)

    df.down <- input_df %>% filter(score <= 0)
    link_down <- colorRampPalette(c(link.ramp.col[1], link.ramp.col[2]))(nrow(df.down))

    df.up <- input_df %>% filter(score > 0)
    link_up <- colorRampPalette(c(link.ramp.col[2], link.ramp.col[3]))(nrow(df.up))

    link_cols <- c(link_down, link_up)
  }

  segments <- unique(c(paste0(input_df$source, "|", input_df$ligand), 
                       paste0(input_df$target, "|", input_df$receptor)))

  grp <- stringr::str_extract(segments, "[^|]+") %>% 
    setNames(segments)

  # Redo colors
  cell_cols2 <- grp
  for (i in unique(grp)) {
    cell_cols2[cell_cols2 == i] <- cell_cols[i]
  }

  # Plot
  input_df %>% 
    select(source_lig, target_rec, score) %>%
    chordDiagram(directional = 1, 
                 group = grp,
                 scale = scale, 
                 diffHeight = 0.005, 
                 direction.type = c("arrows"),
                 link.arr.type = "triangle", 
                 annotationTrack = c(), 
                 preAllocateTracks = list(
                   list(track.height = 0.05),
                   list(track.height = 0.175),
                   list(track.height = 0.05)), 
                 big.gap = big.gap,
                 transparency = 1,
                 link.arr.lwd = arr_wd,
                 link.arr.col = link_cols,
                 link.arr.length = arrow.head.length,
                 link.arr.width = arrow.head.width,
                 small.gap = small.gap
    )

  circos.track(track.index = 2, panel.fun = function(x, y) {
    circos.text(CELL_META$xcenter, 
                CELL_META$ylim[1], 
                stringr::str_extract(CELL_META$sector.index, "[^|]+$"), 
                facing = "clockwise", 
                niceFacing = TRUE, 
                adj = c(0, 0.55), 
                cex = 1)
  }, bg.border = NA)

  # Split segments
  for (l in segments) {
    highlight.sector(l, track.index = 3, col = cell_cols2[l])
  }

  # Add ligand/receptor track
  ## Ligand
  highlight.sector(input_df$source_lig, 
                   track.index = 1, 
                   col = "black", 
                   text = "Ligands", 
                   cex = 1, 
                   text.col = "white", 
                   niceFacing = TRUE)
  ## Receptor
  highlight.sector(input_df$target_rec, 
                   track.index = 1, 
                   col = "white", 
                   text = "Receptors", 
                   cex = 1, 
                   text.col = "black", 
                   border = "black", 
                   niceFacing = TRUE)

  # Legends
  minmax <- input_df %>% 
    pull(score) %>% 
    {pmax(abs(min(.)), max(.))} %>% 
    formatC(digits = 1) %>% 
    as.numeric()

  col.range = c(-minmax, 0, minmax)
  lgd_links = Legend(at = col.range, 
                     col_fun = colorRamp2(col.range, link.ramp.col), 
                     title_position = "topleft", 
                     title = "Links")

  lgd_ct <- Legend(labels = unique(c(input_df$source, input_df$target)), 
                   title = "Cell type", 
                   type = "points", 
                   legend_gp = gpar(col = "transparent"), 
                   background = cell_cols[unique(c(input_df$source, input_df$target))])

  lgd_list_vertical = packLegend(lgd_ct, lgd_links)

  draw(lgd_list_vertical, 
       just = c("left", "bottom"), 
       x = unit(5, "mm"), 
       y = unit(5, "mm"))

  circos.clear()
}
```

Plot

```{r f7d3, fig.width=10, fig.height=8}
dat.plot <- dat.raw %>% 
  dplyr::rename(ligand = ligand_complex,
                receptor = receptor_complex,
                score = lrscore) %>%
  filter(visit %in% c("visit1", "visit3"),
         source == "ASPCs",
         ligand %in% c("COL18A1", "NID1", "JAG1", "FGF2", "VEGFA", "BMP1", "ANGPT1", "MFGE8", "C3", "IGF1", "ANXA1", "ALCAM"),
         receptor %in% c("ITGB5", "ITGAV", "CD46", "NRP1", "BMPR2", "ITGB1", "IGF2R", "INSR", "EGFR")) %>% 
  group_by(visit, ligand, receptor, source, target) %>% 
  summarize(score = mean(score)) %>% 
  ungroup() %>% 
  arrange(visit, source, target, ligand, receptor)

dat.vis1 <- dat.plot %>% 
  filter(visit == "visit1",
         score > 0.8648) %>% 
  mutate(lrst = paste0(ligand, receptor, source, target))

dat.vis3 <- dat.plot %>% 
  mutate(lrst = paste0(ligand, receptor, source, target)) %>% 
  filter(visit == "visit3",
         lrst %in% dat.vis1$lrst)

dat.rel <- dat.vis3 %>% 
  mutate(score = score - dat.vis1$score)

dat.rel %>% 
  lianaCircos()
```

## Figure 7E

Preparation

```{r f7e, fig.width=7, fig.height=4}
# Load Conos object
con <- qread("con_adipocytes_aspc.qs", nthreads = 10)

anno.faps <- qread("anno_aspc.qs")
anno <- factor(c(anno.adipocytes, anno.faps))

cm.merged <- con$getJointCountMatrix(raw = T) %>% 
  Matrix::t() %>% 
  .[, colnames(.) %in% names(anno)]

# Use bins instead of anno
spc <- con$getDatasetPerCell()

anno.bin <- qread("sds_obj.qs")@assays@data@listData$pseudotime %>% 
  as.data.frame() %>% 
  filter(!is.na(Lineage1)) %>% 
  {setNames(.$Lineage1, rownames(.))} %>% 
  {data.frame(pseudotime = unname(.), anno = anno[names(.)], cid = names(.))} %>% 
  mutate(bin = cut(pseudotime, breaks = c(0, 0.005, 0.01, 0.0165, 0.023))) %>% 
  na.omit() %>% 
  mutate(bin = as.character(bin) %>% 
           strsplit(",") %>% 
           sget(2) %>% 
           gsub("]", "", ., fixed = T),
         sample = spc[cid]) %>% 
  mutate(sex = meta$sex[match(sample, meta$sample)],
         visit = strsplit(as.character(sample), "_") %>% sget(3)) %>% 
  mutate(anno.final = paste(sample, bin, sex, sep = "!!"))

# Create pseudo CM
anno.final <- anno.bin %>% 
  pull(anno.final) %>% 
  setNames(anno.bin$cid)

cm.pseudo <- sccore::collapseCellsByType(cm.merged %>% Matrix::t(), 
                                         groups = anno.final, min.cell.count = 1) %>% 
  t() %>%
  apply(2, as, "integer")

cm.pseudo %<>% 
  DESeq2::DESeqDataSetFromMatrix(., 
                                 colnames(.) %>% 
                                   strsplit("_|!!") %>% 
                                   sget(3) %>% 
                                   data.frame() %>% 
                                   `dimnames<-`(list(colnames(cm.pseudo), "group")), 
                                 design = ~ group) %>% 
  DESeq2::estimateSizeFactors() %>% 
  DESeq2::counts(normalized = T)
```

```{r f7e2, fig.width=7, fig.height=4}
genes <- "C3"

idx <- cm.pseudo %>% 
  colnames() %>% 
  data.frame(id = .) %>% 
  mutate(vis = strsplit(id, "_|!!") %>% sget(3),
         bin = strsplit(id, "!!") %>% sget(2),
         sex = strsplit(id, "!!") %>% sget(3)) %>% 
  mutate(ord = order(vis, bin, sex))

x <- cm.pseudo %>% 
  .[match(genes, rownames(.)), match(colnames(na.omit(.)), colnames(.))] %>%
  .[idx$ord]

plot.dat <- x %>% 
  {data.frame(sample = names(.), 
              value = unname(.))} %>% 
  mutate(bin = strsplit(sample, "!!") %>% 
           sget(2),
         visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(labels = c("Visit 1", "Visit 2", "Visit 3")),
         sex = strsplit(sample, "!!") %>% sget(3)) %>% 
  mutate(sex_visit = paste(sex, gsub("Visit", "visit", visit), sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$bin)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(ct = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(ct = factor(ct, levels = unique(plot.dat$bin))) %>% 
  arrange(ct) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  group_by(bin, sex) %>%
  wilcox_test(value~sex_visit, paired = F)

stat.test %<>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "bin", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05) %>% # Adjust line position
  arrange(desc(y.position))

# Plot
ggplot(plot.dat, aes(bin, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1), 
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", fill = "", title = "C3") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(ct, y = 1.5e3, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 0.2, hide.ns = F) # Wilcoxon
```

# Figure S1

## Load data

We provide lists with filtered count matrices per visit. Please note, we re-sequenced some samples that are renamed in the following.

```{r s1load}
cms <- c("vis1_cms_filtered.qs",
         "vis2_cms_filtered.qs",
         "vis3_cms_filtered.qs") %>% 
  lapply(qread, nthreads = 10) %>% 
  mapply(\(x,y) setNames(x, paste0(names(x), "_", y)), x = ., y = c("visit1", "visit2", "visit3")) %>% 
  unlist() %>% 
  .[!names(.) %in% c("Donor_25_poolB_visit3","Donor_29_poolA_visit3")] %>% 
  setNames(gsub("_poolA|_poolB", "", names(.)))

con.major <- qread("con_major.qs", nthreads = 10)

spc <- con.major$getDatasetPerCell()
```

## Figure S1A

```{r s1a, fig.width=8, fig.height=4}
cms %>% 
  lapply(\(x) diff(x@p)) %>% 
  sapply(median) %>% 
  {data.frame(sample = names(.), 
              value = unname(.))} %>% 
  mutate(visit = strsplit(sample, "_") %>% 
           sget(3)) %>% 
  mutate(sample = paste0("Donor", seq(14), "_", visit) %>% 
           factor(., levels = unique(.))) %>% 
  ggplot(aes(sample, value, fill = visit)) +
  geom_col() +
  theme_bw() +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) +
  labs(x = "", y = "Median genes", fill = "") +
  scale_fill_manual(values = ggsci::pal_jama()(3))
```

```{r s1a2}
cms %>% 
  lapply(sparseMatrixStats::colSums2) %>% 
  sapply(median) %>% 
  {data.frame(sample = names(.), 
              value = unname(.))} %>% 
  mutate(visit = strsplit(sample, "_") %>% 
           sget(3)) %>% 
  mutate(sample = paste0("Donor", seq(14), "_", visit) %>% 
           factor(., levels = unique(.))) %>% 
  ggplot(aes(sample, value, fill = visit)) +
  geom_col() +
  theme_bw() +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) +
  labs(x = "", y = "Median UMIs", fill = "") +
  scale_fill_manual(values = ggsci::pal_jama()(3))
```

## Figure S1B

```{r s1b, fig.width=6, fig.height=4}
varToPlot <- spc %>% 
  as.character() %>% 
  strsplit("_vis") %>% 
  sget(1) %>% 
  setNames(names(spc)) %>% 
  factor()

con.major$plotGraph(groups = varToPlot, 
                    plot.na = F, 
                    size = 0.1,
                    alpha = 0.1, 
                    embedding = "UMAP_refined7",
                    mark.groups = F,
                    show.labels = T,
                    show.legend = T,
                    title = "Donor") +
  labs(x = "UMAP1", y = "UMAP2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3)
```

## Figure S1C

```{r s1c, fig.width=6, fig.height=4}
var <- meta$sex

varToPlot <- grepl.replace(spc %>% 
                             as.character(), 
                           meta$sample, 
                           var) %>% 
  setNames(names(spc)) %>% 
  factor(labels = unique(var))

con.major$plotGraph(groups = varToPlot, 
                    plot.na = F, 
                    size = 0.1,
                    alpha = 0.1, 
                    embedding = "UMAP_refined7",
                    mark.groups = F,
                    show.labels = T,
                    show.legend = T,
                    title = "Sex") +
  labs(x = "UMAP1", y = "UMAP2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3) +
  scale_color_manual(values = c("firebrick", "navy"))
```

## Figure S1D

```{r s1d, fig.width=6, fig.height=4}
varToPlot <- spc %>% 
  as.character() %>% 
  strsplit("_|!!") %>% 
  sget(3) %>% 
  setNames(names(spc)) %>% 
  factor()

con.major$plotGraph(groups = varToPlot, 
                    plot.na = F, 
                    size = 0.1,
                    alpha = 0.1, 
                    embedding = "UMAP_refined7",
                    mark.groups = F,
                    show.labels = T,
                    show.legend = T,
                    title = "Visit") +
  labs(x = "UMAP1", y = "UMAP2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3) +
  scale_color_manual(values = ggsci::pal_jama()(3))
```

# Figure S2

## Load data

```{r s2}
con <- qread("con_vascular.qs", nthreads = 10)
anno.vascular <- qread("anno_vascular.qs")

spc <- con$getDatasetPerCell()
```

## Figure S2A

```{r s2a, fig.width=6, fig.height=4}
varToPlot <- spc %>% 
  as.character() %>% 
  strsplit("_vis") %>% 
  sget(1) %>% 
  setNames(names(spc)) %>% 
  factor()

con$plotGraph(groups = varToPlot, 
              plot.na = F, 
              size = 0.3,
              alpha = 0.1, 
              embedding = "UMAP",
              mark.groups = F,
              show.labels = T,
              show.legend = T,
              title = "Donor") +
  labs(x = "UMAP2", y = "UMAP2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3)
```

## Figure S2B

```{r s2b, fig.width=6, fig.height=4}
var <- meta$sex

varToPlot <- grepl.replace(spc %>% 
                             as.character(), 
                           meta$sample, 
                           var) %>% 
  setNames(names(spc)) %>% 
  factor(labels = unique(var))

con$plotGraph(groups = varToPlot, 
              plot.na = F, 
              size = 0.1,
              alpha = 0.1, 
              embedding = "UMAP",
              mark.groups = F,
              show.labels = T,
              show.legend = T,
              title = "Sex") +
  labs(x = "UMAP1", y = "UMAP2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3) +
  scale_color_manual(values = c("firebrick", "navy"))
```

## Figure S2C

```{r s2c, fig.width=6, fig.height=4}
varToPlot <- spc %>% 
  as.character() %>% 
  strsplit("_|!!") %>% 
  sget(3) %>% 
  setNames(names(spc)) %>% 
  factor()

con$plotGraph(groups = varToPlot, 
              plot.na = F, 
              size = 0.3,
              alpha = 0.1, 
              embedding = "UMAP",
              mark.groups = F,
              show.labels = T,
              show.legend = T,
              title = "Visit") +
  labs(x = "UMAP1", y = "UMAP2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3) +
  scale_color_manual(values = ggsci::pal_jama()(3))
```

# Figure S3

## Figure S3B-G

### Load data

```{r s3bg}
con <- qread("con_lymphoid.qs", nthreads = 10)
anno.lymphoid <- qread("anno_lymphoid.qs")

spc <- con$getDatasetPerCell()
```

### Figure S3B

```{r s3b}
con$plotGraph(groups = anno.lymphoid, 
              plot.na = F, 
              size = 0.8,
              alpha = 1, 
              embedding = "largeVis", 
              shuffle.colors = T, 
              show.labels = T, 
              font.size = 5,
              show.legend = T,
              mark.groups = T) + 
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  scale_colour_manual(values = RColorBrewer::brewer.pal(9, "YlOrBr")[-c(1,2)][c(1,4,5,2,6)]) +
  dotSize(3)
```

### Figure S3C

```{r s3c}
varToPlot <- spc %>% 
  as.character() %>% 
  strsplit("_vis") %>% 
  sget(1) %>% 
  setNames(names(spc)) %>% 
  factor(labels = paste0("Donor", seq(14))) %>% 
  .[names(.) %in% names(anno.lymphoid)]

con$plotGraph(groups = varToPlot, 
              plot.na = F, 
              size = 0.5,
              alpha = 0.8, 
              embedding = "largeVis",
              mark.groups = F,
              show.labels = T,
              show.legend = T,
              title = "Donor") +
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3)
```

### Figure S3D

```{r s3d}
var <- meta$sex

varToPlot <- grepl.replace(spc %>% 
                             as.character(), 
                           meta$sample, 
                           var) %>% 
  setNames(names(spc)) %>% 
  factor(labels = unique(var)) %>% 
  .[names(.) %in% names(anno.lymphoid)]

con$plotGraph(groups = varToPlot, 
              plot.na = F, 
              size = 0.5,
              alpha = 0.8, 
              embedding = "largeVis",
              mark.groups = F,
              show.labels = T,
              show.legend = T,
              title = "Sex") +
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3) +
  scale_color_manual(values = c("firebrick", "navy"))
```

### Figure S3E

```{r s3e}
varToPlot <- spc %>% 
  as.character() %>% 
  strsplit("_|!!") %>% 
  sget(3) %>% 
  setNames(names(spc)) %>% 
  factor() %>% 
  .[names(.) %in% names(anno.lymphoid)]

con$plotGraph(groups = varToPlot, 
              plot.na = F, 
              size = 0.5,
              alpha = 0.8, 
              embedding = "largeVis",
              mark.groups = F,
              show.labels = T,
              show.legend = T,
              title = "Visit") +
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3) +
  scale_color_manual(values = ggsci::pal_jama()(3))
```

### Figure S3F

```{r s3f}
cm.merged <- con$getJointCountMatrix() %>% 
  .[rownames(.) %in% names(anno.lymphoid), ]

c("FOXP3", "CTLA4",
  "IL7R", "CD4",
  "CD40LG", "CD8A",
  "GZMK", "GZMH",
  "GNLY", "KLRD1",
  "KLRF1", "NCAM1"
) %>% 
  sccore::dotPlot(., 
                  cm.merged, 
                  anno.lymphoid %>% 
                    factor(levels = c("Treg", "CD4 T cells", "CD8 T cells", "NK-like T cells", "NK")), 
                  gene.order = ., 
                  cols = c("white","yellow4"))
```

### Figure S3G

```{r s3g}
con.major <- qread("con_major.qs", nthreads = 10)
anno.major <- qread("anno_major.qs")

# Use Cacoa to calculate proportions
cao <- Cacoa$new(data.object = con.major, 
                 cell.groups = anno.major, 
                 ref.level = "vis2", 
                 target.level = "vis3",
                 sample.groups = con.major$samples %>% 
                   names() %>% 
                   strsplit("_") %>% 
                   sapply('[[', 3) %>% 
                   gsub(1, 2, .) %>% 
                   `names<-`(con.major$samples %>% 
                               names()),
                 sample.groups.palette = ggsci::pal_jama()(7))

cao$sample.groups <- con.major$samples %>% 
  names() %>% 
  strsplit("_") %>% 
  sapply('[[', 3) %>% 
  `names<-`(con.major$samples %>% 
              names()) %>% 
  gsub("vis", "Visit ", .)

anno.comb <- anno.major[!names(anno.major) %in% names(anno.lymphoid)] %>% 
  {factor(c(., anno.lymphoid))}

p <- cao$plotCellGroupSizes(cell.groups = anno.comb,
                            show.significance = F, 
                       filter.empty.cell.types = F)
```

```{r s3g2, fig.height=4, fig.width=6}
plot.dat <- p$data %>% 
  filter(variable %in% levels(anno.lymphoid)) %>% 
  mutate(sex = rep(meta$sex, anno.lymphoid %>% levels() %>% length()),
         variable = factor(variable)) %>% 
  mutate(sex_visit = paste(sex, group, sep = " ") %>% gsub("Vis", "vis", .),
         variable = factor(variable, levels = c("Treg", "CD4 T cells", "CD8 T cells", "NK-like T cells", "NK")))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$variable)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$group)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(ct = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(ct = factor(ct, levels = unique(plot.dat$variable))) %>% 
  arrange(ct) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(var, sex) %>%
  wilcox_test(value~sex_visit, paired = T) %>%
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "var", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05) %>% # Adjust line position
  arrange(desc(y.position))

# Plot
ggplot(plot.dat, aes(variable, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
        color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "% cells per sample", fill = "") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1)) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(ct, y = 5, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 1, hide.ns = F) # Wilcoxon
```

## Figure S3H-J

### Load data

```{r s3hj}
con <- qread("con_myeloid.qs", nthreads = 10)
anno.immune <- qread("anno_immune.qs")

spc <- con$getDatasetPerCell()
```

### Figure S3H

```{r s3h}
varToPlot <- spc %>% 
  as.character() %>% 
  strsplit("_vis") %>% 
  sget(1) %>% 
  setNames(names(spc)) %>% 
  factor()

con$plotGraph(groups = varToPlot, 
                   plot.na = F, 
                   size = 0.3,
                   alpha = 0.1, 
                   embedding = "largeVis",
                   mark.groups = F,
                   show.labels = T,
                   show.legend = T,
                   title = "Donor") +
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3)
```

### Figure S3I

```{r s3i}
var <- meta$sex

varToPlot <- grepl.replace(spc %>% 
                             as.character(), 
                           meta$sample, 
                           var) %>% 
  setNames(names(spc)) %>% 
  factor(labels = unique(var))

con$plotGraph(groups = varToPlot, 
                   plot.na = F, 
                   size = 0.1,
                   alpha = 0.1, 
                   embedding = "largeVis",
                   mark.groups = F,
                   show.labels = T,
                   show.legend = T,
                   title = "Sex") +
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3) +
  scale_color_manual(values = c("firebrick", "navy"))
```

### Figure S3J

```{r s3j}
varToPlot <- spc %>% 
  as.character() %>% 
  strsplit("_|!!") %>% 
  sget(3) %>% 
  setNames(names(spc)) %>% 
  factor()

con$plotGraph(groups = varToPlot, 
                   plot.na = F, 
                   size = 0.3,
                   alpha = 0.1, 
                   embedding = "largeVis",
                   mark.groups = F,
                   show.labels = T,
                   show.legend = T,
                   title = "Visit") +
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3) +
  scale_color_manual(values = ggsci::pal_jama()(3))
```

## Figure S3L

Preparations

```{r s3l}
cts <- c("Early LAM", "LAM", "mono/mac", "ATM")

cm.merged <- con$getJointCountMatrix(raw = T)

# Create sample-wise annotation
anno.donor <- con$getDatasetPerCell()[rownames(cm.merged)]
anno.subtype <- anno.immune[anno.immune %in% cts] %>% 
  .[!is.na(.)] %>% 
  factor()

idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())

anno.donor %<>% .[idx]
anno.subtype %<>% .[idx] %>% 
  {`names<-`(as.character(.), names(.))}

anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
  `names<-`(anno.donor %>% names()) %>%
  gsub("Foam-like Mac", "Early LAM", .)

cm.pseudo.tmp <- sccore::collapseCellsByType(cm.merged,
                                         groups = anno.final, min.cell.count = 1) %>% 
  t() %>%
  apply(2, as, "integer")

cm.pseudo <- cm.pseudo.tmp %>% 
  DESeq2::DESeqDataSetFromMatrix(., 
                                 colnames(.) %>% 
                                   strsplit("_|!!") %>% 
                                   sget(3) %>% 
                                   data.frame() %>% 
                                   `dimnames<-`(list(colnames(cm.pseudo.tmp), "group")), 
                                 design = ~ group) %>% 
  DESeq2::estimateSizeFactors() %>% 
  DESeq2::counts(normalized = T) %>% 
  t()
```

```{r s3l2}
plot.dat <- cm.pseudo %>% 
  .[, colnames(.) %in% c("HLA-B")] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("ATM", "mono/mac", "Early LAM", "LAM")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>%
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
        color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", title = "HLA-B") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 400, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) # Wilcoxon
```

# Figure S4

## Figure S4A-C

### Load data

```{r s4ac}
con <- qread("con_aspc.qs", nthreads = 10)
anno.aspc <- qread("anno_aspc.qs")

spc <- con$getDatasetPerCell()
```

### Figure S4A

```{r s4a}
varToPlot <- spc %>% 
  as.character() %>% 
  strsplit("_vis") %>% 
  sget(1) %>% 
  setNames(names(spc)) %>% 
  factor()

con$plotGraph(groups = varToPlot, 
                   plot.na = F, 
                   size = 0.3,
                   alpha = 0.1, 
                   embedding = "largeVis",
                   mark.groups = F,
                   show.labels = T,
                   show.legend = T,
                   title = "Donor") +
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3)
```

### Figure S4B

```{r s4b}
var <- meta$sex

varToPlot <- grepl.replace(spc %>% 
                             as.character(), 
                           meta$sample, 
                           var) %>% 
  setNames(names(spc)) %>% 
  factor(labels = unique(var))

con$plotGraph(groups = varToPlot, 
                   plot.na = F, 
                   size = 0.1,
                   alpha = 0.1, 
                   embedding = "largeVis",
                   mark.groups = F,
                   show.labels = T,
                   show.legend = T,
                   title = "Sex") +
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3) +
  scale_color_manual(values = c("firebrick", "navy"))
```

### Figure S4C

```{r s4c}
varToPlot <- spc %>% 
  as.character() %>% 
  strsplit("_|!!") %>% 
  sget(3) %>% 
  setNames(names(spc)) %>% 
  factor()

con$plotGraph(groups = varToPlot, 
                   plot.na = F, 
                   size = 0.3,
                   alpha = 0.1, 
                   embedding = "largeVis",
                   mark.groups = F,
                   show.labels = T,
                   show.legend = T,
                   title = "Visit") +
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3) +
  scale_color_manual(values = ggsci::pal_jama()(3))
```

## Figure S4F-G

### Prepare data

```{r s4fg, fig.width=6, fig.height=4}
cm.merged <- con$getJointCountMatrix(raw = T)

# Create sample-wise annotation
anno.donor <- con$getDatasetPerCell()[rownames(cm.merged)]
anno.subtype <- anno.aspc %>% 
  .[!is.na(.)] %>% 
  factor()

idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())

anno.donor %<>% .[idx]
anno.subtype %<>% .[idx] %>% 
  {`names<-`(as.character(.), names(.))}

anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
  `names<-`(anno.donor %>% names())

cm.pseudo.tmp <- sccore::collapseCellsByType(cm.merged,
                                         groups = anno.final, min.cell.count = 1) %>% 
  t() %>%
  apply(2, as, "integer")

cm.pseudo <- cm.pseudo.tmp %>% 
  DESeq2::DESeqDataSetFromMatrix(., 
                                 colnames(.) %>% 
                                   strsplit("_|!!") %>% 
                                   sget(3) %>% 
                                   data.frame() %>% 
                                   `dimnames<-`(list(colnames(cm.pseudo.tmp), "group")), 
                                 design = ~ group) %>% 
  DESeq2::estimateSizeFactors() %>% 
  DESeq2::counts(normalized = T) %>% 
  t()
```

### Figure S4F

```{r s4f}
gene = "HSP90B1"

plot.dat <- cm.pseudo %>% 
  .[, colnames(.) %in% gene] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>% 
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
        color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", title = gene) +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 300, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) + # Wilcoxon
  guides(fill = "none")
```

### Figure S4G

```{r s4g, fig.width=6, fig.height=4}
gene = "JUN"

plot.dat <- cm.pseudo %>% 
  .[, colnames(.) %in% gene] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>% 
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
        color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", title = gene) +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 600, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) + # Wilcoxon
  guides(fill = "none")
```

```{r s4g2, fig.width=6, fig.height=4}
gene = "FOS"

plot.dat <- cm.pseudo %>% 
  .[, colnames(.) %in% gene] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>% 
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
        color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", title = gene) +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 2e3, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) + # Wilcoxon
  guides(fill = "none")
```

```{r s4g3, fig.width=6, fig.height=4}
gene = "FOSB"

plot.dat <- cm.pseudo %>% 
  .[, colnames(.) %in% gene] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>% 
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
        color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", title = gene) +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 1.6e3, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) + # Wilcoxon
  guides(fill = "none")
```

## Figure S4H

See `Scenic.ipynb` for `mat` object calculation.

```{r s4h, fig.height = 16, fig.width = 16}
mat <- fastMatMR::fmm_to_mat("aspc_adi_auc_mat.mtx") %>% 
  `dimnames<-`(list(
    read.table("aspc_adi_auc_mat.rownames", header = T)[, 1],
    read.table("aspc_adi_auc_mat.colnames", header = T)[, 1]
  ))

con.major <- qread("con_major.qs", nthreads = 10)
spc <- con.major$getDatasetPerCell()
vpc <- spc %>% 
  as.character() %>% 
  strsplit("_|!!") %>% 
  sget(3) %>% 
  setNames(names(spc))
```

```{r s4h2, fig.height = 4, fig.width = 5}
anno.adipocytes <- qread("anno_major.qs") %>% 
  .[. == "Adipocytes"] %>% 
  factor()
anno.comb <- factor(c(anno.adipocytes, anno.aspc))

plot.dat.tmp <- c("FOS(+)", "FOSB(+)") %>% # ZEB1 is just a random pick to make it easier to manipulate data, omitted later
  {mat[, colnames(mat) %in% .]} %>% 
  as.data.frame() %>%
  mutate(., 
         visit = unname(vpc)[match(rownames(.), 
                                   names(vpc) %>% 
                                     strsplit("!!") %>% 
                                     sapply(\(x) paste0(x[2],"!!",x[3]))
         )
         ],
         anno = unname(anno.comb)[match(rownames(.),
                                        names(anno.comb) %>% 
                                          strsplit("!!") %>% 
                                          sapply(\(x) paste0(x[2],"!!",x[3]))
         )
         ],
         sample = unname(spc)[match(rownames(.), 
                                    names(spc) %>% 
                                      strsplit("!!") %>% 
                                      sapply(\(x) paste0(x[2],"!!",x[3]))
         )
         ]
  ) %>% 
  filter(!is.na(anno)) %>% 
  reshape2::melt(id.vars = c("visit", "anno", "sample")) %>% 
  mutate(anno_vis = paste0(anno,"_",visit),
         anno_var = paste0(anno,"_",variable)) %>% 
  group_by(visit, anno, variable, sample) %>% 
  summarize(mm = median(value)) %>% 
  ungroup() %>% 
  filter(#variable == "JUN(+)",
         !anno == "Adipocytes") %>% 
  mutate(anno = factor(anno, levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3", "Adipocytes")),
         visit = factor(visit, levels = c("vis1", "vis2", "vis3"), labels = c("Visit 1", "Visit 2", "Visit 3")),
         sex = meta$sex[match(.$sample, meta$sample)],
         variable = factor(variable)) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " ") %>% gsub("Visit", "visit", .),
         anno = factor(anno, levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")))
```

```{r s4h3, fig.height=4, fig.width=8}
plot.dat <- plot.dat.tmp %>% 
  filter(variable == "FOS(+)")

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$mm, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(ct = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(ct = factor(ct, levels = unique(plot.dat$anno))) %>% 
  arrange(ct) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(anno, sex) %>%
  wilcox_test(mm~sex_visit, paired = TRUE) %>%
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05) %>% # Adjust line position
  arrange(desc(y.position))

# Plot
ggplot(plot.dat, aes(anno, mm)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1), 
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Mean AUC", fill = "") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(ct, y = 6, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 2, hide.ns = F) # Wilcoxon
```

```{r s4h4, fig.height=4, fig.width=8}
plot.dat <- plot.dat.tmp %>% 
  filter(variable == "FOSB(+)")

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$mm, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(ct = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(ct = factor(ct, levels = unique(plot.dat$anno))) %>% 
  arrange(ct) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(anno, sex) %>%
  wilcox_test(mm~sex_visit, paired = TRUE) %>%
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05) %>% # Adjust line position
  arrange(desc(y.position))

# Plot
ggplot(plot.dat, aes(anno, mm)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1), 
             color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Mean AUC", fill = "") +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(ct, y = 6, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 2, hide.ns = F) # Wilcoxon
```

# Figure S5

## Load data

```{r s5}
con <- qread("con_adipocytes.qs", nthreads = 10)

spc <- con$getDatasetPerCell()
```

## Figure S5A

```{r s5a}
varToPlot <- spc %>% 
  as.character() %>% 
  strsplit("_vis") %>% 
  sget(1) %>% 
  setNames(names(spc)) %>% 
  factor()

con$plotGraph(groups = varToPlot, 
                   plot.na = F, 
                   size = 0.3,
                   alpha = 0.1, 
                   embedding = "largeVis",
                   mark.groups = F,
                   show.labels = T,
                   show.legend = T,
                   title = "Donor") +
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3)
```

## Figure S5B

```{r s5b}
var <- meta$sex

varToPlot <- grepl.replace(spc %>% 
                             as.character(), 
                           meta$sample, 
                           var) %>% 
  setNames(names(spc)) %>% 
  factor(labels = unique(var))

con$plotGraph(groups = varToPlot, 
                   plot.na = F, 
                   size = 0.3,
                   alpha = 0.1, 
                   embedding = "largeVis",
                   mark.groups = F,
                   show.labels = T,
                   show.legend = T,
                   title = "Sex") +
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3) +
  scale_color_manual(values = c("firebrick", "navy"))
```

## Figure S5C

```{r s5c}
varToPlot <- spc %>% 
  as.character() %>% 
  strsplit("_|!!") %>% 
  sget(3) %>% 
  setNames(names(spc)) %>% 
  factor()

con$plotGraph(groups = varToPlot, 
                   plot.na = F, 
                   size = 0.3,
                   alpha = 0.1, 
                   embedding = "largeVis",
                   mark.groups = F,
                   show.labels = T,
                   show.legend = T,
                   title = "Visit") +
  labs(x = "largeVis1", y = "largeVis2", col = "") + 
  theme(line = element_blank()) +
  dotSize(3) +
  scale_color_manual(values = ggsci::pal_jama()(3))
```

## Figure S5D

Please note, the clustering algorithm holds some randomness, clusters will not always be the same.

```{r s5d, fig.width=8, fig.height=8}
con$findCommunities(name = "leiden1", resolution = 0.5)
con$findCommunities(name = "leiden2", resolution = 0.5)
con$findCommunities(name = "leiden3", resolution = 0.5)
con$findCommunities(name = "leiden4", resolution = 0.5)

seq(4) %>% 
  lapply(\(x) {
    con$plotGraph(embedding = "largeVis", clustering = paste0("leiden",x), font.size = 6) +
  theme(line = element_blank()) +
  scale_color_manual(values = RColorBrewer::brewer.pal(4, "Reds")[-1])
  }) %>% 
  cowplot::plot_grid(plotlist = ., ncol = 2)
```

## Figure S5E

Please note, the clustering algorithm holds some randomness, stabilities will not always be the same.

```{r s5e}
con$findCommunities(resolution = 0.5, test.stability = T, name = "l05stab")

plot_grid(plotlist = list(
  con$plotClusterStability("l05stab", "ari") +
    theme_bw() +
    theme(line = element_blank(),
          axis.text.x = element_blank()) + 
    geom_boxplot(aes(col = "#E41A1C"), notch = T) +
    labs(x = " \n "),
  con$plotClusterStability("l05stab", "hjc") + 
    theme_bw() +
    theme(line = element_blank()) + 
    scale_color_manual(values = RColorBrewer::brewer.pal(4, "Reds")[-1])
), ncol = 2, rel_widths = c(1.3,3))
```

## Figure S5F

Please note, the clustering algorithm holds some randomness, stabilities will not always be the same.

```{r s5f}
con$findCommunities(resolution = 1, test.stability = T, name = "l05stab")

plot_grid(plotlist = list(
  con$plotClusterStability("l05stab", "ari") +
    theme_bw() +
    theme(line = element_blank(),
          axis.text.x = element_blank()) + 
    geom_boxplot(aes(col = "#E41A1C"), notch = T) +
    labs(x = " \n "),
  con$plotClusterStability("l05stab", "hjc") + 
    theme_bw() +
    theme(line = element_blank()) + 
    scale_color_manual(values = RColorBrewer::brewer.pal(8, "Reds")[-1])
), ncol = 2, rel_widths = c(1.3,3))
```

## Figure S5G

```{r s5g}
cm.merged <- con$getJointCountMatrix()

anno.adipocytes <- qread("anno_adipocytes_archetypes.qs")

c("PRSS23","PDE5A","ADIPOQ","DGAT2", "GPAM", "LPL", "DCN", "VIM", "RORA", "CLSTN2", "CRIM1", "ITGA1") %>% 
  sccore::dotPlot(., 
                  cm.merged, 
                  anno.adipocytes, 
                  gene.order = ., 
                  cols = c("white","red3"))
```

## Figure S5H-I

### Preparation

```{r s5hi, fig.width=6, fig.height=4}
cm.merged <- con$getJointCountMatrix(raw = T)

# Create sample-wise annotation
anno.donor <- con$getDatasetPerCell()[rownames(cm.merged)]
anno.subtype <- anno.adipocytes %>% 
  .[!is.na(.)] %>% 
  factor()

idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())

anno.donor %<>% .[idx]
anno.subtype %<>% .[idx] %>% 
  {`names<-`(as.character(.), names(.))}

anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
  `names<-`(anno.donor %>% names())

cm.pseudo.tmp <- sccore::collapseCellsByType(cm.merged,
                                         groups = anno.final, min.cell.count = 1) %>% 
  t() %>%
  apply(2, as, "integer")

cm.pseudo <- cm.pseudo.tmp %>% 
  DESeq2::DESeqDataSetFromMatrix(., 
                                 colnames(.) %>% 
                                   strsplit("_|!!") %>% 
                                   sget(3) %>% 
                                   data.frame() %>% 
                                   `dimnames<-`(list(colnames(cm.pseudo.tmp), "group")), 
                                 design = ~ group) %>% 
  DESeq2::estimateSizeFactors() %>% 
  DESeq2::counts(normalized = T) %>% 
  t()
```

### Figure S5H

```{r s5h, fig.width=6, fig.height=4}
gene = "MAP3K5"

plot.dat <- cm.pseudo %>% 
  .[, colnames(.) %in% gene] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("DGAT2_archetype", "CLSTN2_archetype", "PRSS23_archetype")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>%
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
        color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", title = gene) +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 200, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) + # Wilcoxon
  guides(fill = "none")
```

```{r s5h2, fig.width=6, fig.height=4}
gene = "AKT3"

plot.dat <- cm.pseudo %>% 
  .[, colnames(.) %in% gene] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("DGAT2_archetype", "CLSTN2_archetype", "PRSS23_archetype")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>%
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
        color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", title = gene) +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 100, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) + # Wilcoxon
  guides(fill = "none")
```

### Figure S5I

```{r s5i, fig.width=6, fig.height=4}
gene = "ADAMTS10"

plot.dat <- cm.pseudo %>% 
  .[, colnames(.) %in% gene] %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>% 
  mutate(visit = strsplit(sample, "!!|_") %>% 
           sget(3) %>% 
           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
         anno = strsplit(sample, "!!") %>% 
           sget(2) %>% 
           factor(levels = c("DGAT2_archetype", "CLSTN2_archetype", "PRSS23_archetype")),
         donor = strsplit(sample, "!!") %>% 
           sget(1)) %>% 
  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = " "))

# Kruskal-Wallis
sex.stat <- plot.dat %$% 
  split(., sex) %>% 
  lapply(\(x) split(x, x$anno)) %>% 
  lapply(lapply, \(x) kruskal.test(x$value, g = x$visit)$p.value) %>% 
  lapply(sapply, gtools::stars.pval) %>% 
  {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
  `names<-`(c("Female","Male")) %>% 
  lapply(\(x) data.frame(anno = names(x), sig = unname(x))) %>% 
  {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
  bind_rows() %>% 
  mutate(anno = factor(anno, levels = unique(plot.dat$anno))) %>% 
  arrange(anno) %>% 
  mutate(sig = gsub(".", "", sig, fixed = T))

# Wilcoxon
stat.test <- plot.dat %>%
  dplyr::rename(var = variable) %>% # add_xy... already uses "variable", need to rename
  group_by(anno, sex) %>%
  wilcox_test(value~sex_visit, paired = F) %>%
  mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
  filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
  add_xy_position(x = "anno", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05)

plot.dat %>% 
  ggplot(aes(anno, value)) + 
  geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
  geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1),
        color = "black", size = 0.5, alpha = 0.2) +
  theme_bw() +
  labs(x = "", y = "Normalized pseudobulk expression", title = gene) +
  theme(line = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(color = "black")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  geom_text(inherit.aes = F, data = sex.stat, aes(anno, y = 200, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 3, hide.ns = F) + # Wilcoxon
  guides(fill = "none")
```

# Figure S6

## Figure S6A-C

### Preparations

We downsampled to 15k nuclei due to calculation times. We provide `sds_obj_reduced.qs` for slingshot pseudotime estimations for the downsampled dataset.

Here, we calculate for visit 1 and visit 3.

```{r s6ac, eval = F}
sds_obj <- qread("sds_obj_reduced.qs")

pseudotime <- sds_obj@assays@data@listData$pseudotime[, 1] %>% 
  .[!is.na(.)]

con <- qread("con_adipocytes_aspc.qs", nthreads = 10)

mat <- con$getJointCountMatrix() %>%
  .[match(names(pseudotime), rownames(.)), colSums(.) > 2e2]

mt <- mat %>%
  as.matrix() %>%
  as.data.frame() %>%
  tibble::rownames_to_column() %>%
  {message("Mutating"); mutate(.,
                               pseudotime = pseudotime[rowname],
                               visit = con$getDatasetPerCell()[rowname] %>%
                                 as.character() %>%
                                 strsplit("_") %>%
                                 sget(3),
                               donor = strsplit(rowname, "!!") %>%
                                 sget(1))} %>%
  .[complete.cases(.), ] %>%
  {message("Melting"); reshape2::melt(., id.vars = c("rowname", "pseudotime", "visit", "donor"))} %$%
  {message("Splitting"); split(., variable)}
```

The calculations take app. 11 h with 14k genes and 10k cells. We've added pre-calculated results as `pseudotime_res_novis2.qs`.

```{r s6ac2, eval = F}
res <- mt %>% 
  sccore::plapply(\(y) {
    x = y %>% 
      filter(!visit == "vis2")

    fit_full <- gamm4(data = x, formula = value ~ s(pseudotime, by = factor(visit)), random = ~(1 | donor), REML = F)
    fit_reduced <- gamm4(data = x, formula = value ~ s(pseudotime), random = ~(1 | donor), REML = F)
    fit_none <- gamm4(data = x, formula = value ~ 1, random = ~(1 | donor), REML = F)

    ann <- anova(fit_full$mer, fit_reduced$mer, fit_none$mer)

    if (any(ann$`Pr(>Chisq)` <= 0.05, na.rm = T)) {
      residuals <- predict(fit_full$gam, se.fit = T)$fit %>% 
        unname()

      r.sq <- c(summary(fit_full$gam)$r.sq, summary(fit_reduced$gam)$r.sq) %>% 
        setNames(c("full", "reduced"))

      out <- list(annova = ann,
                  residuals = residuals,
                  r.sq = r.sq)

      return(out)
    }
  }, n.cores = 5, mc.preschedule = T, mc.cleanup = T, progress = F) # 10 cores fails, 5 is max tested that finishes
```

### Figure S6A,B

Load data. For ease, we also provide pseudotime W/O visit 2 nuclei. We need the `sds_obj` from Figure 6A. Also, for ease of use we provide slingshot pseudotime W/O visit 2 as `pseudotime_novis2.qs`.

```{r s6ab}
res <- qread("pseudotime_res_novis2.qs", nthreads = 10) %>% 
  .[!sapply(., is.null)]

sds_obj <- qread("sds_obj.qs")

pseudotime <- qread("pseudotime_novis2.qs")

con <- qread("con_adipocytes_aspc.qs", nthreads = 10)
```

```{r s6ab2}
both <- res %>%
  sapply(\(gene) {
    ann <- gene$annova
    (ann[2, "Pr(>Chisq)"] <= 0.05 & ann[3, "Pr(>Chisq)"] <= 0.05)
  })

res.both <- res[both]

rsq.both <- res.both %>% 
  sapply(\(gene) gene$r.sq[1]) %>% # Ranking by r2 for pseudotime
  sort(decreasing = T)

res.filter <- res.both %>% 
  .[rsq.both %>% .[. > 0.2] %>% names() %>% gsub(".full", "", .)]

cm.merged <- con$getJointCountMatrix() %>% 
  .[match(names(pseudotime), rownames(.)), colnames(.) %in% names(res.filter)] %>% 
  .[rowSums(.) > 0, ]

## Predict smoothend expression 
pseudotime.both <- pseudotime[rownames(cm.merged)]
weights.both <- sds_obj@assays@data@listData$weights %>% .[match(names(pseudotime.both), rownames(.)), 1] + 1E-7

scFit <- cm.merged %>% 
  Matrix::t() %>% 
  tradeSeq::fitGAM(pseudotime = pseudotime.both, cellWeights = weights.both, verbose = T)

Smooth <- tradeSeq::predictSmooth(scFit, gene = colnames(cm.merged), tidy = F, n=100)

# Average across replicates and scale
Smooth <- t(scale(t(Smooth)))

# Seriate the results
Smooth <- Smooth[ seriation::get_order(seriation::seriate(Smooth, method="PCA_angle")), ]

# Omit MALAT1
Smooth %<>% .[!rownames(.) == "MALAT1", ]

# Create heatmap
gene.subset <- which(rownames(Smooth) %in% c("PDGFRB", "DCN", "FBN1", "LPL", "ADIPOQ"))
labels <- rownames(Smooth)[gene.subset]
lannot <- rowAnnotation(link = anno_mark(at = gene.subset,
                                         labels = labels,
                                         labels_gp = grid::gpar(fontsize = 7)))

col_fun = circlize::colorRamp2(c(-4, 0, 4), c("navy", "white", "firebrick"))

Heatmap(Smooth, 
             col = col_fun,
             cluster_columns=F, 
             cluster_rows=F, 
             show_column_names = F, 
             show_row_names = F,
             left_annotation = lannot)
```

```{r s6ab3}
repressed.start <- 1
tempinduced.start = which(rownames(Smooth) == "SMOC2")
tempinduced.end <- which(rownames(Smooth) == "APOD")
repressed.end <- tempinduced.start - 1
induced.start <- tempinduced.end + 1
induced.end <- nrow(Smooth)

go.list <- Smooth %>% 
  rownames() %>% 
  {
    list(
      repressed = .[repressed.start : repressed.end],
      tempinduced = .[tempinduced.start : tempinduced.end],
      induced = .[induced.start : induced.end]
    )
  } %>% 
  lapply(\(x) clusterProfiler::enrichGO(x, 
                                OrgDb = org.Hs.eg.db::org.Hs.eg.db, 
                                keyType = "SYMBOL", 
                                ont = "BP", 
                                universe = rownames(con$samples[[1]])))

go.list %>% 
  lapply(getElement, "result") %>% 
  lapply(head, 10)
```

### Figure S6C

Figure S6A needs to be calculated first. We obtain list with human transcription factors from public resource.

```{r s6c}
tfs <- read.table("https://humantfs.ccbr.utoronto.ca/download/v_1.01/TF_names_v_1.01.txt") %>% 
  pull(V1)

tf.filter <- intersect(names(res.filter), tfs)

cm.merged <- con$getJointCountMatrix() %>% 
  .[match(names(pseudotime), rownames(.)), colnames(.) %in% tf.filter] %>% 
  .[rowSums(.) > 0, ]

## Predict smoothend expression 
pseudotime.both <- pseudotime[rownames(cm.merged)]
weights.both <- sds_obj@assays@data@listData$weights %>% .[match(names(pseudotime.both), rownames(.)), 1] + 1E-7

scFit <- cm.merged %>% 
  Matrix::t() %>% 
  tradeSeq::fitGAM(pseudotime = pseudotime.both, cellWeights = weights.both, verbose = T)

Smooth <- tradeSeq::predictSmooth(scFit, gene = colnames(cm.merged), tidy = F, n=100)

# Average across replicates and scale
Smooth <- t(scale(t(Smooth)))

# Seriate the results
Smooth <- Smooth[ seriation::get_order(seriation::seriate(Smooth, method="PCA_angle")), ]

# Omit MALAT1
Smooth %<>% .[!rownames(.) == "MALAT1", ]

col_fun = circlize::colorRamp2(c(-4, 0, 4), c("navy", "white", "firebrick"))

Heatmap(Smooth, 
             col = col_fun,
             cluster_columns=F, 
             cluster_rows=F, 
             show_column_names = F, 
             row_names_gp = grid::gpar(fontsize = 10)
             )
```

## Figure S6D

We need the `sds_obj` object from Figure 6A.

```{r s6d, fig.height = 16, fig.width = 16}
con <- qread("con_adipocytes_aspc.qs", nthreads = 10)
spc <- con$getDatasetPerCell()
vpc <- spc %>% 
  as.character() %>% 
  strsplit("_|!!") %>% 
  sget(3) %>% 
  setNames(names(spc))

anno.adipocytes <- qread("anno_major.qs") %>% 
  .[. == "Adipocytes"] %>% 
  factor()
anno.aspc <- qread("anno_aspc.qs")
anno <- factor(c(anno.adipocytes, anno.aspc))

cm.merged <- con$getJointCountMatrix()
```

```{r s6d2}
mat.df <- c("ZEB1", "EBF2", "PPARG") %>%
  {cm.merged[, colnames(cm.merged) %in% .]} %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  mutate(., 
         visit = unname(vpc)[match(rownames(.), 
                                   names(vpc)
                                   )
                             ] %>% 
           factor(labels = c("Visit 1", "Visit 2", "Visit 3")),
         anno = unname(anno)[match(rownames(.),
                                   names(anno)
                                   )
                             ],
         sample = unname(spc)[match(rownames(.), 
                                   names(spc)
                                   )
                             ]
         ) %>% 
  filter(!is.na(anno))

pseudotime_all <- qread("sds_obj.qs", nthreads = 10)@assays@data@listData$pseudotime %>% 
  as.data.frame() %>% 
  filter(!is.na(Lineage1)) %>% 
  {setNames(.$Lineage1, rownames(.))}

mat.df %<>% .[match(names(pseudotime_all), rownames(.), nomatch = F), ]

pseudotime_all %<>% .[rownames(mat.df)]

mat.df2 <- mat.df %>% 
  mutate(pseudotime = pseudotime_all) %>% 
  mutate(bin = cut(pseudotime, breaks = c(0, 0.005, 0.01, 0.0165, 0.023))) %>% 
  na.omit() %>% 
  mutate(bin = as.character(bin) %>% 
           strsplit(",") %>% 
           sget(2) %>% 
           gsub("]", "", ., fixed = T)) %>%
  select(-anno, -pseudotime) %>%
  reshape2::melt(id.vars = c("visit", "bin", "sample")) %>% 
  filter(value > 0) %>% 
  group_by(visit, variable, bin, sample) %>% 
  mutate(value = as.numeric(value)) %>% 
  summarize(mm = mean(value)) %>% 
  mutate(sex = meta$sex[match(sample, meta$sample)]) %>% 
  mutate(sex_visit = paste(sex, gsub("Visit", "visit", visit), sep = " "))
```

```{r s6d3, fig.width=16, fig.height=4}
mat.df2 %>% 
  pull(variable) %>% 
  unique() %>% 
  as.character() %>% 
  lapply(\(vv) {
    plot.dat <- mat.df2 %>% 
      filter(variable == vv) %>%
    ungroup() %>%
      mutate(sex = meta$sex[match(.$sample, meta$sample)]) %>% 
      mutate(sex_visit = paste(sex, visit, sep = " ") %>% gsub("Visit", "visit", .))

    # Kruskal-Wallis
    sex.stat <- plot.dat %$% 
      split(., sex) %>% 
      lapply(\(x) split(x, x$bin)) %>% 
      lapply(lapply, \(x) kruskal.test(x$mm, g = x$visit)$p.value) %>% 
      lapply(sapply, gtools::stars.pval) %>% 
      {lapply(seq(length(.)), \(x) gsub("*", c("$","#")[x], .[[x]], fixed = T))} %>% 
      `names<-`(c("Female","Male")) %>% 
      lapply(\(x) data.frame(ct = names(x), sig = unname(x))) %>% 
      {lapply(names(.), \(nn) mutate(.[[nn]], sex = nn))} %>% 
      bind_rows() %>% 
      mutate(ct = factor(ct, levels = unique(plot.dat$bin))) %>% 
      arrange(ct) %>% 
      mutate(sig = gsub(".", "", sig, fixed = T))

    # Wilcoxon
    stat.test <- plot.dat %>%
      group_by(bin, sex) %>%
      wilcox_test(mm~sex_visit, paired = F) %>% 
      filter(p.adj.signif != "ns", !is.na(p)) %>% # Remove those not significant for KW
      add_xy_position(x = "bin", dodge = 0.8, scales = "free_y", fun = "mean_sd", step.increase = 0.05) %>% # Adjust line position
      arrange(desc(y.position))

    # Plot
    ggplot(plot.dat, aes(bin, mm)) + 
      geom_boxplot(aes(fill = sex_visit), position = position_dodge(), outlier.shape = NA) +
      geom_point(aes(fill = sex_visit), position = position_jitterdodge(jitter.width = 0.1), 
                 color = "black", size = 0.5, alpha = 0.2) +
      theme_bw() +
      labs(x = "", y = "Mean AUC", fill = "", title = vv) +
      theme(line = element_blank(),
            axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
            axis.text = element_text(color = "black")) +
      scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
      geom_text(inherit.aes = F, data = sex.stat, aes(ct, y = 1.4, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 0.2, hide.ns = F) # Wilcoxon
  }) %>% 
  cowplot::plot_grid(plotlist = ., ncol = 3)
```

# Session info

Time to knit

```{r}
Sys.time() - tt
```

```{r}
sessionInfo()
```

### 2024-11-18 Objects_preparations.Rmd
Initial commit

---
title: "Objects preparation"
author: "Rasmus Rydbirk"
date: "2024-11-11"
output: 
  html_document:
    toc: True
    toc_float: True
---

# Setup

```{r setup, message=FALSE}
library(pagoda2)
library(conos)
library(qs)
library(dplyr)
library(magrittr)
library(scHelper) # github.com/rrydbirk/scHelper
```

# Conos objects

## Load annotation

```{r}
anno.major <- qread("anno_major.qs")
```

## con_major

```{r, eval = F}
con.major <- append(cms.preprocessed.vis1, 
                    cms.preprocessed.vis2) %>% 
  append(cms.preprocessed.vis3) %>% 
  Conos$new(n.cores = 32)

con.major$buildGraph()
con.major$findCommunities()
con.major$embedGraph(method = "UMAP", embedding.name = "UMAP_refined7", min.dist = 0.0001, n.neighbors = 120, spread = 5)
con.major$embedGraph(method = "largeVis", embedding.name = "largeVis")

qsave("con_major.qs", nthreads = 10)
```

## con_adipocytes

```{r, eval = F}
con <- con.major$samples %>% 
  lapply('[[', "misc") %>% 
  lapply('[[', "rawCounts") %>% 
  lapply(Matrix::t) %>% 
  lapply(\(cm) cm[, colnames(cm) %in% names(anno.major)[anno.major == "Adipocytes"]]) %>% 
  lapply(basicP2proc, make.geneknn = F, get.tsne = F, get.largevis = F, n.cores = 32) %>% 
  Conos$new(n.cores = 32)

con$buildGraph()
con$findCommunities()
con$embedGraph(method = "UMAP", embedding.name = "UMAP")
con$embedGraph(method = "largeVis", embedding.name = "largeVis")
con$findCommunities(resolution = 0.5, name = "leiden0.5")
con$findCommunities(resolution = 1.5, name = "leiden1.5")

qsave("con_adipocytes.qs", nthreads = 10)
```

## con_adipocytes_aspc_macrophages

```{r, eval = F}
anno.immune <- qread("anno_immune.qs") %>% 
  .[. %in% c("ATM", "Early LAM", "LAM", "mono/mac")]

anno.aspc <- qread("anno_aspc.qs")

anno <- qread("anno_major_final.qs") %>% 
  .[!names(.) %in% names(anno.immune)] %>% 
  .[!names(.) %in% names(anno.aspc)] %>% 
  {factor(c(., anno.immune, anno.aspc))}

cts <- c("ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3", "ASPC_DPP4", "Adipocytes", "ATM","Early LAM", "LAM", "mono/mac")

anno <- anno[anno %in% cts] %>% 
  factor()

qsave(anno, "anno_adipocytes_aspc_macrophages.qs")

con <- con.major$samples %>% 
  lapply('[[', "misc") %>% 
  lapply('[[', "rawCounts") %>% 
  lapply(Matrix::t) %>% 
  lapply(\(cm) cm[, colnames(cm) %in% names(anno)[anno %in% cts]])

con %<>% lapply(basicP2proc, make.geneknn = F, get.tsne = F, get.largevis = F, n.cores = 32) %>% 
  Conos$new(n.cores = 32)

con$buildGraph()
con$findCommunities()
con$embedGraph(method = "UMAP", embedding.name = "UMAP")
con$embedGraph(method = "largeVis", embedding.name = "largeVis")

qsave(con, "con_adipocytes_aspc_macrophages.qs")
```

## con_aspc

```{r, eval = F}
con <- con.major$samples %>% 
  lapply('[[', "misc") %>% 
  lapply('[[', "rawCounts") %>% 
  lapply(Matrix::t) %>% 
  lapply(\(cm) cm[, colnames(cm) %in% names(anno.major)[anno.major == "ASPC"]]) %>% 
  lapply(basicP2proc, make.geneknn = F, get.tsne = F, get.largevis = F, n.cores = 32) %>% 
  Conos$new(n.cores = 32)

con$buildGraph()
con$findCommunities()
con$embedGraph(method = "UMAP", embedding.name = "UMAP")
con$embedGraph(method = "largeVis", embedding.name = "largeVis")
con$findCommunities(resolution = 0.5, name = "leiden0.5")
con$findCommunities(resolution = 1.5, name = "leiden1.5")

qsave(con, "con_aspc.qs", nthreads = 10)
```

## con_myeloid

```{r, eval = F}
con <- con.major$samples %>% 
  lapply('[[', "misc") %>% 
  lapply('[[', "rawCounts") %>% 
  lapply(Matrix::t) %>% 
  lapply(\(cm) cm[, colnames(cm) %in% names(anno.major)[anno.major == "Myeloid immune cells")]]) %>% 
  lapply(basicP2proc, make.geneknn = F, get.tsne = F, get.largevis = F, n.cores = 32) %>% 
  Conos$new(n.cores = 32)

con$buildGraph()
con$findCommunities()
con$embedGraph(method = "UMAP", embedding.name = "UMAP")
con$embedGraph(method = "largeVis", embedding.name = "largeVis")
con$findCommunities(resolution = 0.5, name = "leiden0.5")
con$findCommunities(resolution = 1.5, name = "leiden1.5")

qsave(con, "con_myeloid.qs", nthreads = 10)
```

## con_lymphoid

```{r, eval = F}
con <- con.major$samples %>% 
  lapply('[[', "misc") %>% 
  lapply('[[', "rawCounts") %>% 
  lapply(Matrix::t) %>% 
  lapply(\(cm) cm[, colnames(cm) %in% names(anno.major)[anno.major == "Lymphoid immune cells"]]) %>% 
  .[sapply(., ncol) > 30] %>% 
  lapply(basicP2proc, make.geneknn = F, get.tsne = F, get.largevis = F, n.cores = 32, nPcs = 20) %>% 
  Conos$new(n.cores = 32)

con$buildGraph()
con$findCommunities()
con$embedGraph(method = "UMAP", embedding.name = "UMAP")
con$embedGraph(method = "largeVis", embedding.name = "largeVis")
con$findCommunities(resolution = 0.5, name = "leiden0.5")
con$findCommunities(resolution = 1.5, name = "leiden1.5")

qsave(con, "con_lymphoid.qs", nthreads = 10)
```

## con_immune

```{r, eval = F}
con <- con.major$samples %>% 
  lapply('[[', "misc") %>% 
  lapply('[[', "rawCounts") %>% 
  lapply(Matrix::t) %>% 
  lapply(\(cm) cm[, colnames(cm) %in% names(anno.major)[anno.major %in% c("Lymphoid immune cells", "Myeloid immune cells")]]) %>% 
  lapply(basicP2proc, make.geneknn = F, get.tsne = F, get.largevis = F, n.cores = 32) %>% 
  Conos$new(n.cores = 32)

con$buildGraph()
con$findCommunities()
con$embedGraph(method = "UMAP", embedding.name = "UMAP")
con$embedGraph(method = "largeVis", embedding.name = "largeVis")
con$findCommunities(resolution = 0.5, name = "leiden0.5")
con$findCommunities(resolution = 1.5, name = "leiden1.5")

qsave(con, "con_immune.qs", nthreads = 10)
```

## con_vascular

```{r, eval = F}
con <- con.major$samples %>% 
  lapply('[[', "misc") %>% 
  lapply('[[', "rawCounts") %>% 
  lapply(Matrix::t) %>% 
  lapply(\(cm) cm[, colnames(cm) %in% names(anno.major)[anno.major %in% c("ECs", "SMCs", "Lymphatic ECs", "Pericytes")]]) %>% 
  lapply(basicP2proc, make.geneknn = F, get.tsne = F, get.largevis = F, n.cores = 32, nPcs = 50) %>% 
  Conos$new(n.cores = 32)

con$buildGraph()
con$findCommunities()
con$embedGraph(method = "UMAP", embedding.name = "UMAP")
con$embedGraph(method = "largeVis", embedding.name = "largeVis")
con$findCommunities(resolution = 0.5, name = "leiden0.5")
con$findCommunities(resolution = 1.5, name = "leiden1.5")

qsave(con, "con_vascular.qs", nthreads = 10)
```

# Export data to Python

## aspc_adi_macro objects

```{r}
con <- qread("con_adipocytes_aspc_macrophages.qs", nthreads = 10)
annotation <- qread("anno_adipocytes_aspc_macrophages.qs")

prepareObjectsForPython(con, 
                        annotation,
                        "",
                        "aspc_adi_macro",
                        "largeVis")
```

## major objects

```{r}
con <- qread("con_major.qs", nthreads = 10)
annotation <- qread("anno_major.qs")

prepareObjectsForPython(con, 
                        annotation,
                        "",
                        "major",
                        "UMAP")

# Visits
spc <- con$getDatasetPerCell()

spc %>%
  {data.frame(
    visit = unname(.) %>% 
      as.character() %>% 
      grepl.replace(patterns = c("vis1","vis2","vis3"), c("visit1","visit2","visit3")), 
    cellids = names(.))} %>% 
  write.table("all_cells.visits", 
              sep = ",", 
              dec = ".", 
              row.names = F, 
              col.names = F)
```

## vascular objects

```{r}
con <- qread("con_vascular.qs", nthreads = 10)
annotation <- qread("anno_vascular.qs")

prepareObjectsForPython(con, 
                        annotation,
                        "",
                        "vascular",
                        "UMAP")
```

### 2024-11-18 Objects_preparations.Rmd
Updated `Objects_preparations.Rmd`

@@
 
+The easiest way to extract the list with raw count matrices is to extract it from the Conos object that we provide for all cells. We do this first and then preprocess.
+
+```{r}
+con.major <- qread("con_major.qs", nthreads = 10)
+
+cms.preprocessed <- con.major$samples %>% 
+  lget("misc") %>% 
+  lget("rawCounts") %>% 
+  lapply(basicP2proc, make.geneknn = F, get.tsne = F, get.largevis = F, n.cores = 32)
+```
+
+Then, we can re-create the Conos object. Please note, we provide additional embeddings in the prepared object that you'll lose if you save this object. Also, the embedding may not be exactly the same.
+
+After the next step, we use the major object to create objects for subpopulations.
+
 ```{r, eval = F}
-con.major <- append(cms.preprocessed.vis1, 
-                    cms.preprocessed.vis2) %>% 
-  append(cms.preprocessed.vis3) %>% 
+con.major <- cms.preprocessed %>%  
   Conos$new(n.cores = 32)
@@
 con.major$embedGraph(method = "UMAP", embedding.name = "UMAP_refined7", min.dist = 0.0001, n.neighbors = 120, spread = 5)
-con.major$embedGraph(method = "largeVis", embedding.name = "largeVis")
+con.major$embedGraph(method = "largeVis", embedding.name = "largeVis") # Not necessary to run, we used the UMAP embedding
 

### 2024-11-26 Manuscript_figures.Rmd
Updated R figures R Markdown and HTML report

@@
 library(sccore)
+library(corrplot)
 
@@
 plot_grid(plotlist = list(p.high, p.low), rel_widths = c(1,4.5))
+```
+
+## Figure 3D
+
+Load data
+
+```{r, f3d}
+## Plot Individual genes from RNA-seq data
+Bulk_norm_counts <- read.delim("Bulk_norm_counts.txt", h=T)
+colData <- read.delim("Meta_Data_WL_Select.txt", h=T, dec = ",")
+```
+
+TREM2
+
+```{r, f3d2}
+gene_of_interest <- "TREM2"
+
+# Subset norm_counts to get counts for the gene of interest
+gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
+rownames(gene_counts)=gene_counts$hgnc_symbol
+gene_counts=gene_counts[4:96]
+
+# Reshape gene_counts into long format and merge with colData
+gene_counts_long <- gene_counts %>%
+  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
+  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")
+
+gene_counts_long <- gene_counts_long %>%
+  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData
+
+# Perform a paired ANOVA separately for each gender
+anova_res_F <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "F"))
+anova_res_M <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "M"))
+
+# Extract ANOVA p-values
+anova_pval_F <- summary(anova_res_F)[[2]][[1]][["Pr(>F)"]][1]
+anova_pval_M <- summary(anova_res_M)[[2]][[1]][["Pr(>F)"]][1]
+
+# Perform paired Wilcoxon tests for each sex
+wilcox_res <- list(
+  visit1_vs_visit2_F = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
+    paired = TRUE
+  ),
+  visit1_vs_visit3_F = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
+    paired = TRUE
+  ),
+  visit2_vs_visit3_F = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
+    paired = TRUE
+  ),
+  visit1_vs_visit2_M = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
+    paired = TRUE
+  ),
+  visit1_vs_visit3_M = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
+    paired = TRUE
+  ),
+  visit2_vs_visit3_M = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
+    paired = TRUE
+  )
+)
+
+wilcox_pvals <- sapply(wilcox_res, function(x) x$p.value)
+
+# Create the boxplot with significance annotations
+plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
+  geom_boxplot(outlier.shape = NA) +  # Remove outliers
+  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
+  #geom_line(aes(group = Patient), color = "grey", size = 0.3, linetype = "solid", alpha = 0.6) +  # Connect paired values with thin lines
+  labs(title = gene_of_interest, x = "Visit and Sex", y = "Gene Count") +
+  theme_minimal() +
+  geom_signif(
+    comparisons = list(
+      c("visit_1.F", "visit_2.F"), 
+      c("visit_1.F", "visit_3.F"), 
+      c("visit_2.F", "visit_3.F"),
+      c("visit_1.M", "visit_2.M"), 
+      c("visit_1.M", "visit_3.M"), 
+      c("visit_2.M", "visit_3.M")
+    ),
+    map_signif_level = TRUE,
+    test = "wilcox.test",
+    test.args = list(paired = TRUE)
+  ) +
+  theme(plot.title = element_text(size = 10))
+
+# Add ANOVA p-values to the plot
+plot <- plot + annotate("text", x = 1, y = max(gene_counts_long$count), label = paste("ANOVA F p-value:", format(anova_pval_F, digits = 3)), hjust = 0, vjust = 1, size = 3)
+plot <- plot + annotate("text", x = 4, y = max(gene_counts_long$count), label = paste("ANOVA M p-value:", format(anova_pval_M, digits = 3)), hjust = 0, vjust = 1, size = 3)
+
+# Display the plot
+print(plot)
+```
+
+SPP1
+
+```{r, f3d3}
+gene_of_interest <- "SPP1"
+
+# Subset norm_counts to get counts for the gene of interest
+gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
+rownames(gene_counts)=gene_counts$hgnc_symbol
+gene_counts=gene_counts[4:96]
+
+# Reshape gene_counts into long format and merge with colData
+gene_counts_long <- gene_counts %>%
+  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
+  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")
+
+gene_counts_long <- gene_counts_long %>%
+  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData
+
+# Perform a paired ANOVA separately for each gender
+anova_res_F <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "F"))
+anova_res_M <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "M"))
+
+# Extract ANOVA p-values
+anova_pval_F <- summary(anova_res_F)[[2]][[1]][["Pr(>F)"]][1]
+anova_pval_M <- summary(anova_res_M)[[2]][[1]][["Pr(>F)"]][1]
+
+# Perform paired Wilcoxon tests for each sex
+wilcox_res <- list(
+  visit1_vs_visit2_F = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
+    paired = TRUE
+  ),
+  visit1_vs_visit3_F = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
+    paired = TRUE
+  ),
+  visit2_vs_visit3_F = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
+    paired = TRUE
+  ),
+  visit1_vs_visit2_M = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
+    paired = TRUE
+  ),
+  visit1_vs_visit3_M = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
+    paired = TRUE
+  ),
+  visit2_vs_visit3_M = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
+    paired = TRUE
+  )
+)
+
+wilcox_pvals <- sapply(wilcox_res, function(x) x$p.value)
+
+# Create the boxplot with significance annotations
+plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
+  geom_boxplot(outlier.shape = NA) +  # Remove outliers
+  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
+  #geom_line(aes(group = Patient), color = "grey", size = 0.3, linetype = "solid", alpha = 0.6) +  # Connect paired values with thin lines
+  labs(title = gene_of_interest, x = "Visit and Sex", y = "Gene Count") +
+  theme_minimal() +
+  geom_signif(
+    comparisons = list(
+      c("visit_1.F", "visit_2.F"), 
+      c("visit_1.F", "visit_3.F"), 
+      c("visit_2.F", "visit_3.F"),
+      c("visit_1.M", "visit_2.M"), 
+      c("visit_1.M", "visit_3.M"), 
+      c("visit_2.M", "visit_3.M")
+    ),
+    map_signif_level = TRUE,
+    test = "wilcox.test",
+    test.args = list(paired = TRUE)
+  ) +
+  theme(plot.title = element_text(size = 10))
+
+# Add ANOVA p-values to the plot
+plot <- plot + annotate("text", x = 1, y = max(gene_counts_long$count), label = paste("ANOVA F p-value:", format(anova_pval_F, digits = 3)), hjust = 0, vjust = 1, size = 3)
+plot <- plot + annotate("text", x = 4, y = max(gene_counts_long$count), label = paste("ANOVA M p-value:", format(anova_pval_M, digits = 3)), hjust = 0, vjust = 1, size = 3)
+
+# Display the plot
+print(plot)
+```
+
+MMP9
+
+```{r}
+gene_of_interest <- "MMP9"
+
+# Subset norm_counts to get counts for the gene of interest
+gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
+rownames(gene_counts)=gene_counts$hgnc_symbol
+gene_counts=gene_counts[4:96]
+
+# Reshape gene_counts into long format and merge with colData
+gene_counts_long <- gene_counts %>%
+  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
+  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")
+
+gene_counts_long <- gene_counts_long %>%
+  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData
+
+# Perform a paired ANOVA separately for each gender
+anova_res_F <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "F"))
+anova_res_M <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "M"))
+
+# Extract ANOVA p-values
+anova_pval_F <- summary(anova_res_F)[[2]][[1]][["Pr(>F)"]][1]
+anova_pval_M <- summary(anova_res_M)[[2]][[1]][["Pr(>F)"]][1]
+
+# Perform paired Wilcoxon tests for each sex
+wilcox_res <- list(
+  visit1_vs_visit2_F = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
+    paired = TRUE
+  ),
+  visit1_vs_visit3_F = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
+    paired = TRUE
+  ),
+  visit2_vs_visit3_F = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
+    paired = TRUE
+  ),
+  visit1_vs_visit2_M = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
+    paired = TRUE
+  ),
+  visit1_vs_visit3_M = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
+    paired = TRUE
+  ),
+  visit2_vs_visit3_M = wilcox.test(
+    x = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
+    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
+    paired = TRUE
+  )
+)
+
+wilcox_pvals <- sapply(wilcox_res, function(x) x$p.value)
+
+# Create the boxplot with significance annotations
+plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
+  geom_boxplot(outlier.shape = NA) +  # Remove outliers
+  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
+  #geom_line(aes(group = Patient), color = "grey", size = 0.3, linetype = "solid", alpha = 0.6) +  # Connect paired values with thin lines
+  labs(title = gene_of_interest, x = "Visit and Sex", y = "Gene Count") +
+  theme_minimal() +
+  geom_signif(
+    comparisons = list(
+      c("visit_1.F", "visit_2.F"), 
+      c("visit_1.F", "visit_3.F"), 
+      c("visit_2.F", "visit_3.F"),
+      c("visit_1.M", "visit_2.M"), 
+      c("visit_1.M", "visit_3.M"), 
+      c("visit_2.M", "visit_3.M")
+    ),
+    map_signif_level = TRUE,
+    test = "wilcox.test",
+    test.args = list(paired = TRUE)
+  ) +
+  theme(plot.title = element_text(size = 10))
+
+# Add ANOVA p-values to the plot
+plot <- plot + annotate("text", x = 1, y = max(gene_counts_long$count), label = paste("ANOVA F p-value:", format(anova_pval_F, digits = 3)), hjust = 0, vjust = 1, size = 3)
+plot <- plot + annotate("text", x = 4, y = max(gene_counts_long$count), label = paste("ANOVA M p-value:", format(anova_pval_M, digits = 3)), hjust = 0, vjust = 1, size = 3)
+
+# Display the plot
+print(plot)
 ```
@@
 
-We provide lists with filtered count matrices per visit. Please note, we re-sequenced some samples that are renamed in the following.
+We provide lists with filtered count matrices per visit.
 
 ```{r s1load}
-cms <- c("vis1_cms_filtered.qs",
-         "vis2_cms_filtered.qs",
-         "vis3_cms_filtered.qs") %>% 
-  lapply(qread, nthreads = 10) %>% 
-  mapply(\(x,y) setNames(x, paste0(names(x), "_", y)), x = ., y = c("visit1", "visit2", "visit3")) %>% 
-  unlist() %>% 
-  .[!names(.) %in% c("Donor_25_poolB_visit3","Donor_29_poolA_visit3")] %>% 
-  setNames(gsub("_poolA|_poolB", "", names(.)))
-
 con.major <- qread("con_major.qs", nthreads = 10)
+
+cms <- con.major$samples %>% 
+  lget("misc") %>% 
+  lget("rawCounts") %>% 
+  lapply(Matrix::t)
 
@@
 
+## Figure S1E
+
+```{r}
+WATLAS <- read.delim("watlas_labels_l1_adipose_anno_super.tsv")
+rownames(WATLAS)=WATLAS$X
+WATLAS=WATLAS[2:21]
+
+# Define the order of columns and rows
+column_order <- c("Adipocyte", "FAP", "VEC", "LEC", "SMC","Pericyte", "B.cell", "CD4..T.cell", "CD8..T.cell", "ILC", "NK.cell", "Mast",  
+                  "Macrophage",  "Monocyte", "DC", "pDC",
+                  "Endometrium", "Mesothelial", "Plasmablast", 
+                  "Schwann" )
+
+row_order <- c("Adipocytes", "ASPC", "EC", "lEC", "SMCs", "Pericytes", "B-cells", "T-cells", "Mast cells", "Myeloid")
+WATLAS_ordered <- WATLAS[, rev(column_order)] # Reorder columns
+WATLAS_ordered <- WATLAS_ordered[row_order, ] # Reorder rows
+
+corrplot(t(WATLAS_ordered), method="color", order = , col.lim=c(0, 1),
+         col = rev(COL2("RdBu",200)),
+         is.corr = F, 
+         tl.col	= "black",
+         tl.pos	= "lower",
+         outline = "#C0C0C0",
+         addgrid.col= NA)
+```
+
 # Figure S2
@@
   scale_color_manual(values = ggsci::pal_jama()(3))
+```
+
+## Figure S2D
+
+```{r}
+WATLAS <- read.delim("watlas_labels_l2_adipose_anno.tsv")
+rownames(WATLAS)=WATLAS$X
+
+# Define Vascular cells
+column_order <- c("SOX5..VEC",  "BTNL9..VEC", "ACKR1..VEC", "LEC", "SMC", "Pericyte")
+row_order <- c("arEC", "cEC", "cEC2","vEC", "lEC", "SMCs", "Pericytes")
+WATLAS_ordered <- WATLAS[, rev(column_order)] # Reorder columns
+WATLAS_ordered <- WATLAS_ordered[row_order, ] # Reorder rows
+
+corrplot(t(WATLAS_ordered), method="color", order = , col.lim=c(0, 1),
+         col = rev(COL2("RdBu",200)),
+         is.corr = F, 
+         tl.col	= "black",
+         tl.pos	= "lower",
+         outline = "#C0C0C0",
+         addgrid.col= NA)
 ```
@@
   scale_color_manual(values = ggsci::pal_jama()(3))
+```
+
+## Figure S3K
+
+```{r}
+WATLAS <- read.delim("watlas_labels_l2_adipose_anno.tsv")
+rownames(WATLAS)=WATLAS$X
+
+# Define MACS
+column_order <- c("LYVE1..Macrophage",  "CYP27A1..Macrophage", "LPL..Macrophage")
+row_order <- c("ATM", "mono/mac", "Early LAM", "LAM")
+WATLAS_ordered <- WATLAS[, rev(column_order)] # Reorder columns
+WATLAS_ordered <- WATLAS_ordered[row_order, ] # Reorder rows
+
+corrplot(t(WATLAS_ordered), method="color", order = , col.lim=c(0.5, 1),
+         col = rev(COL2("RdBu",200)),
+         is.corr = F, 
+         tl.col	= "black",
+         tl.pos	= "lower",
+         outline = "#C0C0C0",
+         addgrid.col= NA)
 ```
@@
   scale_color_manual(values = ggsci::pal_jama()(3))
+```
+
+## Figure S4D
+
+```{r}
+FAP_sub <- read.delim("fap.tsv")
+rownames(FAP_sub)=FAP_sub$anno
+FAP_sub=FAP_sub[2:5]
+
+# Define ASPCs 
+column_order <- c("FAP2","FAP1", "FAP4","FAP3")
+row_order <- c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")
+FAP_ordered <- FAP_sub[, rev(column_order)] # Reorder columns
+FAP_ordered <- FAP_ordered[row_order, ] # Reorder rows
+
+corrplot(t(t(FAP_ordered)), method="color", order = , col.lim=c(-1, 1.09),
+         col = rev(COL2("RdBu",200)),
+         is.corr = F, 
+         tl.col	= "black",
+         tl.pos	= "lower",
+         outline = "#C0C0C0",
+         addgrid.col= NA)
+```
+
+## Figure S4E
+
+```{r}
+WATLAS <- read.delim("watlas_labels_l2_adipose_anno.tsv")
+rownames(WATLAS)=WATLAS$X
+
+# Define ASPCs 
+column_order <- c("DPP4..FAP",  "CXCL14..FAP", "PPARG..FAP", "ICAM1..FAP")
+row_order <- c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")
+WATLAS_ordered <- WATLAS[, rev(column_order)] # Reorder columns
+WATLAS_ordered <- WATLAS_ordered[row_order, ] # Reorder rows
+
+corrplot(t(WATLAS_ordered), method="color", order = , col.lim=c(0.4, 1),
+         col = rev(COL2("RdBu",200)),
+         is.corr = F, 
+         tl.col	= "black",
+         tl.pos	= "lower",
+         outline = "#C0C0C0",
+         addgrid.col= NA)
 ```

### 2024-11-26 Manuscript_figures.Rmd
Updated R manuscript figures R Markdown and report

@@
   group_by(anno, sex) %>%
-  wilcox_test(value~sex_visit, paired = F) %>% 
+  wilcox_test(value~sex_visit, paired = T) %>% 
   mutate(., p.adj.signif = apply(., 1, \(x) if(x[10] > 0.05 && x[10] <= 0.1) "." else x[11])) %>% 
@@
       group_by(bin, sex) %>%
-      wilcox_test(mm~sex_visit, paired = F)
+      wilcox_test(mm~sex_visit, paired = T)
 
@@
   group_by(bin, sex) %>%
-  wilcox_test(value~sex_visit, paired = F)
+  wilcox_test(value~sex_visit, paired = T)
 
@@
   geom_text(inherit.aes = F, data = sex.stat, aes(ct, y = 1.5e3, label = sig, group = sex), position = position_dodge(width = 0.9)) + # KW
-  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 0.2, hide.ns = F) # Wilcoxon
+  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01, bracket.nudge.y = 0.2, hide.ns = F)# Wilcoxon
 ```
@@
 
-## Figure S3H-J
+## Figure S3H-J,L
 
@@
 Preparations
+
+Load data for S3H-J first
 

### 2024-12-05 Manuscript_figures.Rmd
Updated manuscript figures

@@
 
-# Perform a paired ANOVA separately for each gender
-anova_res_F <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "F"))
-anova_res_M <- aov(count ~ visit + Error(patient/visit), data = gene_counts_long %>% filter(sex == "M"))
-
-# Extract ANOVA p-values
-anova_pval_F <- summary(anova_res_F)[[2]][[1]][["Pr(>F)"]][1]
-anova_pval_M <- summary(anova_res_M)[[2]][[1]][["Pr(>F)"]][1]
-
-# Perform paired Wilcoxon tests for each sex
-wilcox_res <- list(
-  visit1_vs_visit2_F = wilcox.test(
-    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
-    y = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
+# Define a helper function to map p-values to stars
+pval_to_stars <- function(p) {
+  if (p < 0.0001) {
+    return("****")
+  } else if (p < 0.001) {
+    return("***")
+  } else if (p < 0.01) {
+    return("**")
+  } else if (p < 0.05) {
+    return("*")
+  } else {
+    return("ns")  # Not significant
+  }
+}
+
+# Perform a Kruskal-Wallis test separately for each gender
+kw_res_F <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "F"))
+kw_res_M <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "M"))
+
+# Extract Kruskal-Wallis p-values
+kw_pval_F <- kw_res_F$p.value
+kw_pval_M <- kw_res_M$p.value
+
+# Perform paired Wilcoxon tests for each gender with Holm's correction
+paired_wilcox <- function(data, sex_group) {
+  data_sex <- data %>% filter(sex == sex_group)
+  pairwise_tests <- list(
+    visit1_vs_visit2 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_2") %>% pull(count),
       paired = TRUE
-  ),
-  visit1_vs_visit3_F = wilcox.test(
-    x = gene_counts_long %>% filter(visit == "visit_1", sex == "F") %>% pull(count),
-    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
+    )$p.value,
+    visit1_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
       paired = TRUE
-  ),
-  visit2_vs_visit3_F = wilcox.test(
-    x = gene_counts_long %>% filter(visit == "visit_2", sex == "F") %>% pull(count),
-    y = gene_counts_long %>% filter(visit == "visit_3", sex == "F") %>% pull(count),
+    )$p.value,
+    visit2_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
       paired = TRUE
-  ),
-  visit1_vs_visit2_M = wilcox.test(
-    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
-    y = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
-    paired = TRUE
-  ),
-  visit1_vs_visit3_M = wilcox.test(
-    x = gene_counts_long %>% filter(visit == "visit_1", sex == "M") %>% pull(count),
-    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
-    paired = TRUE
-  ),
-  visit2_vs_visit3_M = wilcox.test(
-    x = gene_counts_long %>% filter(visit == "visit_2", sex == "M") %>% pull(count),
-    y = gene_counts_long %>% filter(visit == "visit_3", sex == "M") %>% pull(count),
-    paired = TRUE
+    )$p.value
   )
-)
-
-wilcox_pvals <- sapply(wilcox_res, function(x) x$p.value)
+  corrected_pvals <- p.adjust(unlist(pairwise_tests), method = "holm")
+  return(corrected_pvals)
+}
+
+wilcox_pvals_F <- paired_wilcox(gene_counts_long, "F")
+wilcox_pvals_M <- paired_wilcox(gene_counts_long, "M")
+
+# Map the Wilcoxon p-values to stars
+wilcox_stars_F <- sapply(wilcox_pvals_F, pval_to_stars)
+wilcox_stars_M <- sapply(wilcox_pvals_M, pval_to_stars)
 
@@
   geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
-  #geom_line(aes(group = Patient), color = "grey", size = 0.3, linetype = "solid", alpha = 0.6) +  # Connect paired values with thin lines
-  labs(title = gene_of_interest, x = "Visit and Sex", y = "Gene Count") +
+  labs(title = "Gene Expression Across Visits and Genders", x = "Visit and Sex", y = "Gene Count") +
   theme_minimal() +
+  theme(plot.title = element_text(size = 10)) +
   geom_signif(
@@
     ),
-    map_signif_level = TRUE,
-    test = "wilcox.test",
-    test.args = list(paired = TRUE)
-  ) +
-  theme(plot.title = element_text(size = 10))
-
-# Add ANOVA p-values to the plot
-plot <- plot + annotate("text", x = 1, y = max(gene_counts_long$count), label = paste("ANOVA F p-value:", format(anova_pval_F, digits = 3)), hjust = 0, vjust = 1, size = 3)
-plot <- plot + annotate("text", x = 4, y = max(gene_counts_long$count), label = paste("ANOVA M p-value:", format(anova_pval_M, digits = 3)), hjust = 0, vjust = 1, size = 3)
+    annotations = c(
+      wilcox_stars_F["visit1_vs_visit2"],
+      wilcox_stars_F["visit1_vs_visit3"],
+      wilcox_stars_F["visit2_vs_visit3"],
+      wilcox_stars_M["visit1_vs_visit2"],
+      wilcox_stars_M["visit1_vs_visit3"],
+      wilcox_stars_M["visit2_vs_visit3"]
+    ),
+    textsize = 4,
+    map_signif_level = FALSE
+  )
+
+# Add Kruskal-Wallis p-values as text
+plot <- plot + 
+  annotate("text", x = 1, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Female p-value:", format(kw_pval_F, digits = 3)), 
+           hjust = 0, size = 3) +
+  annotate("text", x = 4, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Male p-value:", format(kw_pval_M, digits = 3)), 
+           hjust = 0, size = 3)
 
@@
       group_by(bin, sex) %>%
-      wilcox_test(mm~sex_visit, paired = T)
+      wilcox_test(mm~sex_visit, paired = F)
 
@@
   }) %>% 
-  cowplot::plot_grid(plotlist = ., ncol = 3)
+  cowplot::plot_grid(plotlist = ., ncol = 3) -> p
 ```
@@
 
-We provide lists with filtered count matrices per visit.
-
 ```{r s1load}
@@
            sget(3)) %>% 
-  mutate(sample = paste0("Donor", seq(14), "_", visit) %>% 
-           factor(., levels = unique(.))) %>% 
+  mutate(sample = factor(sample, levels = sample)) %>%
   ggplot(aes(sample, value, fill = visit)) +
@@
 # Figure S3
+
+## Figure S3A
+
+Load data
+
+```{r, fs3a}
+## Plot Individual genes from RNA-seq data
+Bulk_norm_counts <- read.delim("Bulk_norm_counts.txt", h=T)
+colData <- read.delim("Meta_Data_WL_Select.txt", h=T, dec = ",")
+```
+
+```{r}
+gene_of_interest <- "TNF"
+
+# Subset norm_counts to get counts for the gene of interest
+gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
+rownames(gene_counts)=gene_counts$hgnc_symbol
+gene_counts=gene_counts[4:96]
+
+# Reshape gene_counts into long format and merge with colData
+gene_counts_long <- gene_counts %>%
+  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
+  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")
+
+gene_counts_long <- gene_counts_long %>%
+  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData
+
+# Define a helper function to map p-values to stars
+pval_to_stars <- function(p) {
+  if (p < 0.0001) {
+    return("****")
+  } else if (p < 0.001) {
+    return("***")
+  } else if (p < 0.01) {
+    return("**")
+  } else if (p < 0.05) {
+    return("*")
+  } else {
+    return("ns")  # Not significant
+  }
+}
+
+# Perform a Kruskal-Wallis test separately for each gender
+kw_res_F <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "F"))
+kw_res_M <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "M"))
+
+# Extract Kruskal-Wallis p-values
+kw_pval_F <- kw_res_F$p.value
+kw_pval_M <- kw_res_M$p.value
+
+# Perform paired Wilcoxon tests for each gender with Holm's correction
+paired_wilcox <- function(data, sex_group) {
+  data_sex <- data %>% filter(sex == sex_group)
+  pairwise_tests <- list(
+    visit1_vs_visit2 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      paired = TRUE
+    )$p.value,
+    visit1_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
+      paired = TRUE
+    )$p.value,
+    visit2_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
+      paired = TRUE
+    )$p.value
+  )
+  corrected_pvals <- p.adjust(unlist(pairwise_tests), method = "holm")
+  return(corrected_pvals)
+}
+
+wilcox_pvals_F <- paired_wilcox(gene_counts_long, "F")
+wilcox_pvals_M <- paired_wilcox(gene_counts_long, "M")
+
+# Map the Wilcoxon p-values to stars
+wilcox_stars_F <- sapply(wilcox_pvals_F, pval_to_stars)
+wilcox_stars_M <- sapply(wilcox_pvals_M, pval_to_stars)
+
+# Create the boxplot with significance annotations
+plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
+  geom_boxplot(outlier.shape = NA) +  # Remove outliers
+  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
+  labs(title = "Gene Expression Across Visits and Genders", x = "Visit and Sex", y = "Gene Count") +
+  theme_minimal() +
+  theme(plot.title = element_text(size = 10)) +
+  geom_signif(
+    comparisons = list(
+      c("visit_1.F", "visit_2.F"), 
+      c("visit_1.F", "visit_3.F"), 
+      c("visit_2.F", "visit_3.F"),
+      c("visit_1.M", "visit_2.M"), 
+      c("visit_1.M", "visit_3.M"), 
+      c("visit_2.M", "visit_3.M")
+    ),
+    annotations = c(
+      wilcox_stars_F["visit1_vs_visit2"],
+      wilcox_stars_F["visit1_vs_visit3"],
+      wilcox_stars_F["visit2_vs_visit3"],
+      wilcox_stars_M["visit1_vs_visit2"],
+      wilcox_stars_M["visit1_vs_visit3"],
+      wilcox_stars_M["visit2_vs_visit3"]
+    ),
+    textsize = 4,
+    map_signif_level = FALSE
+  )
+
+# Add Kruskal-Wallis p-values as text
+plot <- plot + 
+  annotate("text", x = 1, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Female p-value:", format(kw_pval_F, digits = 3)), 
+           hjust = 0, size = 3) +
+  annotate("text", x = 4, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Male p-value:", format(kw_pval_M, digits = 3)), 
+           hjust = 0, size = 3)
+
+# Display the plot
+print(plot)
+```
+
+```{r}
+gene_of_interest <- "CCL3"
+
+# Subset norm_counts to get counts for the gene of interest
+gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
+rownames(gene_counts)=gene_counts$hgnc_symbol
+gene_counts=gene_counts[4:96]
+
+# Reshape gene_counts into long format and merge with colData
+gene_counts_long <- gene_counts %>%
+  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
+  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")
+
+gene_counts_long <- gene_counts_long %>%
+  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData
+
+# Define a helper function to map p-values to stars
+pval_to_stars <- function(p) {
+  if (p < 0.0001) {
+    return("****")
+  } else if (p < 0.001) {
+    return("***")
+  } else if (p < 0.01) {
+    return("**")
+  } else if (p < 0.05) {
+    return("*")
+  } else {
+    return("ns")  # Not significant
+  }
+}
+
+# Perform a Kruskal-Wallis test separately for each gender
+kw_res_F <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "F"))
+kw_res_M <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "M"))
+
+# Extract Kruskal-Wallis p-values
+kw_pval_F <- kw_res_F$p.value
+kw_pval_M <- kw_res_M$p.value
+
+# Perform paired Wilcoxon tests for each gender with Holm's correction
+paired_wilcox <- function(data, sex_group) {
+  data_sex <- data %>% filter(sex == sex_group)
+  pairwise_tests <- list(
+    visit1_vs_visit2 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      paired = TRUE
+    )$p.value,
+    visit1_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
+      paired = TRUE
+    )$p.value,
+    visit2_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
+      paired = TRUE
+    )$p.value
+  )
+  corrected_pvals <- p.adjust(unlist(pairwise_tests), method = "holm")
+  return(corrected_pvals)
+}
+
+wilcox_pvals_F <- paired_wilcox(gene_counts_long, "F")
+wilcox_pvals_M <- paired_wilcox(gene_counts_long, "M")
+
+# Map the Wilcoxon p-values to stars
+wilcox_stars_F <- sapply(wilcox_pvals_F, pval_to_stars)
+wilcox_stars_M <- sapply(wilcox_pvals_M, pval_to_stars)
+
+# Create the boxplot with significance annotations
+plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
+  geom_boxplot(outlier.shape = NA) +  # Remove outliers
+  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
+  labs(title = "Gene Expression Across Visits and Genders", x = "Visit and Sex", y = "Gene Count") +
+  theme_minimal() +
+  theme(plot.title = element_text(size = 10)) +
+  geom_signif(
+    comparisons = list(
+      c("visit_1.F", "visit_2.F"), 
+      c("visit_1.F", "visit_3.F"), 
+      c("visit_2.F", "visit_3.F"),
+      c("visit_1.M", "visit_2.M"), 
+      c("visit_1.M", "visit_3.M"), 
+      c("visit_2.M", "visit_3.M")
+    ),
+    annotations = c(
+      wilcox_stars_F["visit1_vs_visit2"],
+      wilcox_stars_F["visit1_vs_visit3"],
+      wilcox_stars_F["visit2_vs_visit3"],
+      wilcox_stars_M["visit1_vs_visit2"],
+      wilcox_stars_M["visit1_vs_visit3"],
+      wilcox_stars_M["visit2_vs_visit3"]
+    ),
+    textsize = 4,
+    map_signif_level = FALSE
+  )
+
+# Add Kruskal-Wallis p-values as text
+plot <- plot + 
+  annotate("text", x = 1, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Female p-value:", format(kw_pval_F, digits = 3)), 
+           hjust = 0, size = 3) +
+  annotate("text", x = 4, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Male p-value:", format(kw_pval_M, digits = 3)), 
+           hjust = 0, size = 3)
+
+# Display the plot
+print(plot)
+```
+
+```{r}
+gene_of_interest <- "CCL5"
+
+# Subset norm_counts to get counts for the gene of interest
+gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
+rownames(gene_counts)=gene_counts$hgnc_symbol
+gene_counts=gene_counts[4:96]
+
+# Reshape gene_counts into long format and merge with colData
+gene_counts_long <- gene_counts %>%
+  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
+  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")
+
+gene_counts_long <- gene_counts_long %>%
+  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData
+
+# Define a helper function to map p-values to stars
+pval_to_stars <- function(p) {
+  if (p < 0.0001) {
+    return("****")
+  } else if (p < 0.001) {
+    return("***")
+  } else if (p < 0.01) {
+    return("**")
+  } else if (p < 0.05) {
+    return("*")
+  } else {
+    return("ns")  # Not significant
+  }
+}
+
+# Perform a Kruskal-Wallis test separately for each gender
+kw_res_F <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "F"))
+kw_res_M <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "M"))
+
+# Extract Kruskal-Wallis p-values
+kw_pval_F <- kw_res_F$p.value
+kw_pval_M <- kw_res_M$p.value
+
+# Perform paired Wilcoxon tests for each gender with Holm's correction
+paired_wilcox <- function(data, sex_group) {
+  data_sex <- data %>% filter(sex == sex_group)
+  pairwise_tests <- list(
+    visit1_vs_visit2 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      paired = TRUE
+    )$p.value,
+    visit1_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
+      paired = TRUE
+    )$p.value,
+    visit2_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
+      paired = TRUE
+    )$p.value
+  )
+  corrected_pvals <- p.adjust(unlist(pairwise_tests), method = "holm")
+  return(corrected_pvals)
+}
+
+wilcox_pvals_F <- paired_wilcox(gene_counts_long, "F")
+wilcox_pvals_M <- paired_wilcox(gene_counts_long, "M")
+
+# Map the Wilcoxon p-values to stars
+wilcox_stars_F <- sapply(wilcox_pvals_F, pval_to_stars)
+wilcox_stars_M <- sapply(wilcox_pvals_M, pval_to_stars)
+
+# Create the boxplot with significance annotations
+plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
+  geom_boxplot(outlier.shape = NA) +  # Remove outliers
+  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
+  labs(title = "Gene Expression Across Visits and Genders", x = "Visit and Sex", y = "Gene Count") +
+  theme_minimal() +
+  theme(plot.title = element_text(size = 10)) +
+  geom_signif(
+    comparisons = list(
+      c("visit_1.F", "visit_2.F"), 
+      c("visit_1.F", "visit_3.F"), 
+      c("visit_2.F", "visit_3.F"),
+      c("visit_1.M", "visit_2.M"), 
+      c("visit_1.M", "visit_3.M"), 
+      c("visit_2.M", "visit_3.M")
+    ),
+    annotations = c(
+      wilcox_stars_F["visit1_vs_visit2"],
+      wilcox_stars_F["visit1_vs_visit3"],
+      wilcox_stars_F["visit2_vs_visit3"],
+      wilcox_stars_M["visit1_vs_visit2"],
+      wilcox_stars_M["visit1_vs_visit3"],
+      wilcox_stars_M["visit2_vs_visit3"]
+    ),
+    textsize = 4,
+    map_signif_level = FALSE
+  )
+
+# Add Kruskal-Wallis p-values as text
+plot <- plot + 
+  annotate("text", x = 1, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Female p-value:", format(kw_pval_F, digits = 3)), 
+           hjust = 0, size = 3) +
+  annotate("text", x = 4, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Male p-value:", format(kw_pval_M, digits = 3)), 
+           hjust = 0, size = 3)
+
+# Display the plot
+print(plot)
+```
 
@@
 
-```{r s3c}
+```{r s3c, fig.width=7, fig.height=4}
+pdf("/work/02_data/00_prints/S3A.pdf", width = 7, height = 4)
 varToPlot <- spc %>% 
@@
   setNames(names(spc)) %>%
-  factor(labels = paste0("Donor", seq(14))) %>% 
   .[names(.) %in% names(anno.lymphoid)]
@@
   dotSize(3)
+dev.off()
 ```

### 2024-12-16 Manuscript_figures.Rmd
Updated `Manuscript_figures`

@@
   split(diff, sex_visit_anno) %>% 
-  lapply(t.test)
+  lapply(t.test, alternative = "less")
 ```
@@
   }) %>% 
-  cowplot::plot_grid(plotlist = ., ncol = 3) -> p
+  cowplot::plot_grid(plotlist = ., ncol = 3)
 ```

### 2025-04-23 Adipocyte_archetypes.Rmd
Added **Adipocyte_archetypes.Rmd** and updated **README**

---
title: "Adipocyte archetypes"
author: "Rasmus Rydbirk"
date: "'Sys.Date()'"
output: 
  html_document:
    toc: yes
    toc_float: yes
---

# Setup

```{r setup, message=FALSE}
library(archetypes)
library(magrittr)
library(dplyr)
library(qs)
library(scHelper) # github.com/rrydbirk/scHelper
```

# Identify archetypes

This is the code we used to identify archetypes. We set k = 3 based on cluster stability using *findCommunities* in **Conos** with parameter *test.stability = TRUE*. Please note, the results may not be exactly reproducible.

```{r}
con <- qread("con_adipocytes.qs", nthreads = 10)
emb <- con$embeddings$largeVis

res <- lapply(seq(10), \(x) {
  message(paste0("Iteration ",x))
  # Calculate archetypes
  tmp <- archetypes(data = emb, 3, verbose = F)

  # Right
  x <- 0.25
  a <- tmp$alphas %>% 
    as.data.frame() %>% 
    `rownames<-`(tmp$residuals %>% rownames()) %>% 
    filter(V1 < x, V2 > (1-x)) %>% 
    rownames()

  # Left
  x <- 0.30
  b <- tmp$alphas %>% 
    as.data.frame() %>% 
    `rownames<-`(tmp$residuals %>% rownames()) %>% 
    filter(V1 > (1-x), V2 < x) %>% 
    rownames()

  # Top
  x <- 0.25
  c <- tmp$alphas %>% 
    as.data.frame() %>% 
    `rownames<-`(tmp$residuals %>% rownames()) %>% 
    filter(V1 < x, V2 < x) %>% 
    rownames()

  out <- list(a,b,c)

  nn <- out %>% 
    sapply(\(x) unname(anno[x])) %>% 
    lapply(table) %>% 
    lapply(sort, T) %>% 
    lapply(names) %>% 
    sget(1)

  if (any(table(nn) > 1)) out <- NA else names(out) <- nn

  return(out)
}) %>% 
  .[!lapply(res, is.na) %>% sapply(any)]

anno.arc <- c("Adipocytes_RPSS23","Adipocytes_DGAT2","Adipocytes_CLSTN2") %>% 
  `names<-`(lapply(., \(x) sget(res, x) %>% 
                     Reduce(c, .) %>% 
                     table() %>% 
                     sort(T) %>% 
                     .[seq(2e3)] %>% 
                     names()), .) %>% 
  {lapply(seq(3), \(x) `names<-`(rep(names(.)[x], 2e3), .[[x]]))} %>% 
  Reduce(c, .) %>% 
  factor()

anno.arc %<>% factor(labels = c("CLSTN2_archetype", "DGAT2_archetype", "PRSS23_archetype"))

qsave(anno.arc, "anno_adipocytes_archetypes.qs")
```

### 2025-07-07 Manuscript_figures.Rmd
Update for revision

@@
 author: "Rasmus Rydbirk"
-date: "2024-11-11"
+date: "2025-07-07"
 output: 
@@
 library(corrplot)
+library(RColorBrewer)
+library(glmnet)
 
 meta <- qread("meta.qs")
+
+# Palettes
+pal.major <- RColorBrewer::brewer.pal(9, "Set1") %>% 
+  setNames(c("Adipocytes", "Myeloid immune cells", "ASPCs", "ECs", "SMCs", "Lymphoid immune cells", "Mast cells", "Lymphatic ECs", "Pericytes")) %>% 
+  {setNames(c(., ., sapply(unname(.), \(color) {
+    hsv_vals <- rgb2hsv(col2rgb(color))
+    new_hues <- hsv(pmin(1, hsv_vals[1] * 0.9), hsv_vals[2], hsv_vals[3])
+  })), c(names(.), paste0(names(.), ", female"), paste0(names(.), ", male"))
+  )} %>% 
+  {setNames(c(., .[c("Lymphoid immune cells", "Myeloid immune cells")]), c(names(.), c("Lymphoid ICs", "Myeloid ICs")))}
 
@@
 
-## Figure 1G
+## Figures 1b-1f
+
+These figures were made in GraphPad Prism.
+
+## Figure 1g
 
@@
 
-## Figure 1H
+## Figure 1h
 
@@
 
-## Figure 1I
+## Figure 1i
 
@@
 
+## Figure 1k
+
+```{r}
+size_data <- read.csv("Adipocyte_size.csv", sep = ",", header = TRUE)
+sample_colours <- c("#F8B3BB", "#BD6563", "#7D2A25", "#ABD1D4", "#62649E", "#3E4282")
+
+patient_averages <- size_data %>%
+  group_by(ID, Visit, Sex) %>%
+  summarise(Average_Diameter = mean(Diameter, na.rm = TRUE)) %>%
+  ungroup() %>% 
+  mutate(sex_visit = interaction(Visit, Sex) %>% 
+           factor(labels = c("Female Visit 1", "Female Visit 2", "Female Visit 3", "Male Visit 1", "Male Visit 2", "Male Visit 3")))
+
+plot <- ggplot(patient_averages, aes(x = sex_visit, 
+                                     y = Average_Diameter, 
+                                     fill = sex_visit)) +
+  geom_boxplot(position = position_dodge(width = 0.75), outlier.shape = NA) +
+  geom_jitter(size = 1, width = 0.2, alpha = 0.3, show.legend = F) +  # Add individual points
+  theme_bw() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"), 
+    text = element_text(family = "serif", size = 12),
+    line = element_blank(),
+    axis.text.x = element_blank()
+  ) +
+  scale_fill_manual(values = sample_colours) +
+  theme(rect = element_rect(fill = "transparent")) +
+  labs(fill = "", x = "", y = "Diameter (µm)")
+
+plot
+```
+
+## Figure 1m
+
+```{r}
+# Load the CSV file
+CLS <- read.delim("Human_WL_CLS_density_for_plotting.csv", sep = ",")
+
+names(CLS)[names(CLS) == "CLS_density_.number_per_mm3."] <- "CLS_density"
+
+# Define a helper function to map p-values to stars
+pval_to_stars <- function(p) {
+  if (p < 0.0001) {
+    return("****")
+  } else if (p < 0.001) {
+    return("***")
+  } else if (p < 0.01) {
+    return("**")
+  } else if (p < 0.05) {
+    return("*")
+  } else {
+    return("ns")  # Not significant
+  }
+}
+
+# Perform a Kruskal-Wallis test separately for each sex
+kw_res_F <- kruskal.test(CLS_density ~ Visit, data = CLS %>% filter(Sex == "F"))
+kw_res_M <- kruskal.test(CLS_density ~ Visit, data = CLS %>% filter(Sex == "M"))
+
+# Extract Kruskal-Wallis p-values
+kw_pval_F <- kw_res_F$p.value
+kw_pval_M <- kw_res_M$p.value
+
+# Perform paired Wilcoxon tests for each sex with Holm's correction
+paired_wilcox <- function(CLS, sex_group) {
+  data_sex <- CLS %>% filter(Sex == sex_group)
+  pairwise_tests <- list(
+    visit1_vs_visit2 = wilcox.test(
+      x = data_sex %>% filter(Visit == "1") %>% pull(CLS_density),
+      y = data_sex %>% filter(Visit == "2") %>% pull(CLS_density),
+      paired = F
+    )$p.value,
+    visit1_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(Visit == "1") %>% pull(CLS_density),
+      y = data_sex %>% filter(Visit == "3") %>% pull(CLS_density),
+      paired = F
+    )$p.value,
+    visit2_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(Visit == "2") %>% pull(CLS_density),
+      y = data_sex %>% filter(Visit == "3") %>% pull(CLS_density),
+      paired = F
+    )$p.value
+  )
+  corrected_pvals <- p.adjust(unlist(pairwise_tests), method = "holm")
+  return(corrected_pvals)
+}
+
+wilcox_pvals_F <- paired_wilcox(CLS, "F")
+wilcox_pvals_M <- paired_wilcox(CLS, "M")
+
+# Map the Wilcoxon p-values to stars
+wilcox_stars_F <- sapply(wilcox_pvals_F, pval_to_stars)
+wilcox_stars_M <- sapply(wilcox_pvals_M, pval_to_stars)
+
+plot <- ggplot(CLS, aes(x = interaction(Visit, Sex), y = CLS_density, fill = interaction(Visit, Sex))) +
+  geom_boxplot(outlier.shape = NA) +  # Remove outliers
+  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
+  labs(title = "CLS Density Across Visits and Genders", x = "Visit and Sex", y = "# CLS per mm3") +
+  theme_bw() +
+  theme(plot.title = element_text(size = 10)) +
+  geom_signif(
+    comparisons = list(
+      c("1.F", "2.F"), 
+      c("1.F", "3.F"), 
+      c("2.F", "3.F"),
+      c("1.M", "2.M"), 
+      c("1.M", "3.M"), 
+      c("2.M", "3.M")
+    ),
+    annotations = c(
+      wilcox_stars_F["visit1_vs_visit2"],
+      wilcox_stars_F["visit1_vs_visit3"],
+      wilcox_stars_F["visit2_vs_visit3"],
+      wilcox_stars_M["visit1_vs_visit2"],
+      wilcox_stars_M["visit1_vs_visit3"],
+      wilcox_stars_M["visit2_vs_visit3"]
+    ),
+    textsize = 4,
+    map_signif_level = FALSE
+  )
+
+# Add Kruskal-Wallis p-values as text
+plot <- plot + 
+  annotate("text", x = 1, y = max(CLS$CLS_density) * 1.1, 
+           label = paste("KW Female p-value:", format(kw_pval_F, digits = 3)), 
+           hjust = 0, size = 3) +
+  annotate("text", x = 4, y = max(CLS$CLS_density) * 1.1, 
+           label = paste("KW Male p-value:", format(kw_pval_M, digits = 3)), 
+           hjust = 0, size = 3) +
+  scale_fill_manual(values = sample_colours)
+
+plot
+```
+
+## Figure 1o
+
+```{r}
+size_data <- read.csv("2D_histology_averaged_data_output.csv",sep = ",", header = TRUE)
+condition_levels <- c("Female Visit 1", "Female Visit 2", "Female Visit 3", "Male Visit 1", "Male Visit 2", "Male Visit 3")
+
+# 1. Calculate per-patient averages for each visit
+patient_averages <- size_data %>%
+  group_by(Donor, Visit, Sex) %>%
+  summarise(Vascular_Count = mean(Vascular_count_per_area, na.rm = TRUE)) %>%
+  ungroup() %>% 
+  mutate(sex_visit = interaction(Visit, Sex) %>% 
+           factor(labels = c("Female Visit 1", "Female Visit 2", "Female Visit 3", "Male Visit 1", "Male Visit 2", "Male Visit 3")))
+
+# 3. Create the boxplot
+plot <- ggplot(patient_averages, aes(x = sex_visit, y = Vascular_Count, fill = sex_visit)) +
+  geom_boxplot(position = position_dodge(width = 0.75), outlier.shape = NA) +
+  geom_jitter(size = 1, width = 0.2, alpha = 0.3) +  # Add individual points
+  theme_bw() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"), 
+    text = element_text(family = "serif", size = 12),
+    axis.text.x = element_blank(),
+    line = element_blank()
+  ) +
+  scale_fill_manual(values = sample_colours) +
+  labs(x = "", fill = "", y = "Blood vessels per m2")
+
+plot
+```
+
 # Figure 2
@@
 anno.vascular <- qread("anno_vascular.qs")
-```
-
-## Figure 2A
+
+# For 2d,e
+cm.bulk <- read.delim("Bulk_norm_counts.txt", header = T) %>% 
+  .[match(unique(.$hgnc_symbol), .$hgnc_symbol), ] %>% 
+  `rownames<-`(.$hgnc_symbol) %>% 
+  select(-hgnc_symbol, -ensembl_gene_id, -entrezgene_id) %>% 
+  scale()
+
+colnames(cm.bulk) %<>% 
+  gsub("D", "Donor_", .) %>% 
+  gsub("V", "vis", .)
+
+# Metadata
+metadata.all <- read.delim("Meta_Data_WL_Select.txt", h=T, dec = ",") %>% 
+  mutate(sex = factor(sex, labels = c("Female", "Male")),
+         donor = gsub("D", "Donor_", recordid) %>% 
+           gsub("V", "vis", .))
+```
+
+## Figure 2a
 
@@
 
-## Figure 2B
+## Figure 2b
 
@@
 
-## Figure 2D
+## Figure 2c
 
@@
 
+## Figures 2d,e
+
+Preparations
+
+```{r}
+# Use Cacoa to calculate proportions
+cao <- Cacoa$new(data.object = con.major, 
+                 cell.groups = anno.major, 
+                 ref.level = "vis2", 
+                 target.level = "vis3", 
+                 sample.groups = con.major$samples %>% 
+                   names() %>% 
+                   strsplit("_") %>% 
+                   sapply('[[', 3) %>% 
+                   gsub(1, 2, .) %>% 
+                   `names<-`(con.major$samples %>% 
+                               names()),
+                 sample.groups.palette = ggsci::pal_jama()(7))
+
+cao$sample.groups <- con.major$samples %>% 
+  names() %>% 
+  strsplit("_") %>% 
+  sapply('[[', 3) %>% 
+  `names<-`(con.major$samples %>% 
+              names()) %>% 
+  gsub("vis", "Visit ", .)
+
+anno.comb <- anno.major[!names(anno.major) %in% names(anno.vascular)] %>% 
+  {factor(c(., anno.vascular))}
+```
+
+Here, we're showing the preparations to create models for prediction. This takes some time and is not run again.
+
+```{r, eval = F}
+# Get markers
+markers.vascular.all <- con.major$getDifferentialGenes(groups = anno.comb, z.threshold = 1, upregulated.only = T, verbose = T)
+
+# vEC
+# Get proportions
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(vEC, rownames(.))
+
+# Extract markers. Please note, the marker metrics may change a little in each iteration
+markers.all <- markers.vascular.all$vEC %>% 
+  filter(Specificity > 0.9,
+         AUC > 0.65) %>% 
+  arrange(desc(AUC)) %>% 
+  pull(Gene)
+
+# Subset bulk data
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% markers.all) %>% 
+  as.matrix() %>% 
+  t()
+
+# Now, train a final model on all data for visualization and gene selection
+cm.test <- cm.bulk.mat %>% 
+  .[rownames(.) %in% names(proportions), ]
+
+lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 40)
+best_lambda <- lasso_cv$lambda.min
+lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)
+
+# Predict on all samples using the final model
+lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)
+
+# Save model
+list(lasso_cv = lasso_cv,
+     lasso_model = lasso_model) %>%
+  qsave("vEC_model.qs")
+
+# cEC
+## Get proportions
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(cEC, rownames(.))
+
+## Markers
+markers.all <- markers.vacsular.all$cEC %>% 
+  filter(Specificity > 0.9,
+         AUC > 0.6) %>% 
+  arrange(desc(AUC)) %>% 
+  pull(Gene)
+
+## Subset bulk data
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% markers.all) %>% 
+  as.matrix() %>% 
+  t()
+
+# Now, train a final model on all data for visualization and gene selection
+cm.test <- cm.bulk.mat %>% 
+  .[rownames(.) %in% names(proportions), ]
+
+lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 40)
+best_lambda <- lasso_cv$lambda.min
+lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)
+
+# Predict on all samples using the final model
+lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)
+
+# Save model
+list(lasso_cv = lasso_cv,
+     lasso_model = lasso_model) %>%
+  qsave("cEC_model.qs")
+```
+
+### Figure 2d
+
+```{r}
+# Load and prepare
+tmp <- qread("cEC_model.qs")
+
+lasso_model <- tmp$lasso_model
+best_lambda <- tmp$lasso_cv$lambda.min
+
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% rownames(coef(lasso_model))) %>% 
+  as.matrix() %>% 
+  t()
+
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(cEC, rownames(.))
+
+## Predict proportions
+proportions.bulk <-  predict(lasso_model, newx = cm.bulk.mat, s = best_lambda)
+
+## Prepare plot
+plot.df <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+trend.trained <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  filter(donor %in% names(proportions)) %>%
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  group_by(visit, sex) %>% 
+  summarize(s1 = median(s1)) %>%
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+trend.untrained <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  filter(!donor %in% names(proportions)) %>%
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  group_by(visit, sex) %>% 
+  summarize(s1 = median(s1)) %>% 
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+plot.df %>% 
+  ggplot() +
+  geom_boxplot(mapping = aes(sex_visit, s1, fill = sex_visit), outliers = F) +
+  geom_jitter(mapping = aes(sex_visit, s1, col = sex_visit), alpha = 0.3, col = "black", width = 0.3) +
+  theme_bw() +
+  labs(y = "Predicted proportions (%)", title = "Predicted vECs proportions", x = "", col = "Median trends", fill = "") +
+  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Trained data"), data = trend.trained, linewidth = 1) +
+  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Untrained data"), data = trend.untrained, linewidth = 1) +
+  scale_color_manual(values = c("Trained data" = "black", "Untrained data" = "grey40")) +
+  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
+  theme(line = element_blank(),
+        axis.text = element_text(color = "black"),
+        axis.text.x = element_blank())
+```
+
+Statistics
+
+```{r}
+plot.df %>% 
+  filter(sex == "Female") %>% 
+  rstatix::kruskal_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Female") %>% 
+  rstatix::wilcox_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Male") %>% 
+  rstatix::kruskal_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Male") %>% 
+  rstatix::wilcox_test(s1 ~ visit)
+```
+
+### Figure 2e
+
+```{r}
+# Load and prepare
+tmp <- qread("vEC_model.qs")
+
+lasso_model <- tmp$lasso_model
+best_lambda <- tmp$lasso_cv$lambda.min
+
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% rownames(coef(lasso_model))) %>% 
+  as.matrix() %>% 
+  t()
+
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(vEC, rownames(.))
+
+## Predict proportions
+proportions.bulk <-  predict(lasso_model, newx = cm.bulk.mat, s = best_lambda)
+
+plot.df <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+trend.trained <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  filter(donor %in% names(proportions)) %>%
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  group_by(visit, sex) %>% 
+  summarize(s1 = median(s1)) %>%
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+trend.untrained <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  filter(!donor %in% names(proportions)) %>%
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  group_by(visit, sex) %>% 
+  summarize(s1 = median(s1)) %>% 
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+plot.df %>% 
+  ggplot() +
+  geom_boxplot(mapping = aes(sex_visit, s1, fill = sex_visit), outliers = F) +
+  geom_jitter(mapping = aes(sex_visit, s1, col = sex_visit), alpha = 0.3, col = "black", width = 0.3) +
+  theme_bw() +
+  labs(y = "Predicted proportions (%)", title = "Predicted cECs proportions", x = "", col = "Median trends", fill = "") +
+  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Trained data"), data = trend.trained, linewidth = 1) +
+  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Untrained data"), data = trend.untrained, linewidth = 1) +
+  scale_color_manual(values = c("Trained data" = "black", "Untrained data" = "grey40")) +
+  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
+  theme(line = element_blank(),
+        axis.text = element_text(color = "black"),
+        axis.text.x = element_blank())
+```
+
+Statistics
+
+```{r}
+plot.df %>% 
+  filter(sex == "Female") %>% 
+  rstatix::kruskal_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Female") %>% 
+  rstatix::wilcox_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Male") %>% 
+  rstatix::kruskal_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Male") %>% 
+  rstatix::wilcox_test(s1 ~ visit)
+```
+
 # Figure 3
@@
 
-## Figure 3A
+## Figure 3a
 
@@
 
-## Figure 3B
+## Figure 3b
 
@@
 
-## Figure 3C
-
-Here, we're plotting proportions of total cells, so we load the Conos object and the annotation for all cells first.
+## Figure 3c
+
+Here, we're plotting proportions of total cells, so we use the Conos object and the annotation for all cells first.
+
 ```{r f3c}
-con.major <- qread("con_major.qs", nthreads = 10)
-anno.major <- qread("anno_major.qs")
-
 # Use Cacoa to calculate proportions
@@
 
-## Figure 3D
-
-Load data
-
-```{r, f3d}
-## Plot Individual genes from RNA-seq data
-Bulk_norm_counts <- read.delim("Bulk_norm_counts.txt", h=T)
-colData <- read.delim("Meta_Data_WL_Select.txt", h=T, dec = ",")
-```
-
-TREM2
-
-```{r, f3d2}
-gene_of_interest <- "TREM2"
-
-# Subset norm_counts to get counts for the gene of interest
-gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
-rownames(gene_counts)=gene_counts$hgnc_symbol
-gene_counts=gene_counts[4:96]
-
-# Reshape gene_counts into long format and merge with colData
-gene_counts_long <- gene_counts %>%
-  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
-  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")
-
-gene_counts_long <- gene_counts_long %>%
-  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData
-
-# Define a helper function to map p-values to stars
-pval_to_stars <- function(p) {
-  if (p < 0.0001) {
-    return("****")
-  } else if (p < 0.001) {
-    return("***")
-  } else if (p < 0.01) {
-    return("**")
-  } else if (p < 0.05) {
-    return("*")
-  } else {
-    return("ns")  # Not significant
-  }
-}
-
-# Perform a Kruskal-Wallis test separately for each gender
-kw_res_F <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "F"))
-kw_res_M <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "M"))
-
-# Extract Kruskal-Wallis p-values
-kw_pval_F <- kw_res_F$p.value
-kw_pval_M <- kw_res_M$p.value
-
-# Perform paired Wilcoxon tests for each gender with Holm's correction
-paired_wilcox <- function(data, sex_group) {
-  data_sex <- data %>% filter(sex == sex_group)
-  pairwise_tests <- list(
-    visit1_vs_visit2 = wilcox.test(
-      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
-      y = data_sex %>% filter(visit == "visit_2") %>% pull(count),
-      paired = TRUE
-    )$p.value,
-    visit1_vs_visit3 = wilcox.test(
-      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
-      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
-      paired = TRUE
-    )$p.value,
-    visit2_vs_visit3 = wilcox.test(
-      x = data_sex %>% filter(visit == "visit_2") %>% pull(count),
-      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
-      paired = TRUE
-    )$p.value
-  )
-  corrected_pvals <- p.adjust(unlist(pairwise_tests), method = "holm")
-  return(corrected_pvals)
-}
-
-wilcox_pvals_F <- paired_wilcox(gene_counts_long, "F")
-wilcox_pvals_M <- paired_wilcox(gene_counts_long, "M")
-
-# Map the Wilcoxon p-values to stars
-wilcox_stars_F <- sapply(wilcox_pvals_F, pval_to_stars)
-wilcox_stars_M <- sapply(wilcox_pvals_M, pval_to_stars)
-
-# Create the boxplot with significance annotations
-plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
-  geom_boxplot(outlier.shape = NA) +  # Remove outliers
-  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
-  labs(title = "Gene Expression Across Visits and Genders", x = "Visit and Sex", y = "Gene Count") +
-  theme_minimal() +
-  theme(plot.title = element_text(size = 10)) +
-  geom_signif(
-    comparisons = list(
-      c("visit_1.F", "visit_2.F"), 
-      c("visit_1.F", "visit_3.F"), 
-      c("visit_2.F", "visit_3.F"),
-      c("visit_1.M", "visit_2.M"), 
-      c("visit_1.M", "visit_3.M"), 
-      c("visit_2.M", "visit_3.M")
-    ),
-    annotations = c(
-      wilcox_stars_F["visit1_vs_visit2"],
-      wilcox_stars_F["visit1_vs_visit3"],
-      wilcox_stars_F["visit2_vs_visit3"],
-      wilcox_stars_M["visit1_vs_visit2"],
-      wilcox_stars_M["visit1_vs_visit3"],
-      wilcox_stars_M["visit2_vs_visit3"]
-    ),
-    textsize = 4,
-    map_signif_level = FALSE
-  )
-
-# Add Kruskal-Wallis p-values as text
-plot <- plot + 
-  annotate("text", x = 1, y = max(gene_counts_long$count) * 1.1, 
-           label = paste("KW Female p-value:", format(kw_pval_F, digits = 3)), 
-           hjust = 0, size = 3) +
-  annotate("text", x = 4, y = max(gene_counts_long$count) * 1.1, 
-           label = paste("KW Male p-value:", format(kw_pval_M, digits = 3)), 
-           hjust = 0, size = 3)
-
-# Display the plot
-print(plot)
-```
-
-SPP1
-
-```{r, f3d3}
-gene_of_interest <- "SPP1"
-
-# Subset norm_counts to get counts for the gene of interest
-gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
-rownames(gene_counts)=gene_counts$hgnc_symbol
-gene_counts=gene_counts[4:96]
-
-# Reshape gene_counts into long format and merge with colData
-gene_counts_long <- gene_counts %>%
-  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
-  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")
-
-gene_counts_long <- gene_counts_long %>%
-  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData
-
-# Define a helper function to map p-values to stars
-pval_to_stars <- function(p) {
-  if (p < 0.0001) {
-    return("****")
-  } else if (p < 0.001) {
-    return("***")
-  } else if (p < 0.01) {
-    return("**")
-  } else if (p < 0.05) {
-    return("*")
-  } else {
-    return("ns")  # Not significant
-  }
-}
-
-# Perform a Kruskal-Wallis test separately for each gender
-kw_res_F <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "F"))
-kw_res_M <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "M"))
-
-# Extract Kruskal-Wallis p-values
-kw_pval_F <- kw_res_F$p.value
-kw_pval_M <- kw_res_M$p.value
-
-# Perform paired Wilcoxon tests for each gender with Holm's correction
-paired_wilcox <- function(data, sex_group) {
-  data_sex <- data %>% filter(sex == sex_group)
-  pairwise_tests <- list(
-    visit1_vs_visit2 = wilcox.test(
-      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
-      y = data_sex %>% filter(visit == "visit_2") %>% pull(count),
-      paired = TRUE
-    )$p.value,
-    visit1_vs_visit3 = wilcox.test(
-      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
-      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
-      paired = TRUE
-    )$p.value,
-    visit2_vs_visit3 = wilcox.test(
-      x = data_sex %>% filter(visit == "visit_2") %>% pull(count),
-      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
-      paired = TRUE
-    )$p.value
-  )
-  corrected_pvals <- p.adjust(unlist(pairwise_tests), method = "holm")
-  return(corrected_pvals)
-}
-
-wilcox_pvals_F <- paired_wilcox(gene_counts_long, "F")
-wilcox_pvals_M <- paired_wilcox(gene_counts_long, "M")
-
-# Map the Wilcoxon p-values to stars
-wilcox_stars_F <- sapply(wilcox_pvals_F, pval_to_stars)
-wilcox_stars_M <- sapply(wilcox_pvals_M, pval_to_stars)
-
-# Create the boxplot with significance annotations
-plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
-  geom_boxplot(outlier.shape = NA) +  # Remove outliers
-  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
-  labs(title = "Gene Expression Across Visits and Genders", x = "Visit and Sex", y = "Gene Count") +
-  theme_minimal() +
-  theme(plot.title = element_text(size = 10)) +
-  geom_signif(
-    comparisons = list(
-      c("visit_1.F", "visit_2.F"), 
-      c("visit_1.F", "visit_3.F"), 
-      c("visit_2.F", "visit_3.F"),
-      c("visit_1.M", "visit_2.M"), 
-      c("visit_1.M", "visit_3.M"), 
-      c("visit_2.M", "visit_3.M")
-    ),
-    annotations = c(
-      wilcox_stars_F["visit1_vs_visit2"],
-      wilcox_stars_F["visit1_vs_visit3"],
-      wilcox_stars_F["visit2_vs_visit3"],
-      wilcox_stars_M["visit1_vs_visit2"],
-      wilcox_stars_M["visit1_vs_visit3"],
-      wilcox_stars_M["visit2_vs_visit3"]
-    ),
-    textsize = 4,
-    map_signif_level = FALSE
-  )
-
-# Add Kruskal-Wallis p-values as text
-plot <- plot + 
-  annotate("text", x = 1, y = max(gene_counts_long$count) * 1.1, 
-           label = paste("KW Female p-value:", format(kw_pval_F, digits = 3)), 
-           hjust = 0, size = 3) +
-  annotate("text", x = 4, y = max(gene_counts_long$count) * 1.1, 
-           label = paste("KW Male p-value:", format(kw_pval_M, digits = 3)), 
-           hjust = 0, size = 3)
-
-# Display the plot
-print(plot)
-```
-
-MMP9
+## Figures 3d,e
+
+Preparations
 
 ```{r}
-gene_of_interest <- "MMP9"
-
-# Subset norm_counts to get counts for the gene of interest
-gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
-rownames(gene_counts)=gene_counts$hgnc_symbol
-gene_counts=gene_counts[4:96]
-
-# Reshape gene_counts into long format and merge with colData
-gene_counts_long <- gene_counts %>%
-  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
-  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")
-
-gene_counts_long <- gene_counts_long %>%
-  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData
-
-# Define a helper function to map p-values to stars
-pval_to_stars <- function(p) {
-  if (p < 0.0001) {
-    return("****")
-  } else if (p < 0.001) {
-    return("***")
-  } else if (p < 0.01) {
-    return("**")
-  } else if (p < 0.05) {
-    return("*")
-  } else {
-    return("ns")  # Not significant
-  }
-}
-
-# Perform a Kruskal-Wallis test separately for each gender
-kw_res_F <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "F"))
-kw_res_M <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "M"))
-
-# Extract Kruskal-Wallis p-values
-kw_pval_F <- kw_res_F$p.value
-kw_pval_M <- kw_res_M$p.value
-
-# Perform paired Wilcoxon tests for each gender with Holm's correction
-paired_wilcox <- function(data, sex_group) {
-  data_sex <- data %>% filter(sex == sex_group)
-  pairwise_tests <- list(
-    visit1_vs_visit2 = wilcox.test(
-      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
-      y = data_sex %>% filter(visit == "visit_2") %>% pull(count),
-      paired = TRUE
-    )$p.value,
-    visit1_vs_visit3 = wilcox.test(
-      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
-      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
-      paired = TRUE
-    )$p.value,
-    visit2_vs_visit3 = wilcox.test(
-      x = data_sex %>% filter(visit == "visit_2") %>% pull(count),
-      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
-      paired = TRUE
-    )$p.value
-  )
-  corrected_pvals <- p.adjust(unlist(pairwise_tests), method = "holm")
-  return(corrected_pvals)
-}
-
-wilcox_pvals_F <- paired_wilcox(gene_counts_long, "F")
-wilcox_pvals_M <- paired_wilcox(gene_counts_long, "M")
-
-# Map the Wilcoxon p-values to stars
-wilcox_stars_F <- sapply(wilcox_pvals_F, pval_to_stars)
-wilcox_stars_M <- sapply(wilcox_pvals_M, pval_to_stars)
-
-# Create the boxplot with significance annotations
-plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
-  geom_boxplot(outlier.shape = NA) +  # Remove outliers
-  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
-  labs(title = "Gene Expression Across Visits and Genders", x = "Visit and Sex", y = "Gene Count") +
-  theme_minimal() +
-  theme(plot.title = element_text(size = 10)) +
-  geom_signif(
-    comparisons = list(
-      c("visit_1.F", "visit_2.F"), 
-      c("visit_1.F", "visit_3.F"), 
-      c("visit_2.F", "visit_3.F"),
-      c("visit_1.M", "visit_2.M"), 
-      c("visit_1.M", "visit_3.M"), 
-      c("visit_2.M", "visit_3.M")
-    ),
-    annotations = c(
-      wilcox_stars_F["visit1_vs_visit2"],
-      wilcox_stars_F["visit1_vs_visit3"],
-      wilcox_stars_F["visit2_vs_visit3"],
-      wilcox_stars_M["visit1_vs_visit2"],
-      wilcox_stars_M["visit1_vs_visit3"],
-      wilcox_stars_M["visit2_vs_visit3"]
-    ),
-    textsize = 4,
-    map_signif_level = FALSE
-  )
-
-# Add Kruskal-Wallis p-values as text
-plot <- plot + 
-  annotate("text", x = 1, y = max(gene_counts_long$count) * 1.1, 
-           label = paste("KW Female p-value:", format(kw_pval_F, digits = 3)), 
-           hjust = 0, size = 3) +
-  annotate("text", x = 4, y = max(gene_counts_long$count) * 1.1, 
-           label = paste("KW Male p-value:", format(kw_pval_M, digits = 3)), 
-           hjust = 0, size = 3)
-
-# Display the plot
-print(plot)
-```
-
-## Figure 3E
+# Use Cacoa to calculate proportions
+cao <- Cacoa$new(data.object = con.major, 
+                 cell.groups = anno.major, 
+                 ref.level = "vis2", 
+                 target.level = "vis3", 
+                 sample.groups = con.major$samples %>% 
+                   names() %>% 
+                   strsplit("_") %>% 
+                   sapply('[[', 3) %>% 
+                   gsub(1, 2, .) %>% 
+                   `names<-`(con.major$samples %>% 
+                               names()),
+                 sample.groups.palette = ggsci::pal_jama()(7))
+
+cao$sample.groups <- con.major$samples %>% 
+  names() %>% 
+  strsplit("_") %>% 
+  sapply('[[', 3) %>% 
+  `names<-`(con.major$samples %>% 
+              names()) %>% 
+  gsub("vis", "Visit ", .)
+
+anno.comb <- anno.major[!names(anno.major) %in% names(anno.immune)] %>% 
+  {factor(c(., anno.immune))}
+```
+
+Here, we're showing the preparations to create models for prediction. This takes some time and is not run again.
+
+```{r, eval = F}
+# Get markers
+markers.immune.all <- con.major$getDifferentialGenes(groups = anno.comb, z.threshold = 1, upregulated.only = T, verbose = T)
+
+# vEC
+# Get proportions
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(`Early LAM`, rownames(.))
+
+# Extract markers. Please note, the marker metrics may change a little in each iteration
+markers.all <- markers.immune.all$`Early LAM` %>% 
+  filter(Specificity > 0.9,
+         AUC > 0.65) %>% 
+  arrange(desc(AUC)) %>% 
+  pull(Gene)
+
+# Subset bulk data
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% markers.all) %>% 
+  as.matrix() %>% 
+  t()
+
+# Now, train a final model on all data for visualization and gene selection
+cm.test <- cm.bulk.mat %>% 
+  .[rownames(.) %in% names(proportions), ]
+
+lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 5)
+best_lambda <- lasso_cv$lambda.min
+lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)
+
+# Predict on all samples using the final model
+lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)
+
+# Save model
+list(lasso_cv = lasso_cv,
+     lasso_model = lasso_model) %>%
+  qsave("eLAM_model.qs")
+
+# cEC
+## Get proportions
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(LAM, rownames(.))
+
+## Markers
+markers.all <- markers.immune.all$LAM %>% 
+  filter(Specificity > 0.9,
+         AUC > 0.7) %>% 
+  arrange(desc(AUC)) %>% 
+  pull(Gene)
+
+## Subset bulk data
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% markers.all) %>% 
+  as.matrix() %>% 
+  t()
+
+# Now, train a final model on all data for visualization and gene selection
+cm.test <- cm.bulk.mat %>% 
+  .[rownames(.) %in% names(proportions), ]
+
+lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 15)
+best_lambda <- lasso_cv$lambda.min
+lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)
+
+# Predict on all samples using the final model
+lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)
+
+# Save model
+list(lasso_cv = lasso_cv,
+     lasso_model = lasso_model) %>%
+  qsave("LAM_model.qs")
+```
+
+### Figure 3d
+
+```{r}
+# Load and prepare
+tmp <- qread("eLAM_model.qs")
+
+lasso_model <- tmp$lasso_model
+best_lambda <- tmp$lasso_cv$lambda.min
+
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% rownames(coef(lasso_model))) %>% 
+  as.matrix() %>% 
+  t()
+
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(`Early LAM`, rownames(.))
+
+## Predict proportions
+proportions.bulk <-  predict(lasso_model, newx = cm.bulk.mat, s = best_lambda)
+
+## Prepare plot
+plot.df <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+trend.trained <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  filter(donor %in% names(proportions)) %>%
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  group_by(visit, sex) %>% 
+  summarize(s1 = median(s1)) %>%
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+trend.untrained <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  filter(!donor %in% names(proportions)) %>%
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  group_by(visit, sex) %>% 
+  summarize(s1 = median(s1)) %>% 
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+plot.df %>% 
+  ggplot() +
+  geom_boxplot(mapping = aes(sex_visit, s1, fill = sex_visit), outliers = F) +
+  geom_jitter(mapping = aes(sex_visit, s1, col = sex_visit), alpha = 0.3, col = "black", width = 0.3) +
+  theme_bw() +
+  labs(y = "Predicted proportions (%)", title = "Predicted eLAM proportions", x = "", col = "Median trends", fill = "") +
+  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Trained data"), data = trend.trained, linewidth = 1) +
+  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Untrained data"), data = trend.untrained, linewidth = 1) +
+  scale_color_manual(values = c("Trained data" = "black", "Untrained data" = "grey40")) +
+  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
+  theme(line = element_blank(),
+        axis.text = element_text(color = "black"),
+        axis.text.x = element_blank())
+```
+
+Statistics
+
+```{r}
+plot.df %>% 
+  filter(sex == "Female") %>% 
+  rstatix::kruskal_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Female") %>% 
+  rstatix::wilcox_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Male") %>% 
+  rstatix::kruskal_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Male") %>% 
+  rstatix::wilcox_test(s1 ~ visit)
+```
+
+### Figure 3e
+
+```{r}
+# Load and prepare
+tmp <- qread("LAM_model.qs")
+
+lasso_model <- tmp$lasso_model
+best_lambda <- tmp$lasso_cv$lambda.min
+
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% rownames(coef(lasso_model))) %>% 
+  as.matrix() %>% 
+  t()
+
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(LAM, rownames(.))
+
+## Predict proportions
+proportions.bulk <-  predict(lasso_model, newx = cm.bulk.mat, s = best_lambda)
+
+plot.df <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+trend.trained <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  filter(donor %in% names(proportions)) %>%
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  group_by(visit, sex) %>% 
+  summarize(s1 = median(s1)) %>%
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+trend.untrained <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  filter(!donor %in% names(proportions)) %>%
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  group_by(visit, sex) %>% 
+  summarize(s1 = median(s1)) %>% 
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+plot.df %>% 
+  ggplot() +
+  geom_boxplot(mapping = aes(sex_visit, s1, fill = sex_visit), outliers = F) +
+  geom_jitter(mapping = aes(sex_visit, s1, col = sex_visit), alpha = 0.3, col = "black", width = 0.3) +
+  theme_bw() +
+  labs(y = "Predicted proportions (%)", title = "Predicted LAM  proportions", x = "", col = "Median trends", fill = "") +
+  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Trained data"), data = trend.trained, linewidth = 1) +
+  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Untrained data"), data = trend.untrained, linewidth = 1) +
+  scale_color_manual(values = c("Trained data" = "black", "Untrained data" = "grey40")) +
+  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
+  theme(line = element_blank(),
+        axis.text = element_text(color = "black"),
+        axis.text.x = element_blank())
+```
+
+Statistics
+
+```{r}
+plot.df %>% 
+  filter(sex == "Female") %>% 
+  rstatix::kruskal_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Female") %>% 
+  rstatix::wilcox_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Male") %>% 
+  rstatix::kruskal_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Male") %>% 
+  rstatix::wilcox_test(s1 ~ visit)
+```
+
+## Figure 3f
 
@@
 
-genes <- de[c("ATM","Early LAM","LAM","mono/mac")] %>%
-  {`names<-`(lapply(names(.), \(nn) mutate(.[[nn]], ct = nn)), names(.))} %>% 
-  lapply(filter, padj <= 0.05, abs(log2FoldChange) > 1) %>%
-  lapply(pull, Gene) %>% 
-  {sapply(names(.), \(nn) .[[nn]] %>% `names<-`(., rep(nn, length(.))))} %>% 
-  Reduce(c, .) %>% 
-  .[match(unique(.), .)]
+# Initially, we used this approach to select genes. Since the clusters are not exactly reproducible, here we load the gene lists
+
+# genes <- de[c("ATM","Early LAM","LAM","mono/mac")] %>%
+#   {`names<-`(lapply(names(.), \(nn) mutate(.[[nn]], ct = nn)), names(.))} %>% 
+#   lapply(filter, padj <= 0.05, abs(log2FoldChange) > 1) %>%
+#   lapply(pull, Gene) %>% 
+#   {sapply(names(.), \(nn) .[[nn]] %>% `names<-`(., rep(nn, length(.))))} %>% 
+#   Reduce(c, .) %>% 
+#   .[match(unique(.), .)]
+
+genes <- read.delim("Macrophages_DEG_heatmap_genes.tsv") %$% 
+  setNames(gene, cluster) %>%
+  .[order(names(.))]
 
@@
                               cluster_columns=FALSE, 
-                              show_row_names=FALSE, 
+                              show_row_names=F, 
                               show_column_names=FALSE, 
@@
                               show_row_dend = FALSE,
-                              row_km = 5,
-                              row_km_repeats = 500,
+                              cluster_rows = F,
+                              # row_km = 5, # Only used initially
+                              # row_km_repeats = 500, # Only used initially
+                              row_split = names(genes),
                               column_split = groups[colnames(x)] %>% strsplit("!!") %>% sapply('[[', 2) %>% unname() %>% factor(levels = c("ATM", "mono/mac", "Early LAM", "LAM")))
@@
 
-## Figures 3F,G
+Save order for later
+
+```{r, eval = F}
+cls %>% 
+  lapply(\(y) rownames(x)[y]) %>% 
+  lapply(data.frame) %>% 
+  lapply(setNames, "gene") %>% 
+  bind_rows(.id = "cluster") %>%
+  select(gene, cluster) %>% 
+  write.table("Macrophages_DEG_heatmap_genes.tsv", sep = "\t")
+```
+
+## Figures 3g,h
 
@@
 
-### Figure 3F
+### Figure 3g
 
@@
 
-### Figure 3G
-
-```{r f3g}
-plot.dat <- cm.pseudo %>%
-  .[, colnames(.) %in% genes] %>% 
-  as.data.frame() %>% 
-  tibble::rownames_to_column(var = "sample") %>% 
-  mutate(visit = strsplit(sample, "!!|_") %>% 
-           sget(3) %>% 
-           factor(levels = c("vis1", "vis2", "vis3"), labels = c("visit 1", "visit 2", "visit 3")),
-         anno = strsplit(sample, "!!") %>% 
-           sget(2) %>% 
-           factor(levels = c("ATM", "mono/mac", "Early LAM", "LAM")),
-         donor = strsplit(sample, "!!") %>% 
-           sget(1)) %>% 
-  reshape2::melt(id.vars = c("sample", "visit", "anno", "donor")) %>%
-  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
-  mutate(sex_visit = paste(sex, visit, sep = " ")) %>% 
-  group_by(visit, sample, anno, donor, sex, sex_visit) %>%
+### Figure 3h
+
+```{r}
+anno <- anno.myeloid %>% 
+  .[. %in% cts] %>% 
+  factor()
+
+cm.clean <- cm.merged %>% 
+  .[rownames(.) %in% names(anno), ]
+
+spc <- con$getDatasetPerCell()
+
+lam.scores <- list(LAM.activity = genes)
+seu_obj <- CreateSeuratObject(counts = cm.clean %>% Matrix::t() %>% drop0(),
+                              min.cells = 0, 
+                              min.features = 0)
+m.scores <- UCell::AddModuleScore_UCell(seu_obj, features = lam.scores, ncores = 32)
+```
+
+```{r, fig.width = 6, fig.height = 4}
+plot.df <- reshape2::melt(m.scores$LAM.activity_UCell) %>% 
+  mutate(anno = anno[rownames(.)] %>% factor(levels = c("ATM", "mono/mac", "Early LAM", "LAM")), 
+         visit = spc[rownames(.)] %>% as.character() %>% strsplit("_") %>% sget(3),
+         donor = spc[rownames(.)] %>% as.character() %>% strsplit("_") %>% sapply(\(x) paste(x[1], x[2], sep = "_")),
+         donor_vis = spc[rownames(.)]) %>% 
+  mutate(sex = meta$sex[match(donor_vis, meta$sample)],
+         donor_anno = paste(donor, anno, sep = "-")) %>%
+  group_by(sex, visit, donor_anno, donor, anno) %>%
   summarize(value = mean(value)) %>% 
-  filter(visit != "visit 2") %>% 
-  mutate(visit = factor(visit),
-         donor_anno = paste(strsplit(donor, "_") %>% sapply(\(x) paste(x[1], x[2], sep = "_")), anno, sep = "_")) %>% 
-  arrange(visit, sample) %$% 
+  ungroup() %>% 
+  filter(visit != "vis2") %$% 
   split(., visit) %>% 
   {lapply(., filter, donor_anno %in% .[[2]]$donor_anno)} %>% 
-  {data.frame(.[[1]], diff = .[[2]]$value - .[[1]]$value)}
-
-plot.dat %>% 
-  ggplot(aes(anno, diff)) +
-  geom_boxplot(aes(fill = sex)) +
+  {rbind(data.frame(.[[2]], diff = .[[2]]$value - .[[1]]$value))}
+
+plot.df %>% 
+  ggplot(aes(anno, diff, fill = sex)) +
+  geom_boxplot() +
+  geom_point(aes(col = sex), alpha = 0.3, position = position_dodge(width = 0.8), col = "black") +
   geom_hline(yintercept = 0, color = "black") +
-  geom_point(aes(fill = sex), position = position_jitterdodge(jitter.width = 0.1),
-             color = "black", size = 0.5, alpha = 0.2) +
   theme_bw() +
-  labs(x = "", y = "Mean normalized pseudobulk difference\nin top gene marker expression", title = "LAM top markers") +
+  labs(x = "", y = "Mean module score difference (V3 - V1)", title = "LAM signature post-surgery weight loss", fill = "", col = "") +
   theme(line = element_blank(),
@@
         axis.text = element_text(color = "black")) +
-  scale_fill_manual(values = c("darkred", "blue3")) +
-  stat_compare_means(mapping = aes(fill = sex), method = "wilcox", label = "p.signif")
-```
-
-```{r f3g2}
-plot.dat %>% 
-  mutate(sex_visit_anno = paste(sex_visit, anno, sep = "_")) %$% 
-  split(diff, sex_visit_anno) %>% 
-  lapply(t.test, alternative = "less")
+  scale_fill_manual(values = c("darkred", "blue3"))
 ```
@@
 
-## Figure 4A
+## Figure 4a
 
@@
 
-## Figure 4B
+## Figure 4b
 
@@
 
-## Figure 4C
-
-Here, we're plotting proportions of total cells, so we load the Conos object and the annotation for all cells first.
+## Figure 4c
+
+Here, we're plotting proportions of total cells, so we use the Conos object and the annotation for all cells.
+
 ```{r f4c}
-con.major <- qread("con_major.qs", nthreads = 10)
-anno.major <- qread("anno_major.qs")
-
 # Use Cacoa to calculate proportions
@@
 
-## Figure 4D
+## Figure 4d
+
+Preparations
+
+```{r}
+# Use Cacoa to calculate proportions
+cao <- Cacoa$new(data.object = con.major, 
+                 cell.groups = anno.major, 
+                 ref.level = "vis2", 
+                 target.level = "vis3", 
+                 sample.groups = con.major$samples %>% 
+                   names() %>% 
+                   strsplit("_") %>% 
+                   sapply('[[', 3) %>% 
+                   gsub(1, 2, .) %>% 
+                   `names<-`(con.major$samples %>% 
+                               names()),
+                 sample.groups.palette = ggsci::pal_jama()(7))
+
+cao$sample.groups <- con.major$samples %>% 
+  names() %>% 
+  strsplit("_") %>% 
+  sapply('[[', 3) %>% 
+  `names<-`(con.major$samples %>% 
+              names()) %>% 
+  gsub("vis", "Visit ", .)
+
+anno.comb <- anno.major[!names(anno.major) %in% names(anno.aspc)] %>% 
+  {factor(c(., anno.aspc))}
+```
+
+Here, we're showing the preparations to create models for prediction. This takes some time and is not run again.
+
+```{r, eval = F}
+# Get markers
+markers.aspc.all <- con.major$getDifferentialGenes(groups = anno.comb, z.threshold = 1, upregulated.only = T, verbose = T)
+
+# Get proportions
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(`Early LAM`, rownames(.))
+
+# Extract markers. Please note, the marker metrics may change a little in each iteration
+markers.all <- markers.aspc.all$ASPC_DPP4 %>% 
+  filter(PAdj < 0.05,
+         Specificity > 0.7) %>% 
+  arrange(desc(AUC)) %>% 
+  pull(Gene)
+
+# Subset bulk data
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% markers.all) %>% 
+  as.matrix() %>% 
+  t()
+
+# Now, train a final model on all data for visualization and gene selection
+cm.test <- cm.bulk.mat %>% 
+  .[rownames(.) %in% names(proportions), ]
+
+lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 5)
+best_lambda <- lasso_cv$lambda.min
+lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)
+
+# Predict on all samples using the final model
+lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)
+
+# Save model
+list(lasso_cv = lasso_cv,
+     lasso_model = lasso_model) %>%
+  qsave("DPP4_model.qs")
+```
+
+```{r}
+# Load and prepare
+tmp <- qread("DPP4_model.qs")
+
+lasso_model <- tmp$lasso_model
+best_lambda <- tmp$lasso_cv$lambda.min
+
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% rownames(coef(lasso_model))) %>% 
+  as.matrix() %>% 
+  t()
+
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(ASPC_DPP4, rownames(.))
+
+## Predict proportions
+proportions.bulk <-  predict(lasso_model, newx = cm.bulk.mat, s = best_lambda)
+
+## Prepare plot
+plot.df <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+trend.trained <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  filter(donor %in% names(proportions)) %>%
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  group_by(visit, sex) %>% 
+  summarize(s1 = median(s1)) %>%
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+trend.untrained <- proportions.bulk %>% 
+  as.data.frame() %>%
+  mutate(donor = rownames(.)) %>% 
+  filter(!donor %in% names(proportions)) %>%
+  mutate(visit = strsplit(donor, "_") %>% sget(3),
+         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
+  group_by(visit, sex) %>% 
+  summarize(s1 = median(s1)) %>% 
+  mutate(sex_visit = paste(sex, visit, sep = "_"))
+
+plot.df %>% 
+  ggplot() +
+  geom_boxplot(mapping = aes(sex_visit, s1, fill = sex_visit), outliers = F) +
+  geom_jitter(mapping = aes(sex_visit, s1, col = sex_visit), alpha = 0.3, col = "black", width = 0.3) +
+  theme_bw() +
+  labs(y = "Predicted proportions (%)", title = "Predicted ASPC_DPP4 proportions", x = "", col = "Median trends", fill = "") +
+  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Trained data"), data = trend.trained, linewidth = 1) +
+  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Untrained data"), data = trend.untrained, linewidth = 1) +
+  scale_color_manual(values = c("Trained data" = "black", "Untrained data" = "grey40")) +
+  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
+  theme(line = element_blank(),
+        axis.text = element_text(color = "black"),
+        axis.text.x = element_blank())
+```
+
+Statistics
+
+```{r}
+plot.df %>% 
+  filter(sex == "Female") %>% 
+  rstatix::kruskal_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Female") %>% 
+  rstatix::wilcox_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Male") %>% 
+  rstatix::kruskal_test(s1 ~ visit)
+
+plot.df %>% 
+  filter(sex == "Male") %>% 
+  rstatix::wilcox_test(s1 ~ visit)
+```
+
+## Figure 4e
 
@@
 
-genes <- de %>%
-  {`names<-`(lapply(names(.), \(nn) mutate(.[[nn]], ct = nn)), names(.))} %>% 
-  lapply(filter, padj <= 0.05, abs(log2FoldChange) > 1) %>%
-  lapply(pull, Gene) %>% 
-  {sapply(names(.), \(nn) .[[nn]] %>% `names<-`(., rep(nn, length(.))))} %>% 
-  Reduce(c, .) %>% 
-  .[match(unique(.), .)]
+# See comment for Figure 3f
+
+# genes <- de %>%
+#   {`names<-`(lapply(names(.), \(nn) mutate(.[[nn]], ct = nn)), names(.))} %>% 
+#   lapply(filter, padj <= 0.05, abs(log2FoldChange) > 1) %>%
+#   lapply(pull, Gene) %>% 
+#   {sapply(names(.), \(nn) .[[nn]] %>% `names<-`(., rep(nn, length(.))))} %>% 
+#   Reduce(c, .) %>% 
+#   .[match(unique(.), .)]
+
+genes <- read.delim("ASPC_DEG_heatmap_genes.tsv") %$% 
+  setNames(gene, cluster) %>%
+  .[order(names(.))]
 
@@
                               show_row_dend = FALSE,
-                              row_km = 3,
-                              row_km_repeats = 100,
+                              # row_km = 3, # Only used initially
+                              # row_km_repeats = 100, # Only used initially
                               column_split = groups[colnames(x)] %>% strsplit("!!") %>% sapply('[[', 2) %>% unname() %>% factor(levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")),
-                              row_dend_reorder = TRUE)
+                              row_split = names(genes))
 
@@
 
-## Figure 4E
+Save order for later
+
+```{r, eval = F}
+cls %>% 
+  lapply(\(y) rownames(x)[y]) %>% 
+  lapply(data.frame) %>% 
+  lapply(setNames, "gene") %>% 
+  bind_rows(.id = "cluster") %>%
+  select(gene, cluster) %>% 
+  write.table("ASPC_DEG_heatmap_genes.tsv", sep = "\t")
+```
+
+## Figure 4f
 
@@
 
-## Figure 4G
+## Figure 4g
+
+See `Scenic.ipynb`.
+
+## Figure 4h
 
@@
 
-con.major <- qread("con_major.qs", nthreads = 10)
 spc <- con.major$getDatasetPerCell()
@@
 ```{r f4g2, fig.height = 4, fig.width = 5}
-anno.adipocytes <- qread("anno_major.qs") %>% 
-  .[. == "Adipocytes"] %>% 
-  factor()
-anno.comb <- factor(c(anno.adipocytes, anno.aspc))
+anno.comb <- factor(anno.aspc)
 
@@
   ungroup() %>% 
-  filter(variable == "JUN(+)",
-         !anno == "Adipocytes") %>% 
-  mutate(anno = factor(anno, levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3", "Adipocytes")),
+  filter(variable == "JUN(+)") %>% 
+  mutate(anno = factor(anno, levels = c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")),
          visit = factor(visit, levels = c("vis1", "vis2", "vis3"), labels = c("Visit 1", "Visit 2", "Visit 3")),
@@
   theme_bw() +
-  labs(x = "", y = "Mean AUC", fill = "") +
+  labs(x = "", y = "Regulon activity (mean AUC)", fill = "") +
   theme(line = element_blank(),
@@
 
-## Figure 5A
+## Figure 5a
 
@@
 
-## Figure 5B
+## Figure 5b
 
@@
 
-## Figure 5C
+## Figure 5c
 
@@
 
-genes <- de %>%
-  {`names<-`(lapply(names(.), \(nn) mutate(.[[nn]], ct = nn)), names(.))} %>% 
-  lapply(filter, padj <= 0.05, abs(log2FoldChange) > 1) %>%
-  lapply(pull, Gene) %>% 
-  {sapply(names(.), \(nn) .[[nn]] %>% `names<-`(., rep(nn, length(.))))} %>% 
-  Reduce(c, .) %>% 
-  .[match(unique(.), .)]
+# See comment for Figure 3f
+
+# genes <- de %>%
+#   {`names<-`(lapply(names(.), \(nn) mutate(.[[nn]], ct = nn)), names(.))} %>% 
+#   lapply(filter, padj <= 0.05, abs(log2FoldChange) > 1) %>%
+#   lapply(pull, Gene) %>% 
+#   {sapply(names(.), \(nn) .[[nn]] %>% `names<-`(., rep(nn, length(.))))} %>% 
+#   Reduce(c, .) %>% 
+#   .[match(unique(.), .)]
+
+genes <- read.delim("Archetypes_DEG_heatmap_genes.tsv") %$% 
+  setNames(gene, cluster) %>%
+  .[order(names(.))]
 
@@
                               show_row_dend = FALSE,
-                              row_km = 4,
-                              row_km_repeats = 200, # Here, this parameter greatly influences cluster composition
+                              # row_km = 4, # Only used initially
+                              # row_km_repeats = 200, # Only used initially
                               column_split = groups[colnames(x)] %>% strsplit("!!") %>% sapply('[[', 2) %>% unname() %>% factor(levels = c("CLSTN2_archetype", "DGAT2_archetype", "PRSS23_archetype")),
-                              row_dend_reorder = TRUE)
+                              row_split = names(genes))
 
@@
 
-## Figure 5D
+Save order for later
+
+```{r, eval = F}
+cls %>% 
+  lapply(\(y) rownames(x)[y]) %>% 
+  lapply(data.frame) %>% 
+  lapply(setNames, "gene") %>% 
+  bind_rows(.id = "cluster") %>%
+  select(gene, cluster) %>% 
+  write.table("Archetypes_DEG_heatmap_genes.tsv", sep = "\t")
+```
+
+## Figure 5d
 
@@
 
-## Figure 6A
+## Figure 6a
 
@@
 
-## Figure 6B
-
-We need the `sds_obj` object from Figure 6A.
+## Figure 6b
 
 ```{r f6b, fig.width=5, fig.height=2}
+# We need the `sds_obj` object from Figure 6a
 sds_obj <- qread("sds_obj.qs")
@@
 
-## Figure 6C
-
-We need the `sds_obj` object from Figure 6A.
+## Figure 6c
 
 ```{r f6c, fig.width=6, fig.height=2}
+# We need the `sds_obj` object from Figure 6a
 sds_obj <- qread("sds_obj.qs")
@@
 
-## Figure 6D
+## Figure 6d
 
@@
 
-We need the `sds_obj` object from Figure 6A.
-
 ```{r f6d2, fig.height = 4, fig.width = 5}
+# We need the `sds_obj` object from Figure 6a
 sds_obj <- qread("sds_obj.qs")
@@
 
-## Figure 6E
+## Figure 6e
 
@@
 
-## Figure 7D
+## Figures 7a-c
+
+See `Liana.ipynb`.
+
+## Figure 7d
 
@@
 ```{r f7d}
-dat.raw <- read.delim("/work/02_data/00_other/liana_res.csv",
+dat.raw <- read.delim("liana_res.csv",
                       sep = ",",
@@
 
-## Figure 7E
+## Figure 7e
 
@@
 
-# Figure S1
+# Extended Data Figure 1
 
@@
 ```{r s1load}
-con.major <- qread("con_major.qs", nthreads = 10)
-
 cms <- con.major$samples %>% 
@@
 
-## Figure S1A
+## Extended Data Figure 1a
 
@@
 
-## Figure S1B
+## Extended Data Figure 1b
+
+```{r}
+WATLAS <- read.delim("watlas_labels_l1_adipose_anno_super.tsv")
+rownames(WATLAS)=WATLAS$X
+WATLAS=WATLAS[2:21]
+
+# Define the order of columns and rows
+column_order <- c("Adipocyte", "FAP", "VEC", "LEC", "SMC","Pericyte", "B.cell", "CD4..T.cell", "CD8..T.cell", "ILC", "NK.cell", "Mast",  
+                  "Macrophage",  "Monocyte", "DC", "pDC",
+                  "Endometrium", "Mesothelial", "Plasmablast", 
+                  "Schwann" )
+
+row_order <- c("Adipocytes", "ASPC", "EC", "lEC", "SMCs", "Pericytes", "B-cells", "T-cells", "Mast cells", "Myeloid")
+WATLAS_ordered <- WATLAS[, rev(column_order)] # Reorder columns
+WATLAS_ordered <- WATLAS_ordered[row_order, ] # Reorder rows
+
+corrplot(WATLAS_ordered %>% t() %>% t(), method="color", col.lim=c(0, 1),
+         col = rev(COL2("RdBu",200)),
+         is.corr = F, 
+         tl.col	= "black",
+         tl.pos	= "lower",
+         outline = "#C0C0C0",
+         addgrid.col= NA)
+```
+
+## Extended Data Figure 1c
 
@@
 
-## Figure S1C
+## Extended Data Figure 1d
 
@@
 
-## Figure S1D
+## Extended Data Figure 1e
 
@@
 
-## Figure S1E
+## Extended Data Figure 1f
+
+Calculate
 
 ```{r}
-WATLAS <- read.delim("watlas_labels_l1_adipose_anno_super.tsv")
-rownames(WATLAS)=WATLAS$X
-WATLAS=WATLAS[2:21]
-
-# Define the order of columns and rows
-column_order <- c("Adipocyte", "FAP", "VEC", "LEC", "SMC","Pericyte", "B.cell", "CD4..T.cell", "CD8..T.cell", "ILC", "NK.cell", "Mast",  
-                  "Macrophage",  "Monocyte", "DC", "pDC",
-                  "Endometrium", "Mesothelial", "Plasmablast", 
-                  "Schwann" )
-
-row_order <- c("Adipocytes", "ASPC", "EC", "lEC", "SMCs", "Pericytes", "B-cells", "T-cells", "Mast cells", "Myeloid")
-WATLAS_ordered <- WATLAS[, rev(column_order)] # Reorder columns
-WATLAS_ordered <- WATLAS_ordered[row_order, ] # Reorder rows
-
-corrplot(t(WATLAS_ordered), method="color", order = , col.lim=c(0, 1),
-         col = rev(COL2("RdBu",200)),
-         is.corr = F, 
-         tl.col	= "black",
-         tl.pos	= "lower",
-         outline = "#C0C0C0",
-         addgrid.col= NA)
-```
-
-# Figure S2
+spc <- con.major$getDatasetPerCell()
+
+anno.sex <- data.frame(anno = unname(anno.major), 
+                       cid = names(anno.major)) %>% 
+  mutate(donor = spc[match(cid, names(spc))]) %>% 
+  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
+  mutate(anno_sex = paste(anno, tolower(sex), sep = ", ")) %>% 
+  pull(anno_sex, cid)
+
+res.major <- c("vis1", "vis2", "vis3") %>% 
+  sapply(\(vis) {
+    con <- con.major$samples %>% 
+      .[!grepl(vis, names(.))] %>%
+      Conos$new()
+
+    if (vis == "vis1") {
+      vis.ref <- "vis2"
+      vis.target <- "vis3" 
+    } else {
+      vis.ref <- "vis1"
+      if (vis == "vis2") {
+        vis.target <- "vis3"
+      } else {
+        vis.target <- "vis2"
+      }
+    }
+
+    sample.groups <- con$samples %>%
+      names() %>%
+      `names<-`(ifelse(grepl(vis.ref, .), vis.ref, vis.target), .)
+
+    cao <- Cacoa$new(con,
+                     sample.groups,
+                     anno.sex,
+                     ref.level = vis.ref,
+                     target.level = vis.target,
+                     n.cores = 32)
+
+    cao$estimateCellLoadings()
+
+    return(cao)
+  }, simplify = F, USE.NAMES = T)
+```
+
+Plot
+
+```{r}
+visits <- c("vis1", "vis2", "vis3")
+
+p <- visits %>% 
+  lapply(\(vis) {
+    comp <- visits[!visits == vis]
+
+    dat.p <- res.major[[vis]]$test.results$coda$padj
+
+    dat.plot <- res.major[[vis]]$test.results$coda$loadings %>% 
+      tibble::rownames_to_column(var = "anno") %>% 
+      reshape2::melt(id.var = "anno") %>% 
+      mutate(anno = renameAnnotation(anno, "Myeloid immune cells", "Myeloid ICs") %>% 
+               renameAnnotation("Lymphoid immune cells", "Lymphoid ICs"))
+
+    yvar <- dat.plot %>%
+      group_by(anno) %>%
+      summarize(var = mean(value)) %>%
+      mutate(padj = dat.p[match(anno, names(dat.p))]) %>%
+      arrange(padj, desc(abs(var)))
+
+    dat.plot %<>% 
+      mutate(anno = factor(anno, levels = rev(unique(yvar$anno))))
+
+    ymax <- dat.plot %>%
+      pull(value) %>%
+      abs() %>%
+      max()
+
+    out <- dat.plot %>% 
+      ggplot(aes(anno, value, fill = anno)) + 
+      geom_hline(yintercept = 0, col = "black") +
+      geom_jitter(alpha = 0.2, size = 0.3, col = "grey") +
+      geom_boxplot(outliers = F) +
+      coord_flip() +
+      ylim(c(-ymax, ymax)) +
+      theme_bw() + 
+      theme(line = element_blank(),
+            axis.text = element_text(color = "black")) +
+      guides(fill = "none") +
+      labs(title = paste0(comp[1], " vs ", comp[2]), x = "", y = "Loadings [AU]") +
+      scale_fill_manual(values = pal.major)
+
+    return(out)
+  })
+
+p
+```
+
+Statistics
+
+```{r}
+for (vis in visits) {
+  comp <- visits[!visits == vis]
+  message("Comparison: ", comp)
+  print(res.major[[vis]]$test.results$coda$padj)
+}
+```
+
+# Extended Data Figure 2
+
+We integrated our data with the data from [Hinte *et al.*](https://www.nature.com/articles/s41586-024-08165-7) that were graciously provided by the authors.
+As these data are not public, here we will show how we integrated the data and performed label transferring, but we will not run those steps here.
+
+```{r, eval = F}
+## Load data
+dat.nefa <- qread("scAT_NEFA.qs", nthreads = 10)
+con <- qread("con_major.qs", nthreads = 10)
+
+## Integrate
+cm <- GetAssayData(dat.nefa, assay = "RNA", layer = "counts")
+p2.list <- split(colnames(dat.nefa), dat.nefa$condition) %>% 
+  lapply(\(cid) cm[, cid]) %>% 
+  lapply(basicP2proc, n.cores = 32, get.largevis = F, get.tsne = F, make.geneknn = F)
+
+con$addSamples(p2.list)
+con$n.cores <- 64
+con$buildGraph()
+con$findCommunities(method = leiden.community, resolution = 1)
+con$embedGraph(method = "UMAP")
+
+nefa.ids <- cm %>% 
+  colnames() %>% 
+  `names<-`(rep("NEFA", length(.)), .)
+
+hinte.cls <- dat.nefa$cellType %>% 
+  setNames(colnames(dat.nefa))
+
+hinte.cond <- dat.nefa$condition %>% 
+  setNames(colnames(dat.nefa))
+
+# Label transfer, minor
+anno.minor <- c("adipocytes", "fap", "immune", "vascular") %>% 
+  sapply(\(x) paste0("anno_",x,".qs")) %>% 
+  lapply(qread) %>% 
+  setNames(rep("", 4)) %>% 
+  unlist() %>% 
+  scHelper::collapseAnnotation("Adipocytes") %>% 
+  scHelper::renameAnnotation(c("Th","Th1"), "CD4 T cells") %>% 
+  scHelper::renameAnnotation("Tem", "CD8 T cells") %>% 
+  scHelper::renameAnnotation("NKT", "NK-like T cells") %>% 
+  scHelper::renameAnnotation("cEC2", "cvEC") %>% 
+  .[unique(names(.))]
+
+anno.minor %<>% 
+  c(qread("anno_major.qs") %>% .[. == "Mast cells"]) %>% 
+  factor()
+
+# Major
+anno <- qread("anno_major") %>% 
+  scHelper::renameAnnotation("FAPs", "ASPCs") %>% 
+  scHelper::renameAnnotation("Tissue-resident macrophages", "Myeloid immune cells") %>% 
+  scHelper::renameAnnotation("Endothelial cells", "ECs") %>% 
+  scHelper::renameAnnotation("Lymphatic endothelial cells", "Lymphatic ECs") %>% 
+  .[!. == "SMCs"]
+
+anno.vasc <- qread("anno_vascular.qs") %>% 
+  .[. %in% c("SMCs","Pericytes")]
+
+anno <- factor(c(anno, anno.vasc)) %>% 
+  .[match(unique(names(.)), names(.))]
+
+anno.major <- anno[names(anno.minor)]
+
+# Fix annotation issues
+anno.minor <- anno.minor[!is.na(anno.minor)]
+anno.major <- anno.major[!is.na(anno.major)]
+matched <- names(anno.minor)[ (names(anno.minor) %in% names(anno.major))]
+anno.major <- anno.major[ names(anno.major) %in% matched]
+anno.minor <- anno.minor[ names(anno.minor) %in% matched]
+anno.major <- anno.major[ duplicated(names(anno.major)) == FALSE]
+anno.minor <- anno.minor[ duplicated(names(anno.minor)) == FALSE]
+
+qsave(anno.minor, "anno_minor.qs")
+
+## Label transfer
+loft.transfer <- con$propagateLabels(labels = anno.major)
+loft.transfer.minor <- con$propagateLabels(labels = anno.minor)
+
+# Loft annotations
+con$clusters$loft_transfer <- list(groups = loft.transfer$labels)
+con$clusters$loft_transfer_uncertainty <- list(groups = formatC(loft.transfer$uncertainty, 1) %>% {setNames(as.numeric(.), names(.))} %>% .[!grepl("!!", names(.))])
+con$clusters$loft_transfer_minor <- list(groups = loft.transfer.minor$labels)
+con$clusters$loft_transfer_minor_uncertainty <- list(groups = formatC(loft.transfer.minor$uncertainty, 1) %>% {setNames(as.numeric(.), names(.))} %>% .[!grepl("!!", names(.))])
+con$clusters$loft_clusters <- list(groups = anno.major)
+con$clusters$loft_clusters_minor <- list(groups = anno.minor)
+
+# Hinte annotations
+con$clusters$hinte_transfer <- list(groups = hinte.transfer$labels)
+con$clusters$hinte_clusters <- list(groups = hinte.cls)
+con$clusters$hinte_condition <- list(groups = hinte.cond)
+con$clusters$hinte_clusters_major <- list(groups = con$clusters$hinte_clusters$groups %>%
+                                            collapseAnnotation(c("APCs", "Endo", "Macro")) %>%
+                                            renameAnnotation("Endo", "EndoCs") %>% 
+                                            renameAnnotation("FAPs", "APCs"))
+
+qsave(con, "hinte_integration.qs", nthreads = 10)
+```
+
+## Load and prepare
+
+```{r}
+con.integration <- qread("hinte_integration.qs")
+loft.transfer <- con.integration$propagateLabels(anno.major)
+cm.merged <- con.integration$getJointCountMatrix()
+spc <- con.integration$getDatasetPerCell()
+```
+
+## Extended Data Figure 2a
+
+```{r}
+con.integration$plotGraph(groups = loft.transfer$labels,
+                          font.size = 5) +
+  scale_color_manual(values = pal.major) +
+  theme(line = element_blank())
+```
+
+## Extended Data Figure 2b
+
+```{r}
+con.integration$plotGraph(groups = loft.transfer$labels %>% .[!grepl("!!", names(.))], 
+                          plot.na = F,
+                          font.size = 5) +
+  scale_color_manual(values = pal.major) +
+  theme(line = element_blank())
+```
+
+## Extended Data Figure 2c
+
+```{r, fig.width=7, fig.height=4}
+# Calculate
+cm.pseudo <- cm.merged %>%
+  sccore::collapseCellsByType(spc) %>% 
+  t() %>%
+  apply(2, as, "integer")
+
+cm.pseudo %<>% 
+  DESeq2::DESeqDataSetFromMatrix(., 
+                                 colnames(.) %>% 
+                                   strsplit("_") %>% 
+                                   sget(1) %>% 
+                                   data.frame() %>% 
+                                   `dimnames<-`(list(colnames(cm.pseudo), "group")), 
+                                 design = ~ group) %>% 
+  DESeq2::estimateSizeFactors() %>% 
+  DESeq2::counts(normalized = T)
+
+# We rotate Hinte into Loft sample PCA space
+cm.loft <- cm.pseudo[, grepl("vis", colnames(cm.pseudo))] %>% 
+  as.data.frame() %>%
+  filter(rowSums(.) > 0)
+cm.hinte <- cm.pseudo[, !grepl("vis", colnames(cm.pseudo))] %>% 
+  as.data.frame() %>%
+  filter(rowSums(.) > 0)
+
+genesToKeep <- intersect(rownames(cm.loft), rownames(cm.hinte))
+
+pc.res <- 
+  cm.loft[genesToKeep, ] %>% 
+  t() %>% 
+  prcomp(center = T, 
+         scale = T)
+
+pc.tmp <- cm.hinte[genesToKeep, ] %>% 
+  as.data.frame() %>% 
+  filter(rowSums(.) > 0) %>% 
+  t() %>% 
+  scale(pc.res$center, pc.res$scale) %*% pc.res$rotation
+
+# Plot
+dat.plot <- rbind(pc.res$x, pc.tmp) %>%
+  data.frame() %>% 
+  mutate(id = rownames(.)) %>% 
+  mutate(type = grepl.replace(id, c("lean", "t0", "vis1", "vis2", "t1", "vis3"), c("lean", "obese", "obese", "short-term WL", "regressed", "regressed")) %>% factor(levels = c("lean", "obese", "short-term WL", "regressed")),
+         sex = meta$sex[match(id, meta$sample)] %>% 
+           as.character())
+
+dat.plot$sex[is.na(dat.plot$sex)] <- "Hinte"
+
+dat.plot %>% 
+  mutate(sex = factor(sex, levels = c("Female", "Male", "Hinte"), labels = c("Loft, female", "Loft, male", "Hinte"))) %>% 
+  ggplot(aes(PC1, PC2, col = type, shape = sex)) +
+  geom_point(
+    data = filter(dat.plot, sex == "Hinte"),
+    aes(PC1, PC2), color = "yellow3", shape = 15, size = 5, alpha = 0.7) +
+  geom_point(size = 3) +
+  theme_bw() +
+  scale_color_manual(values = c("purple", brewer.pal(9, "Greens")[c(3, 6, 9)])) +
+  theme(line = element_blank()) +
+  labs(shape = "Study and sex", col = "Sample type")
+```
+
+Variation
+
+```{r}
+pc.res$x %>% 
+  data.frame() %>% 
+  mutate(var = pc.res$sdev^2 / sum(pc.res$sdev^2)) %>% 
+  pull(var)
+```
+
+## Extended Data Figure 2d
+
+```{r}
+pc.res <- 
+  cm.hinte[genesToKeep, ] %>% 
+  t() %>% 
+  prcomp(center = T, 
+         scale = T)
+
+pc.tmp <- cm.loft[genesToKeep, ] %>% 
+  as.data.frame() %>% 
+  filter(rowSums(.) > 0) %>% 
+  t() %>% 
+  scale(pc.res$center, pc.res$scale) %*% pc.res$rotation
+
+# Plot
+dat.plot <- rbind(pc.res$x, pc.tmp) %>%
+  data.frame() %>% 
+  mutate(id = rownames(.)) %>% 
+  mutate(type = grepl.replace(id, c("lean", "t0", "vis1", "vis2", "t1", "vis3"), c("lean", "obese", "obese", "short-term WL", "regressed", "regressed")) %>% factor(levels = c("lean", "obese", "short-term WL", "regressed")),
+         sex = meta$sex[match(id, meta$sample)] %>% 
+           as.character())
+
+dat.plot$sex[is.na(dat.plot$sex)] <- "Hinte"
+
+dat.plot %>% 
+  mutate(sex = factor(sex, levels = c("Female", "Male", "Hinte"), labels = c("Loft, female", "Loft, male", "Hinte"))) %>% 
+  ggplot(aes(PC1, PC2, col = type, shape = sex)) +
+  geom_point(
+    data = filter(dat.plot, sex == "Hinte"),
+    aes(PC1, PC2), color = "yellow3", shape = 15, size = 5, alpha = 0.7) +
+  geom_point(size = 3) +
+  theme_bw() +
+  scale_color_manual(values = c("purple", brewer.pal(9, "Greens")[c(3, 6, 9)])) +
+  theme(line = element_blank()) +
+  labs(shape = "Study and sex", col = "Sample type")
+```
+
+Variation
+
+```{r}
+pc.res$x %>% 
+  data.frame() %>% 
+  mutate(var = pc.res$sdev^2 / sum(pc.res$sdev^2)) %>% 
+  pull(var)
+```
+
+## Extended Data Figure 2e
+
+```{r, fig.width = 6, fig.height = 4}
+tmp <- con.integration$clusters$loft_transfer$groups %>%
+  renameAnnotation("Endothelial cells", "ECs") %>%
+  renameAnnotation("FAPs", "ASPCs") %>% 
+  renameAnnotation("Lymphoid immune cells", "Lymphoid ICs") %>%
+  renameAnnotation(c("Tissue-resident macrophages", "Myeloid immune cells"), "Myeloid ICs") %>% 
+  {data.frame(cid = names(.), anno = unname(.))} %>% 
+  filter(!grepl("!!", cid)) %>% 
+  mutate(sample = con.integration$clusters$hinte_condition$groups[match(cid, names(con.integration$clusters$hinte_condition$groups))])
+
+tmp.sum <- tmp %>% 
+  group_by(sample) %>% 
+  summarize(total = n()) %>% 
+  split(., .$sample)
+
+tmp.sum.ct <- tmp %>% 
+  group_by(sample, anno) %>% 
+  summarize(total = n()) %>% 
+  split(., .$sample)
+
+tmp.prop <- Map(\(ct, tot, an) setNames(ct$total / tot$total, an), ct = tmp.sum.ct, tot = tmp.sum, an = lapply(tmp.sum.ct, pull, anno))
+
+df.plot <- tmp.prop %>% 
+  {Map(\(nn, x) data.frame(sample = nn, anno = names(x), value = unname(x)), nn = names(.), x = .)} %>% 
+  bind_rows() %>% 
+  mutate(anno = factor(anno) %>% 
+           factor(levels = levels(.)[c(1, 2, 5, 8, 3, 7, 6, 4)]) %>% 
+           renameAnnotation("Lymphatic endothelial cells", "LECs"))
+
+df.plot %>% 
+  rbind(c("st_t0", "LECs", 0)) %>% # Adding an artificial 0 as LECs aren't detected in all conditions
+  mutate(sample = strsplit(sample, "_") %>% sapply(\(x) if (length(x) == 1) x[1] else x[2]),
+         anno = factor(anno, levels = c("Adipocytes", "ASPCs", "ECs", "Myeloid ICs", "LECs", "SMCs", "Pericytes", "Lymphoid ICs", "Mast cells")),
+         sample = factor(sample, levels = c("lean", "t0", "t1"), labels = c("Lean", "Obese", "Regressed")),
+         value = as.numeric(value)) %>% 
+  ggplot(aes(anno, value)) +
+  stat_summary(aes(fill = sample), fun = "mean", geom = "bar", position = position_dodge()) +
+  geom_point(aes(col = sample), position = position_dodge(width = 0.9)) +
+  theme_bw() + 
+  labs(x = "", y = "% proportion per sample", title = "Loft to Hinte transfer, major annotation", fill = "", col = "") +
+  theme(line = element_blank(),
+        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), 
+        axis.text = element_text(colour = "black")) +
+  scale_fill_manual(values = c(brewer.pal(9, "Greens")[c(3,6,9)])) +
+  scale_color_manual(values = c(brewer.pal(9, "Purples")[c(9,7,5)]))
+```
+
+## Extended Data Figure 2g
+
+```{r}
+size_data <- read.csv("Adipocyte_size.csv",sep = ",", header = TRUE)
+condition_levels <- c("F1", "F2", "F3", "M1", "M2", "M3")
+
+ggplot(size_data, aes(x = Diameter, y = factor(Condition, levels = rev(condition_levels)), fill = Condition)) +
+  ggridges::geom_density_ridges(scale = 0.9) +
+  xlab(expression(paste("Diameter (µm)"))) + 
+  ylab("Condition") + 
+  ggtitle("Adipocyte diameter distribution") + 
+  theme(legend.position="none", plot.title = element_text(hjust = 0.5, size = 14, face = "bold"), text = element_text(family = "serif", size = 12)) +
+  scale_fill_manual(values = c("#F8B3BB", "#BD6563", "#7D2A25", "#ABD1D4", "#62649E", "#3E4282")) +
+  theme(rect = element_rect(fill = "transparent"))
+```
+
+# Extended Data Figure 3
 
@@
 
-## Figure S2A
+## Extended Data Figure 3a
 
@@
 
-## Figure S2B
+## Extended Data Figure 3b
 
@@
 
-## Figure S2C
+## Extended Data Figure 3c
 
@@
 
-## Figure S2D
+## Extended Data Figure 3d
 
@@
 
-corrplot(t(WATLAS_ordered), method="color", order = , col.lim=c(0, 1),
+corrplot(WATLAS_ordered %>% t() %>% t(), method="color", col.lim=c(0, 1),
          col = rev(COL2("RdBu",200)),
@@
 
-# Figure S3
-
-## Figure S3A
+## Extended Data Figure 3e
+
+Calculate
+
+```{r}
+anno.minor <- qread("anno_minor.qs")
+
+spc <- con.major$getDatasetPerCell() %>% 
+  .[match(names(anno.minor), names(.))]
+
+anno.minor.sex <- anno.minor %>% 
+  data.frame(anno = ., cid = names(.), donor = unname(spc)) %>% 
+  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
+  mutate(anno_sex = paste(anno, sex, sep = "_")) %$% 
+  setNames(anno_sex, cid) %>% 
+  factor()
+
+# Calculate
+res.minor <- c("vis1", "vis2", "vis3") %>% 
+  sapply(\(vis) {
+    con <- con.major$samples %>% 
+      .[!grepl(vis, names(.))] %>%
+      Conos$new()
+
+    if (vis == "vis1") {
+      vis.ref <- "vis2"
+      vis.target <- "vis3" 
+    } else {
+      vis.ref <- "vis1"
+      if (vis == "vis2") {
+        vis.target <- "vis3"
+      } else {
+        vis.target <- "vis2"
+      }
+    }
+
+    sample.groups <- con$samples %>%
+      names() %>%
+      `names<-`(ifelse(grepl(vis.ref, .), vis.ref, vis.target), .)
+
+    cao <- Cacoa$new(con,
+                     sample.groups,
+                     anno.minor.sex,
+                     ref.level = vis.ref,
+                     target.level = vis.target,
+                     n.cores = 32)
+
+    cao$estimateCellLoadings()
+
+    return(cao)
+  }, simplify = F, USE.NAMES = T)
+```
+
+Plot
+
+```{r}
+vasc.pal <- RColorBrewer::brewer.pal(9, "Purples")[-c(1,2)][c(1,4,7,2,6,3,5)] %>% 
+  setNames(levels(qread("anno_vascular.qs"))) %>% 
+  {setNames(c(rep(., 3)), c(names(.), paste(names(.), "Female", sep = "_"), paste(names(.), "Male", sep = "_")))}
+
+visits <- c("vis1", "vis2", "vis3")
+
+p <- visits %>% 
+  lapply(\(vis) {
+    comp <- visits[!visits == vis]
+
+    dat.p <- res.minor[[vis]]$test.results$coda$padj
+
+    dat.plot <- res.minor[[vis]]$test.results$coda$loadings %>% 
+      tibble::rownames_to_column(var = "anno") %>% 
+      reshape2::melt(id.var = "anno") %>% 
+      filter(anno %in% names(vasc.pal))
+
+    yvar <- dat.plot %>%
+      group_by(anno) %>%
+      summarize(var = mean(value)) %>%
+      mutate(padj = dat.p[match(anno, names(dat.p))]) %>%
+      arrange(padj, desc(abs(var)))
+
+    dat.plot %<>% 
+      mutate(anno = factor(anno, levels = rev(unique(yvar$anno))))
+
+    ymax <- dat.plot %>%
+      pull(value) %>%
+      abs() %>%
+      max()
+
+    out <- dat.plot %>% 
+      ggplot(aes(anno, value, fill = anno)) + 
+      geom_hline(yintercept = 0, col = "black") +
+      geom_jitter(alpha = 0.2, size = 0.3, col = "grey") +
+      geom_boxplot(outliers = F) +
+      coord_flip() +
+      ylim(c(-ymax, ymax)) +
+      theme_bw() + 
+      theme(line = element_blank(),
+            axis.text = element_text(color = "black")) +
+      guides(fill = "none") +
+      labs(title = paste0(comp[1], " vs ", comp[2]), x = "", y = "Loadings [AU]") +
+      scale_fill_manual(values = vasc.pal)
+
+    return(out)
+  })
+
+p
+```
+
+Statistics
+
+```{r}
+for (vis in visits) {
+  comp <- visits[!visits == vis]
+  message("Comparison: ", comp)
+  print(res.minor[[vis]]$test.results$coda$padj %>% .[names(.) %in% names(vasc.pal)])
+}
+```
+
+## Extended Data Figure 3f
+
+```{r}
+watlas_data <- read.delim("watlas_mapped_percent.tsv", stringsAsFactors = FALSE)
+watlas_data_count <- read.delim("watlas_mapped_counts.tsv", stringsAsFactors = FALSE)
+watlas_data_count <- watlas_data_count %>%
+  rowwise() %>%
+  mutate(row_sum = sum(c_across(where(is.numeric)), na.rm = TRUE))  # Sum across numeric columns
+valid_samples <- watlas_data_count %>%
+  filter(row_sum >= 1000) %>%
+  pull(sample)  # Extracts the valid sample IDs
+
+filtered_data <- watlas_data %>%
+  filter(type == "sn", tissue == "AT", depot == "SAT", grepl("Hs", sample))
+filtered_data <- filtered_data %>%
+  filter(sample %in% valid_samples)
+
+long_data_VEC <- filtered_data %>%
+  select(wtstatus, arEC, cEC, vEC,  lEC, SMCs, Pericytes) %>%  # Select relevant columns
+  tidyr::pivot_longer(cols = -wtstatus, names_to = "Cell_Type", values_to = "Percentage")
+```
+
+```{r, fig.width=6, fig.height=4}
+long_data_VEC %>% 
+  mutate(Cell_Type = factor(Cell_Type, levels = c("arEC", "cEC", "vEC", "lEC", "Pericytes", "SMCs"))) %>% 
+  ggplot(aes(x = Cell_Type, y = Percentage, fill = wtstatus)) +
+  geom_boxplot(position = position_dodge(width = 0.8)) +
+  geom_point(aes(group = wtstatus), position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.8), alpha = 0.3, ) +
+  theme_bw() +
+  labs(title = "Comparison of Cell Types in Lean vs Obese",
+       x = "Cell Type",
+       y = "Percentage",
+       fill = "") +
+  scale_fill_manual(values = c(lean = "#26547c", obese = "#ef476f")) +
+  stat_compare_means(aes(group = wtstatus), method = "wilcox.test", label = "p.signif", ) +
+  theme(axis.text = element_text(color = "black"),
+        line = element_blank())
+```
+
+## Extended Data Figure 3g
+
+```{r, fig.width=6, fig.height=4}
+tmp <- con.integration$clusters$loft_transfer_minor$groups %>%
+  {data.frame(cid = names(.), anno = unname(.))} %>% 
+  filter(!grepl("!!", cid)) %>% 
+  mutate(sample = con.integration$clusters$hinte_condition$groups[match(cid, names(con.integration$clusters$hinte_condition$groups))])
+
+tmp.sum <- tmp %>% 
+  group_by(sample) %>% 
+  summarize(total = n()) %>% 
+  split(., .$sample)
+
+tmp.sum.ct <- tmp %>% 
+  group_by(sample, anno) %>% 
+  summarize(total = n()) %>% 
+  split(., .$sample)
+
+tmp.prop <- Map(\(ct, tot, an) setNames(ct$total / tot$total, an), ct = tmp.sum.ct, tot = tmp.sum, an = lapply(tmp.sum.ct, pull, anno))
+
+df.plot <- tmp.prop %>%
+  {Map(\(nn, x) data.frame(sample = nn, anno = names(x), value = unname(x)), nn = names(.), x = .)} %>% 
+  bind_rows() %>% 
+  mutate(anno = factor(anno, levels = unique(anno)),
+         type = sapply(sample, \(x) if (grepl("lean", x)) "Lean" else if (grepl("t0", x)) "Obese" else "Regressed"))
+
+cts <- c("arEC", "cEC", "vEC", "cvEC", "lEC", "Pericytes", "SMCs")
+
+p <- df.plot %>% 
+  filter(anno %in% cts) %>% 
+  mutate(anno = factor(anno, levels = cts)) %>% 
+  rbind(data.frame(sample = c("rg_t0", "rg_t1"), anno = c("lEC", "lEC"), value = c(0, 0), type = c("Obese", "Regressed"))) %>%
+  ggplot(aes(anno, value)) + 
+  stat_summary(mapping = aes(fill = type), geom = "bar", fun = "mean", position = "dodge") +
+  geom_point(aes(col = type), position = position_dodge(width = 0.9)) +
+  theme_bw() +
+  labs(x = "", y = "% proportion per sample", title = "Loft to Hinte transfer, minor annotation", col = "", fill = "") +
+  theme(axis.text = element_text(colour = "black"),
+        line = element_blank(),
+        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
+  scale_fill_manual(values = brewer.pal(9, "Greens")[c(3, 6, 9)]) +
+  scale_color_manual(values = brewer.pal(9, "Purples")[c(9, 7, 5)])
+
+p
+```
+
+## Extended Data Figures 3h,i
+
+```{r}
+# Metadata
+metadata.all <- read.delim("Meta_Data_WL_Select.txt", h=T, dec = ",") %>% 
+  mutate(sex = factor(sex, labels = c("Female", "Male")),
+         donor = gsub("D", "Donor_", recordid) %>% 
+           gsub("V", "vis", .))
+
+# Bulk data
+cm.bulk <- read.delim("Bulk_norm_counts.txt", header = T) %>% 
+  .[match(unique(.$hgnc_symbol), .$hgnc_symbol), ] %>% 
+  `rownames<-`(.$hgnc_symbol) %>% 
+  select(-hgnc_symbol, -ensembl_gene_id, -entrezgene_id) %>% 
+  scale()
+
+colnames(cm.bulk) %<>% 
+  gsub("D", "Donor_", .) %>% 
+  gsub("V", "vis", .)
+```
+
+### Extended Data Figure 3h
+
+See Figure 2d for model setup
+
+```{r}
+# Use Cacoa to calculate proportions
+cao <- Cacoa$new(data.object = con.major, 
+                 cell.groups = anno.major, 
+                 ref.level = "vis2", 
+                 target.level = "vis3", 
+                 sample.groups = con.major$samples %>% 
+                   names() %>% 
+                   strsplit("_") %>% 
+                   sapply('[[', 3) %>% 
+                   gsub(1, 2, .) %>% 
+                   `names<-`(con.major$samples %>% 
+                               names()),
+                 sample.groups.palette = ggsci::pal_jama()(7))
+
+cao$sample.groups <- con.major$samples %>% 
+  names() %>% 
+  strsplit("_") %>% 
+  sapply('[[', 3) %>% 
+  `names<-`(con.major$samples %>% 
+              names()) %>% 
+  gsub("vis", "Visit ", .)
+
+anno.comb <- anno.major[!names(anno.major) %in% names(anno.vascular)] %>% 
+  {factor(c(., anno.vascular))}
+
+# Get proportions
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(cEC, rownames(.))
+
+tmp <- qread("cEC_model.qs")
+
+lasso_model <- tmp$lasso_model
+best_lambda <- tmp$lasso_cv$lambda.min
+
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% rownames(coef(lasso_model, s = best_lambda))) %>% 
+  as.matrix() %>% 
+  t()
+
+# Now, train a final model on all data for visualization and gene selection
+cm.test <- cm.bulk.mat %>% 
+  .[rownames(.) %in% names(proportions), ]
+
+lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 40)
+best_lambda <- lasso_cv$lambda.min
+lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)
+
+# Predict on all samples using the final model
+lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)
+r2.all <- cor(proportions, lasso.predict.all)^2
+
+data.frame(Proportions = proportions,
+           Predictions = unname(lasso.predict.all)) %>% 
+  ggplot(aes(Proportions, Predictions)) + 
+  geom_point() +
+  theme_bw() +
+  theme(axis.text = element_text(colour = "black")) +
+  labs(title = paste0("r2 = ", formatC(r2.all, digits = 3)), y = "Predicted proportions (%)", x = "Proportions (%)") +
+  geom_abline(intercept = 0, slope = 1, colour = "firebrick")
+```
+
+### Extended Data Figure 3i
+
+See Figure 2e for model setup
+
+```{r}
+# Use Cacoa to calculate proportions
+cao <- Cacoa$new(data.object = con.major, 
+                 cell.groups = anno.major, 
+                 ref.level = "vis2", 
+                 target.level = "vis3", 
+                 sample.groups = con.major$samples %>% 
+                   names() %>% 
+                   strsplit("_") %>% 
+                   sapply('[[', 3) %>% 
+                   gsub(1, 2, .) %>% 
+                   `names<-`(con.major$samples %>% 
+                               names()),
+                 sample.groups.palette = ggsci::pal_jama()(7))
+
+cao$sample.groups <- con.major$samples %>% 
+  names() %>% 
+  strsplit("_") %>% 
+  sapply('[[', 3) %>% 
+  `names<-`(con.major$samples %>% 
+              names()) %>% 
+  gsub("vis", "Visit ", .)
+
+anno.comb <- anno.major[!names(anno.major) %in% names(anno.vascular)] %>% 
+  {factor(c(., anno.vascular))}
+
+# Get proportions
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(vEC, rownames(.))
+
+tmp <- qread("vEC_model.qs")
+
+lasso_model <- tmp$lasso_model
+best_lambda <- tmp$lasso_cv$lambda.min
+
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% rownames(coef(lasso_model, s = best_lambda))) %>% 
+  as.matrix() %>% 
+  t()
+
+# Now, train a final model on all data for visualization and gene selection
+cm.test <- cm.bulk.mat %>% 
+  .[rownames(.) %in% names(proportions), ]
+
+lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 40)
+best_lambda <- lasso_cv$lambda.min
+lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)
+
+# Predict on all samples using the final model
+lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)
+r2.all <- cor(proportions, lasso.predict.all)^2
+
+data.frame(Proportions = proportions,
+           Predictions = unname(lasso.predict.all)) %>% 
+  ggplot(aes(Proportions, Predictions)) + 
+  geom_point() +
+  theme_bw() +
+  theme(axis.text = element_text(colour = "black")) +
+  labs(title = paste0("r2 = ", formatC(r2.all, digits = 3)), y = "Predicted proportions (%)", x = "Proportions (%)") +
+  geom_abline(intercept = 0, slope = 1, colour = "firebrick")
+```
+
+# Extended Data Figure 4
+
+## Extended Data Figure 4a
 
@@
 
-## Figure S3B-G
+## Extended Data Figure 4b-g
 
@@
 
-### Figure S3B
+### Extended Data Figure 4b
 
@@
 
-### Figure S3C
+### Extended Data Figure 4c
 
 ```{r s3c, fig.width=7, fig.height=4}
-pdf("/work/02_data/00_prints/S3A.pdf", width = 7, height = 4)
 varToPlot <- spc %>% 
@@
   dotSize(3)
-dev.off()
-```
-
-### Figure S3D
+```
+
+### Extended Data Figure 4d
 
@@
 
-### Figure S3E
+### Extended Data Figure 4e
 
@@
 
-### Figure S3F
+### Extended Data Figure 4f
 
@@
 
-### Figure S3G
+### Extended Data Figure 4g
 
 ```{r s3g}
-con.major <- qread("con_major.qs", nthreads = 10)
-anno.major <- qread("anno_major.qs")
-
 # Use Cacoa to calculate proportions
@@
 
-## Figure S3H-J,L
+# Extended Data Figure 5
+
+## Extended Data Figure 5a-c
 
@@
 
-### Figure S3H
+### Extended Data Figure 5a
 
@@
 
-### Figure S3I
+### Extended Data Figure 5b
 
@@
 
-### Figure S3J
+### Extended Data Figure 5c
 
@@
 
-## Figure S3K
+## Extended Data Figure 5d
 
@@
 
-## Figure S3L
+## Extended Data Figure 5e
+
+Calculate
+
+```{r}
+anno.minor <- qread("anno_minor.qs")
+
+spc <- con.major$getDatasetPerCell() %>% 
+  .[match(names(anno.minor), names(.))]
+
+anno.minor.sex <- anno.minor %>% 
+  data.frame(anno = ., cid = names(.), donor = unname(spc)) %>% 
+  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
+  mutate(anno_sex = paste(anno, sex, sep = "_")) %$% 
+  setNames(anno_sex, cid) %>% 
+  factor()
+
+# Calculate
+res.minor <- c("vis1", "vis2", "vis3") %>% 
+  sapply(\(vis) {
+    con <- con.major$samples %>% 
+      .[!grepl(vis, names(.))] %>%
+      Conos$new()
+
+    if (vis == "vis1") {
+      vis.ref <- "vis2"
+      vis.target <- "vis3" 
+    } else {
+      vis.ref <- "vis1"
+      if (vis == "vis2") {
+        vis.target <- "vis3"
+      } else {
+        vis.target <- "vis2"
+      }
+    }
+
+    sample.groups <- con$samples %>%
+      names() %>%
+      `names<-`(ifelse(grepl(vis.ref, .), vis.ref, vis.target), .)
+
+    cao <- Cacoa$new(con,
+                     sample.groups,
+                     anno.minor.sex,
+                     ref.level = vis.ref,
+                     target.level = vis.target,
+                     n.cores = 32)
+
+    cao$estimateCellLoadings()
+
+    return(cao)
+  }, simplify = F, USE.NAMES = T)
+```
+
+Plot
+
+```{r}
+myeloid.pal <- c("#174D86", "#9EC3B5", "#356790", "#ADD0BA", "#628EA0", "#80A9AA", "#719CA5") %>% 
+  setNames(levels(qread("/work/02_data/09_export/anno_immune.qs") %>%  
+                    .[!. %in% c("NKT","NK","Tem","Th1","Th","Treg","B-cells")] %>% 
+                    factor(levels = c("ATM", "mono/mac", "Early LAM", "LAM", "cDC1", "cDC2", "mDC")))) %>%
+  {setNames(c(rep(., 3)), c(names(.), paste(names(.), "Female", sep = "_"), paste(names(.), "Male", sep = "_")))} %>% 
+  .[grepl("ATM|LAM|mono", names(.))]
+
+visits <- c("vis1", "vis2", "vis3")
+
+p <- visits %>% 
+  lapply(\(vis) {
+    comp <- visits[!visits == vis]
+
+    dat.p <- res.minor[[vis]]$test.results$coda$padj
+
+    dat.plot <- res.minor[[vis]]$test.results$coda$loadings %>% 
+      tibble::rownames_to_column(var = "anno") %>% 
+      reshape2::melt(id.var = "anno") %>% 
+      filter(anno %in% names(myeloid.pal))
+
+    yvar <- dat.plot %>%
+      group_by(anno) %>%
+      summarize(var = mean(value)) %>%
+      mutate(padj = dat.p[match(anno, names(dat.p))]) %>%
+      arrange(padj, desc(abs(var)))
+
+    dat.plot %<>% 
+      mutate(anno = factor(anno, levels = rev(unique(yvar$anno))))
+
+    ymax <- dat.plot %>%
+      pull(value) %>%
+      abs() %>%
+      max()
+
+    out <- dat.plot %>% 
+      ggplot(aes(anno, value, fill = anno)) + 
+      geom_hline(yintercept = 0, col = "black") +
+      geom_jitter(alpha = 0.2, size = 0.3, col = "grey") +
+      geom_boxplot(outliers = F) +
+      coord_flip() +
+      ylim(c(-ymax, ymax)) +
+      theme_bw() + 
+      theme(line = element_blank(),
+            axis.text = element_text(color = "black")) +
+      guides(fill = "none") +
+      labs(title = paste0(comp[1], " vs ", comp[2]), x = "", y = "Loadings [AU]") +
+      scale_fill_manual(values = myeloid.pal)
+
+    return(out)
+  })
+
+p
+```
+
+Statistics
+
+```{r}
+for (vis in visits) {
+  comp <- visits[!visits == vis]
+  message("Comparison: ", comp)
+  print(res.minor[[vis]]$test.results$coda$padj %>% .[names(.) %in% names(myeloid.pal)])
+}
+```
+
+## Extended Data Figure 5f
+
+Load data
+
+```{r, f3d}
+## Plot Individual genes from RNA-seq data
+Bulk_norm_counts <- read.delim("Bulk_norm_counts.txt", h=T)
+colData <- read.delim("Meta_Data_WL_Select.txt", h=T, dec = ",")
+```
+
+TREM2
+
+```{r, f3d2}
+gene_of_interest <- "TREM2"
+
+# Subset norm_counts to get counts for the gene of interest
+gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
+rownames(gene_counts)=gene_counts$hgnc_symbol
+gene_counts=gene_counts[4:96]
+
+# Reshape gene_counts into long format and merge with colData
+gene_counts_long <- gene_counts %>%
+  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
+  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")
+
+gene_counts_long <- gene_counts_long %>%
+  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData
+
+# Define a helper function to map p-values to stars
+pval_to_stars <- function(p) {
+  if (p < 0.0001) {
+    return("****")
+  } else if (p < 0.001) {
+    return("***")
+  } else if (p < 0.01) {
+    return("**")
+  } else if (p < 0.05) {
+    return("*")
+  } else {
+    return("ns")  # Not significant
+  }
+}
+
+# Perform a Kruskal-Wallis test separately for each gender
+kw_res_F <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "F"))
+kw_res_M <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "M"))
+
+# Extract Kruskal-Wallis p-values
+kw_pval_F <- kw_res_F$p.value
+kw_pval_M <- kw_res_M$p.value
+
+# Perform paired Wilcoxon tests for each gender with Holm's correction
+paired_wilcox <- function(data, sex_group) {
+  data_sex <- data %>% filter(sex == sex_group)
+  pairwise_tests <- list(
+    visit1_vs_visit2 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      paired = TRUE
+    )$p.value,
+    visit1_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
+      paired = TRUE
+    )$p.value,
+    visit2_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
+      paired = TRUE
+    )$p.value
+  )
+  corrected_pvals <- p.adjust(unlist(pairwise_tests), method = "holm")
+  return(corrected_pvals)
+}
+
+wilcox_pvals_F <- paired_wilcox(gene_counts_long, "F")
+wilcox_pvals_M <- paired_wilcox(gene_counts_long, "M")
+
+# Map the Wilcoxon p-values to stars
+wilcox_stars_F <- sapply(wilcox_pvals_F, pval_to_stars)
+wilcox_stars_M <- sapply(wilcox_pvals_M, pval_to_stars)
+
+# Create the boxplot with significance annotations
+plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
+  geom_boxplot(outlier.shape = NA) +  # Remove outliers
+  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
+  labs(title = "Gene Expression Across Visits and Genders", x = "Visit and Sex", y = "Gene Count") +
+  theme_minimal() +
+  theme(plot.title = element_text(size = 10)) +
+  geom_signif(
+    comparisons = list(
+      c("visit_1.F", "visit_2.F"), 
+      c("visit_1.F", "visit_3.F"), 
+      c("visit_2.F", "visit_3.F"),
+      c("visit_1.M", "visit_2.M"), 
+      c("visit_1.M", "visit_3.M"), 
+      c("visit_2.M", "visit_3.M")
+    ),
+    annotations = c(
+      wilcox_stars_F["visit1_vs_visit2"],
+      wilcox_stars_F["visit1_vs_visit3"],
+      wilcox_stars_F["visit2_vs_visit3"],
+      wilcox_stars_M["visit1_vs_visit2"],
+      wilcox_stars_M["visit1_vs_visit3"],
+      wilcox_stars_M["visit2_vs_visit3"]
+    ),
+    textsize = 4,
+    map_signif_level = FALSE
+  )
+
+# Add Kruskal-Wallis p-values as text
+plot <- plot + 
+  annotate("text", x = 1, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Female p-value:", format(kw_pval_F, digits = 3)), 
+           hjust = 0, size = 3) +
+  annotate("text", x = 4, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Male p-value:", format(kw_pval_M, digits = 3)), 
+           hjust = 0, size = 3)
+
+# Display the plot
+print(plot)
+```
+
+SPP1
+
+```{r, f3d3}
+gene_of_interest <- "SPP1"
+
+# Subset norm_counts to get counts for the gene of interest
+gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
+rownames(gene_counts)=gene_counts$hgnc_symbol
+gene_counts=gene_counts[4:96]
+
+# Reshape gene_counts into long format and merge with colData
+gene_counts_long <- gene_counts %>%
+  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
+  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")
+
+gene_counts_long <- gene_counts_long %>%
+  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData
+
+# Define a helper function to map p-values to stars
+pval_to_stars <- function(p) {
+  if (p < 0.0001) {
+    return("****")
+  } else if (p < 0.001) {
+    return("***")
+  } else if (p < 0.01) {
+    return("**")
+  } else if (p < 0.05) {
+    return("*")
+  } else {
+    return("ns")  # Not significant
+  }
+}
+
+# Perform a Kruskal-Wallis test separately for each gender
+kw_res_F <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "F"))
+kw_res_M <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "M"))
+
+# Extract Kruskal-Wallis p-values
+kw_pval_F <- kw_res_F$p.value
+kw_pval_M <- kw_res_M$p.value
+
+# Perform paired Wilcoxon tests for each gender with Holm's correction
+paired_wilcox <- function(data, sex_group) {
+  data_sex <- data %>% filter(sex == sex_group)
+  pairwise_tests <- list(
+    visit1_vs_visit2 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      paired = TRUE
+    )$p.value,
+    visit1_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
+      paired = TRUE
+    )$p.value,
+    visit2_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
+      paired = TRUE
+    )$p.value
+  )
+  corrected_pvals <- p.adjust(unlist(pairwise_tests), method = "holm")
+  return(corrected_pvals)
+}
+
+wilcox_pvals_F <- paired_wilcox(gene_counts_long, "F")
+wilcox_pvals_M <- paired_wilcox(gene_counts_long, "M")
+
+# Map the Wilcoxon p-values to stars
+wilcox_stars_F <- sapply(wilcox_pvals_F, pval_to_stars)
+wilcox_stars_M <- sapply(wilcox_pvals_M, pval_to_stars)
+
+# Create the boxplot with significance annotations
+plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
+  geom_boxplot(outlier.shape = NA) +  # Remove outliers
+  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
+  labs(title = "Gene Expression Across Visits and Genders", x = "Visit and Sex", y = "Gene Count") +
+  theme_minimal() +
+  theme(plot.title = element_text(size = 10)) +
+  geom_signif(
+    comparisons = list(
+      c("visit_1.F", "visit_2.F"), 
+      c("visit_1.F", "visit_3.F"), 
+      c("visit_2.F", "visit_3.F"),
+      c("visit_1.M", "visit_2.M"), 
+      c("visit_1.M", "visit_3.M"), 
+      c("visit_2.M", "visit_3.M")
+    ),
+    annotations = c(
+      wilcox_stars_F["visit1_vs_visit2"],
+      wilcox_stars_F["visit1_vs_visit3"],
+      wilcox_stars_F["visit2_vs_visit3"],
+      wilcox_stars_M["visit1_vs_visit2"],
+      wilcox_stars_M["visit1_vs_visit3"],
+      wilcox_stars_M["visit2_vs_visit3"]
+    ),
+    textsize = 4,
+    map_signif_level = FALSE
+  )
+
+# Add Kruskal-Wallis p-values as text
+plot <- plot + 
+  annotate("text", x = 1, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Female p-value:", format(kw_pval_F, digits = 3)), 
+           hjust = 0, size = 3) +
+  annotate("text", x = 4, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Male p-value:", format(kw_pval_M, digits = 3)), 
+           hjust = 0, size = 3)
+
+# Display the plot
+print(plot)
+```
+
+MMP9
+
+```{r}
+gene_of_interest <- "MMP9"
+
+# Subset norm_counts to get counts for the gene of interest
+gene_counts <- Bulk_norm_counts[Bulk_norm_counts$hgnc_symbol == gene_of_interest, ]
+rownames(gene_counts)=gene_counts$hgnc_symbol
+gene_counts=gene_counts[4:96]
+
+# Reshape gene_counts into long format and merge with colData
+gene_counts_long <- gene_counts %>%
+  tibble::rownames_to_column(var = "gene") %>%  # Create a 'gene' column from row names
+  tidyr::pivot_longer(cols = -gene, names_to = "recordid", values_to = "count")
+
+gene_counts_long <- gene_counts_long %>%
+  left_join(colData, by = "recordid")  # Assuming recordid matches between gene_counts and colData
+
+# Define a helper function to map p-values to stars
+pval_to_stars <- function(p) {
+  if (p < 0.0001) {
+    return("****")
+  } else if (p < 0.001) {
+    return("***")
+  } else if (p < 0.01) {
+    return("**")
+  } else if (p < 0.05) {
+    return("*")
+  } else {
+    return("ns")  # Not significant
+  }
+}
+
+# Perform a Kruskal-Wallis test separately for each gender
+kw_res_F <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "F"))
+kw_res_M <- kruskal.test(count ~ visit, data = gene_counts_long %>% filter(sex == "M"))
+
+# Extract Kruskal-Wallis p-values
+kw_pval_F <- kw_res_F$p.value
+kw_pval_M <- kw_res_M$p.value
+
+# Perform paired Wilcoxon tests for each gender with Holm's correction
+paired_wilcox <- function(data, sex_group) {
+  data_sex <- data %>% filter(sex == sex_group)
+  pairwise_tests <- list(
+    visit1_vs_visit2 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      paired = TRUE
+    )$p.value,
+    visit1_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_1") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
+      paired = TRUE
+    )$p.value,
+    visit2_vs_visit3 = wilcox.test(
+      x = data_sex %>% filter(visit == "visit_2") %>% pull(count),
+      y = data_sex %>% filter(visit == "visit_3") %>% pull(count),
+      paired = TRUE
+    )$p.value
+  )
+  corrected_pvals <- p.adjust(unlist(pairwise_tests), method = "holm")
+  return(corrected_pvals)
+}
+
+wilcox_pvals_F <- paired_wilcox(gene_counts_long, "F")
+wilcox_pvals_M <- paired_wilcox(gene_counts_long, "M")
+
+# Map the Wilcoxon p-values to stars
+wilcox_stars_F <- sapply(wilcox_pvals_F, pval_to_stars)
+wilcox_stars_M <- sapply(wilcox_pvals_M, pval_to_stars)
+
+# Create the boxplot with significance annotations
+plot <- ggplot(gene_counts_long, aes(x = interaction(visit, sex), y = count, fill = visit)) +
+  geom_boxplot(outlier.shape = NA) +  # Remove outliers
+  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
+  labs(title = "Gene Expression Across Visits and Genders", x = "Visit and Sex", y = "Gene Count") +
+  theme_minimal() +
+  theme(plot.title = element_text(size = 10)) +
+  geom_signif(
+    comparisons = list(
+      c("visit_1.F", "visit_2.F"), 
+      c("visit_1.F", "visit_3.F"), 
+      c("visit_2.F", "visit_3.F"),
+      c("visit_1.M", "visit_2.M"), 
+      c("visit_1.M", "visit_3.M"), 
+      c("visit_2.M", "visit_3.M")
+    ),
+    annotations = c(
+      wilcox_stars_F["visit1_vs_visit2"],
+      wilcox_stars_F["visit1_vs_visit3"],
+      wilcox_stars_F["visit2_vs_visit3"],
+      wilcox_stars_M["visit1_vs_visit2"],
+      wilcox_stars_M["visit1_vs_visit3"],
+      wilcox_stars_M["visit2_vs_visit3"]
+    ),
+    textsize = 4,
+    map_signif_level = FALSE
+  )
+
+# Add Kruskal-Wallis p-values as text
+plot <- plot + 
+  annotate("text", x = 1, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Female p-value:", format(kw_pval_F, digits = 3)), 
+           hjust = 0, size = 3) +
+  annotate("text", x = 4, y = max(gene_counts_long$count) * 1.1, 
+           label = paste("KW Male p-value:", format(kw_pval_M, digits = 3)), 
+           hjust = 0, size = 3)
+
+# Display the plot
+print(plot)
+```
+
+## Extended Data Figure 5g-h
+
+```{r}
+# Metadata
+metadata.all <- read.delim("Meta_Data_WL_Select.txt", h=T, dec = ",") %>% 
+  mutate(sex = factor(sex, labels = c("Female", "Male")),
+         donor = gsub("D", "Donor_", recordid) %>% 
+           gsub("V", "vis", .))
+
+# Bulk data
+cm.bulk <- read.delim("Bulk_norm_counts.txt", header = T) %>% 
+  .[match(unique(.$hgnc_symbol), .$hgnc_symbol), ] %>% 
+  `rownames<-`(.$hgnc_symbol) %>% 
+  select(-hgnc_symbol, -ensembl_gene_id, -entrezgene_id) %>% 
+  scale()
+
+colnames(cm.bulk) %<>% 
+  gsub("D", "Donor_", .) %>% 
+  gsub("V", "vis", .)
+```
+
+### Extended Data Figure 5g
+
+See Figure 3d for model setup
+
+```{r}
+# Use Cacoa to calculate proportions
+cao <- Cacoa$new(data.object = con.major, 
+                 cell.groups = anno.major, 
+                 ref.level = "vis2", 
+                 target.level = "vis3", 
+                 sample.groups = con.major$samples %>% 
+                   names() %>% 
+                   strsplit("_") %>% 
+                   sapply('[[', 3) %>% 
+                   gsub(1, 2, .) %>% 
+                   `names<-`(con.major$samples %>% 
+                               names()),
+                 sample.groups.palette = ggsci::pal_jama()(7))
+
+cao$sample.groups <- con.major$samples %>% 
+  names() %>% 
+  strsplit("_") %>% 
+  sapply('[[', 3) %>% 
+  `names<-`(con.major$samples %>% 
+              names()) %>% 
+  gsub("vis", "Visit ", .)
+
+anno.comb <- anno.major[!names(anno.major) %in% names(anno.immune)] %>% 
+  {factor(c(., anno.immune))}
+
+# Get proportions
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(`Early LAM`, rownames(.))
+
+tmp <- qread("eLAM_model.qs")
+
+lasso_model <- tmp$lasso_model
+best_lambda <- tmp$lasso_cv$lambda.min
+
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% rownames(coef(lasso_model, s = best_lambda))) %>% 
+  as.matrix() %>% 
+  t()
+
+# Now, train a final model on all data for visualization and gene selection
+cm.test <- cm.bulk.mat %>% 
+  .[rownames(.) %in% names(proportions), ]
+
+lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 40)
+best_lambda <- lasso_cv$lambda.min
+lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)
+
+# Predict on all samples using the final model
+lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)
+r2.all <- cor(proportions, lasso.predict.all)^2
+
+data.frame(Proportions = proportions,
+           Predictions = unname(lasso.predict.all)) %>% 
+  ggplot(aes(Proportions, Predictions)) + 
+  geom_point() +
+  theme_bw() +
+  theme(axis.text = element_text(colour = "black")) +
+  labs(title = paste0("r2 = ", formatC(r2.all, digits = 3)), y = "Predicted proportions (%)", x = "Proportions (%)") +
+  geom_abline(intercept = 0, slope = 1, colour = "firebrick")
+```
+
+### Extended Data Figure 5h
+
+See Figure 3e for model setup
+
+```{r}
+# Use Cacoa to calculate proportions
+cao <- Cacoa$new(data.object = con.major, 
+                 cell.groups = anno.major, 
+                 ref.level = "vis2", 
+                 target.level = "vis3", 
+                 sample.groups = con.major$samples %>% 
+                   names() %>% 
+                   strsplit("_") %>% 
+                   sapply('[[', 3) %>% 
+                   gsub(1, 2, .) %>% 
+                   `names<-`(con.major$samples %>% 
+                               names()),
+                 sample.groups.palette = ggsci::pal_jama()(7))
+
+cao$sample.groups <- con.major$samples %>% 
+  names() %>% 
+  strsplit("_") %>% 
+  sapply('[[', 3) %>% 
+  `names<-`(con.major$samples %>% 
+              names()) %>% 
+  gsub("vis", "Visit ", .)
+
+anno.comb <- anno.major[!names(anno.major) %in% names(anno.immune)] %>% 
+  {factor(c(., anno.immune))}
+
+# Get proportions
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(LAM, rownames(.))
+
+tmp <- qread("LAM_model.qs")
+
+lasso_model <- tmp$lasso_model
+best_lambda <- tmp$lasso_cv$lambda.min
+
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% rownames(coef(lasso_model, s = best_lambda))) %>% 
+  as.matrix() %>% 
+  t()
+
+# Now, train a final model on all data for visualization and gene selection
+cm.test <- cm.bulk.mat %>% 
+  .[rownames(.) %in% names(proportions), ]
+
+lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 40)
+best_lambda <- lasso_cv$lambda.min
+lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)
+
+# Predict on all samples using the final model
+lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)
+r2.all <- cor(proportions, lasso.predict.all)^2
+
+data.frame(Proportions = proportions,
+           Predictions = unname(lasso.predict.all)) %>% 
+  ggplot(aes(Proportions, Predictions)) + 
+  geom_point() +
+  theme_bw() +
+  theme(axis.text = element_text(colour = "black")) +
+  labs(title = paste0("r2 = ", formatC(r2.all, digits = 3)), y = "Predicted proportions (%)", x = "Proportions (%)") +
+  geom_abline(intercept = 0, slope = 1, colour = "firebrick")
+```
+
+## Extended Data Figure 5i
+
+```{r, fig.width=5, fig.height=4}
+watlas_data <- read.delim("watlas_mapped_percent.tsv", stringsAsFactors = FALSE)
+watlas_data_count <- read.delim("watlas_mapped_counts.tsv", stringsAsFactors = FALSE)
+watlas_data_count <- watlas_data_count %>%
+  rowwise() %>%
+  mutate(row_sum = sum(c_across(where(is.numeric)), na.rm = TRUE))  # Sum across numeric columns
+valid_samples <- watlas_data_count %>%
+  filter(row_sum >= 1000) %>%
+  pull(sample)  # Extracts the valid sample IDs
+
+filtered_data <- watlas_data %>%
+  filter(type == "sn", tissue == "AT", depot == "SAT", grepl("Hs", sample))
+filtered_data <- filtered_data %>%
+  filter(sample %in% valid_samples)
+
+long_data_mac <- filtered_data %>%
+  select(wtstatus, ATM, Early.LAM, LAM, mono.mac) %>%  # Select relevant columns
+  tidyr::pivot_longer(cols = -wtstatus, names_to = "Cell_Type", values_to = "Percentage")
+
+long_data_mac %>% 
+  ggplot(aes(x = Cell_Type, y = Percentage, fill = wtstatus)) +
+  geom_boxplot(position = position_dodge(width = 0.8)) + 
+  geom_point(aes(group = wtstatus), position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.8), alpha = 0.3, ) +
+  theme_bw() +
+  labs(title = "Comparison of Cell Types in Lean vs Obese",
+       x = "Cell Type",
+       y = "Percentage",
+       fill = "") +
+  scale_fill_manual(values = c(lean = "#26547c", obese = "#ef476f")) +
+  stat_compare_means(aes(group = wtstatus), method = "wilcox.test", label = "p.signif", ) +
+  theme(axis.text = element_text(color = "black"),
+        line = element_blank())
+```
+
+## Extended Data Figure 5j
+
+```{r, fig.width=6, fig.height=4}
+tmp <- con.integration$clusters$loft_transfer_minor$groups %>%
+  {data.frame(cid = names(.), anno = unname(.))} %>% 
+  filter(!grepl("!!", cid)) %>% 
+  mutate(sample = con.integration$clusters$hinte_condition$groups[match(cid, names(con.integration$clusters$hinte_condition$groups))])
+
+tmp.sum <- tmp %>% 
+  group_by(sample) %>% 
+  summarize(total = n()) %>% 
+  split(., .$sample)
+
+tmp.sum.ct <- tmp %>% 
+  group_by(sample, anno) %>% 
+  summarize(total = n()) %>% 
+  split(., .$sample)
+
+tmp.prop <- Map(\(ct, tot, an) setNames(ct$total / tot$total, an), ct = tmp.sum.ct, tot = tmp.sum, an = lapply(tmp.sum.ct, pull, anno))
+
+df.plot <- tmp.prop %>%
+  {Map(\(nn, x) data.frame(sample = nn, anno = names(x), value = unname(x)), nn = names(.), x = .)} %>% 
+  bind_rows() %>% 
+  mutate(anno = factor(anno, levels = unique(anno)),
+         type = sapply(sample, \(x) if (grepl("lean", x)) "Lean" else if (grepl("t0", x)) "Obese" else "Regressed"))
+
+cts <- c("ATM", "mono/mac", "Early LAM", "LAM")
+
+df.plot %>% 
+  filter(anno %in% cts) %>% 
+  mutate(anno = factor(anno, levels = cts)) %>% 
+  ggplot(aes(anno, value)) + 
+  stat_summary(mapping = aes(fill = type), geom = "bar", fun = "mean", position = "dodge") +
+  geom_point(aes(col = type), position = position_dodge(width = 0.9)) +
+  theme_bw() +
+  labs(x = "", y = "% proportion per sample", title = "Loft to Hinte transfer, minor annotation", col = "", fill = "") +
+  theme(axis.text = element_text(colour = "black"),
+        line = element_blank(),
+        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
+  scale_fill_manual(values = brewer.pal(9, "Greens")[c(3, 6, 9)]) +
+  scale_color_manual(values = brewer.pal(9, "Purples")[c(9, 7, 5)])
+```
+
+## Extended Data Figure 5k
 
@@
 
-Load data for S3H-J first
+Load data for S3H-J first !!!
 
@@
 
-# Figure S4
-
-## Figure S4A-C
+## Extended Data Figure 5l
+
+Load and prepare
+
+```{r}
+cm.merged.raw <- con.integration$getJointCountMatrix(raw = T)
+cts <- c("ATM", "mono/mac", "Early LAM", "LAM")
+anno <- getConosCluster(con.integration, "loft_transfer_minor")
+anno.subtype <- anno[anno %in% cts]
+```
+
+```{r, fig.width=6, fig.height=5}
+# Create sample-wise annotation
+anno.donor <- con.integration$getDatasetPerCell()[rownames(cm.merged.raw)] %>% 
+  .[!grepl("!!", names(.))] %>% 
+  factor()
+
+idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())
+
+anno.donor %<>% .[idx]
+anno.subtype %<>% .[idx] %>% 
+  {`names<-`(as.character(.), names(.))}
+
+anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
+  `names<-`(anno.donor %>% names())
+
+cm.pseudo.tmp <- sccore::collapseCellsByType(cm.merged.raw,
+                                             groups = anno.final, min.cell.count = 1) %>% 
+  t() %>%
+  apply(2, as, "integer")
+
+cm.pseudo <- cm.pseudo.tmp %>% 
+  DESeq2::DESeqDataSetFromMatrix(., 
+                                 colnames(.) %>% 
+                                   strsplit("!!") %>% 
+                                   sget(1) %>%
+                                   data.frame() %>% 
+                                   `dimnames<-`(list(colnames(cm.pseudo.tmp), "group")), 
+                                 design = ~ group) %>% 
+  DESeq2::estimateSizeFactors() %>% 
+  DESeq2::counts(normalized = TRUE)
+
+idx <- cm.pseudo %>% 
+  colnames() %>% 
+  data.frame(id = .) %>% 
+  mutate(type = strsplit(id, "_|!!") %>% 
+           sapply(\(x) if (length(x) == 3) x[2] else x[1]),
+         ct = strsplit(id, "!!") %>% sget(2) %>% 
+           factor(levels = cts)) %>% 
+  mutate(ord = order(ct, type))
+
+genes.order <- read.delim("Macrophages_DEG_heatmap_genes.tsv") %$% 
+  setNames(gene, cluster) %>%
+  .[order(names(.))]
+
+x <- cm.pseudo %>% 
+  .[genes.order, ] %>%
+  .[rowSums(.) > 0, idx$ord] %>% 
+  Matrix::t() %>%
+  scale() %>%
+  Matrix::t()
+
+row.order <- names(genes.order)[genes.order %in% rownames(x)]
+
+groups <- colnames(x) %>%
+  `names<-`(colnames(x))
+
+# Limit scale
+x[x > 2] <- 2
+x[x < -1.5] <- -1.5
+
+# Create top annotation
+tannot <- idx %$%
+  .[match(colnames(x), id), ] %>% 
+  select(type) %>% 
+  dplyr::rename(Type = type) %$% 
+  {HeatmapAnnotation(df=.,
+                     border=T,
+                     col=list(Type = setNames(brewer.pal(9, "Greens")[c(3, 6, 9)], unique(Type))), 
+                     show_legend=T)}
+
+# Create color palette
+pal <- colorRampPalette(c('navy','grey95','firebrick'))(1024)
+labeled.gene.subset <- NULL
+
+# Plot
+set.seed(1337)
+
+ha <- ComplexHeatmap::Heatmap(x, 
+                              name='Expression', 
+                              col=pal, 
+                              cluster_columns=FALSE, 
+                              cluster_rows = F,
+                              show_row_names=FALSE, 
+                              show_column_names=FALSE, 
+                              top_annotation=tannot,
+                              left_annotation=NULL,
+                              border=TRUE,
+                              show_column_dend = FALSE, 
+                              show_row_dend = FALSE,
+                              row_split = row.order,
+                              # row_km = 5,
+                              # row_km_repeats = 500,
+                              column_split = idx %$% 
+                                .[match(colnames(x), id), ] %>% 
+                                pull(ct) %>% 
+                                factor(levels = cts) %>% 
+                                levels() %>%
+                                {c("C1", rep(.[1], 4), "C2", rep(.[2], 4), "C3", rep(.[3], 4), "C4", rep(.[4], 4))} %>%
+                                factor(., levels = unique(.)),
+                              column_gap = unit(rep(0:1, 4), "mm") %>% .[-length(.)],
+                              border_gp = gpar(col = "black", lwd = 1.5))
+
+ht = draw(ha)
+```
+
+## Extended Data Figure 5m
+
+```{r}
+cts <- c("Early LAM", "LAM", "mono/mac", "ATM")
+anno <- getConosCluster(con.integration, "loft_transfer_minor") %>% 
+  .[. %in% cts] %>% 
+  factor()
+cm.clean <- con.integration$getJointCountMatrix(raw = T) %>%
+  .[rownames(.) %in% names(anno), ]
+spc <- con.integration$getDatasetPerCell()
+
+markers <- con$getDifferentialGenes(groups = anno, 
+                                    z.threshold = 1, 
+                                    upregulated.only = TRUE, 
+                                    append.specificity.metrics = TRUE, 
+                                    append.auc = TRUE)
+
+genes <- markers$LAM %>% 
+  filter(AUC > 0.8) %>% 
+  pull(Gene)
+
+lam.scores <- list(LAM.activity = genes)
+seu_obj <- CreateSeuratObject(counts = cm.clean %>% Matrix::t() %>% drop0(),
+                              min.cells = 0, 
+                              min.features = 0)
+m.scores <- UCell::AddModuleScore_UCell(seu_obj, features = lam.scores, ncores = 32)
+```
+
+```{r, fig.width = 6, fig.height = 4}
+tmp <- reshape2::melt(m.scores$LAM.activity_UCell) %>% 
+  mutate(anno = anno[rownames(.)],
+         group = spc[rownames(.)] %>% as.character() %>% strsplit("_") %>% sapply(\(x) if (length(x) == 3) x[3] else if (length(x) == 2) paste(x[1], x[2], sep = "_") else x[1])) %>% 
+  mutate(type = grepl.replace(group, unique(group), c("obese", "short-term WL", "regressed", "lean", "obese", "regressed", "obese", "regressed"))) %>% 
+  mutate(anno = anno[rownames(.)] %>% factor(levels = c("ATM", "mono/mac", "Early LAM", "LAM"))) %>% 
+  mutate(type_group = paste(type, group, sep = "_") %>% factor(levels = c("obese_rg_t0", "obese_st_t0", "obese_vis1", "short-term WL_vis2", "lean_lean", "regressed_rg_t1", "regressed_st_t1", "regressed_vis3"))) %>%
+  group_by(group, type, anno) %>%
+  summarize(value = mean(value)) %>%
+  mutate(group = factor(group)) %>% 
+  arrange(group, anno)
+
+dat.plot1 <- tmp %>% 
+  filter(!group %in% c("lean")) %$% 
+  split(., type) %>% 
+  {rbind(data.frame(.[[2]], diff = .[[2]]$value - .[[1]]$value), data.frame(.[[3]], diff = .[[3]]$value - .[[1]]$value[.[[1]]$group == "vis1"]))}  %>%
+  mutate(group = factor(group, labels = c("Hinte_rg", "Hinte_st", "Loft_short-term WL", "Loft_long-term WL"))) %>% 
+  filter(group %in% c("Hinte_rg", "Hinte_st")) %>% 
+  mutate(group = factor(group),
+         group2 = "t1-t0")
+
+dat.plot2 <- tmp %>% 
+  filter(group %in% c("lean", "rg_t0", "st_t0")) %>% 
+  mutate(group = factor(group)) %$% 
+  split(., group) %>% 
+  {rbind(data.frame(.[[2]], diff = .[[1]]$value - .[[2]]$value), data.frame(.[[3]], diff = .[[1]]$value - .[[3]]$value))} %>%
+  mutate(group = factor(group, labels = c("lean_rg", "lean_st"))) %>%
+  mutate(group = factor(group),
+         group2 = "lean-t0")
+
+dat.plot <- rbind(dat.plot1, dat.plot2) %>% 
+  mutate(group = as.character(group) %>% strsplit("_") %>% sget(2) %>% factor())
+
+dat.plot %>% 
+  ggplot(aes(anno, diff, fill = group2)) +
+  stat_summary(geom = "bar", fun = "mean", position = "dodge") +
+  geom_point(aes(col = group, group = group2), position = position_dodge(width = 0.8)) +
+  geom_hline(yintercept = 0, color = "black") +
+  theme_bw() +
+  labs(x = "", y = "Difference in mean UCell module scores for\ntop gene marker expression", title = "LAM top markers module scores", fill = "Difference", col = "Dataset") +
+  theme(line = element_blank(),
+        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
+        axis.text = element_text(color = "black")) +
+  scale_fill_manual(values = c("grey70", "grey40")) +
+  scale_color_manual(values = brewer.pal(9, "Purples")[c(5,9)])
+```
+
+# Extended Data Figure 6
+
+## Extended Data Figure 6a-c
 
@@
 
-### Figure S4A
+### Extended Data Figure 6a
 
@@
 
-### Figure S4B
+### Extended Data Figure 6b
 
@@
 
-### Figure S4C
+### Extended Data Figure 6c
 
@@
 
-## Figure S4D
+## Extended Data Figure 6d
 
@@
 
-## Figure S4E
+## Extended Data Figure 6e
 
@@
 
-## Figure S4F-G
+## Extended Data Figure 6f
+
+Calculate
+
+```{r}
+anno.minor <- qread("anno_minor.qs")
+
+spc <- con.major$getDatasetPerCell() %>% 
+  .[match(names(anno.minor), names(.))]
+
+anno.minor.sex <- anno.minor %>% 
+  data.frame(anno = ., cid = names(.), donor = unname(spc)) %>% 
+  mutate(sex = meta$sex[match(donor, meta$sample)]) %>% 
+  mutate(anno_sex = paste(anno, sex, sep = "_")) %$% 
+  setNames(anno_sex, cid) %>% 
+  factor()
+
+# Calculate
+res.minor <- c("vis1", "vis2", "vis3") %>% 
+  sapply(\(vis) {
+    con <- con.major$samples %>% 
+      .[!grepl(vis, names(.))] %>%
+      Conos$new()
+
+    if (vis == "vis1") {
+      vis.ref <- "vis2"
+      vis.target <- "vis3" 
+    } else {
+      vis.ref <- "vis1"
+      if (vis == "vis2") {
+        vis.target <- "vis3"
+      } else {
+        vis.target <- "vis2"
+      }
+    }
+
+    sample.groups <- con$samples %>%
+      names() %>%
+      `names<-`(ifelse(grepl(vis.ref, .), vis.ref, vis.target), .)
+
+    cao <- Cacoa$new(con,
+                     sample.groups,
+                     anno.minor.sex,
+                     ref.level = vis.ref,
+                     target.level = vis.target,
+                     n.cores = 32)
+
+    cao$estimateCellLoadings()
+
+    return(cao)
+  }, simplify = F, USE.NAMES = T)
+```
+
+Plot
+
+```{r}
+aspc.pal <- RColorBrewer::brewer.pal(5, "Greens")[-1][c(2,1,3,4)] %>% 
+  setNames(levels(qread("/work/02_data/09_export/anno_aspc.qs"))) %>% 
+  {setNames(c(rep(., 3)), c(names(.), paste(names(.), "Female", sep = "_"), paste(names(.), "Male", sep = "_")))}
+
+visits <- c("vis1", "vis2", "vis3")
+
+p <- visits %>% 
+  lapply(\(vis) {
+    comp <- visits[!visits == vis]
+
+    dat.p <- res.minor[[vis]]$test.results$coda$padj
+
+    dat.plot <- res.minor[[vis]]$test.results$coda$loadings %>% 
+      tibble::rownames_to_column(var = "anno") %>% 
+      reshape2::melt(id.var = "anno") %>% 
+      filter(anno %in% names(aspc.pal))
+
+    yvar <- dat.plot %>%
+      group_by(anno) %>%
+      summarize(var = mean(value)) %>%
+      mutate(padj = dat.p[match(anno, names(dat.p))]) %>%
+      arrange(padj, desc(abs(var)))
+
+    dat.plot %<>% 
+      mutate(anno = factor(anno, levels = rev(unique(yvar$anno))))
+
+    ymax <- dat.plot %>%
+      pull(value) %>%
+      abs() %>%
+      max()
+
+    out <- dat.plot %>% 
+      ggplot(aes(anno, value, fill = anno)) + 
+      geom_hline(yintercept = 0, col = "black") +
+      geom_jitter(alpha = 0.2, size = 0.3, col = "grey") +
+      geom_boxplot(outliers = F) +
+      coord_flip() +
+      ylim(c(-ymax, ymax)) +
+      theme_bw() + 
+      theme(line = element_blank(),
+            axis.text = element_text(color = "black")) +
+      guides(fill = "none") +
+      labs(title = paste0(comp[1], " vs ", comp[2]), x = "", y = "Loadings [AU]") +
+      scale_fill_manual(values = aspc.pal)
+
+    return(out)
+  })
+
+p
+```
+
+Statistics
+
+```{r}
+for (vis in visits) {
+  comp <- visits[!visits == vis]
+  message("Comparison: ", comp)
+  print(res.minor[[vis]]$test.results$coda$padj %>% .[names(.) %in% names(aspc.pal)])
+}
+```
+
+## Extended Data Figure 6g
+
+See Figure 4d for model setup
+
+```{r}
+# Use Cacoa to calculate proportions
+cao <- Cacoa$new(data.object = con.major, 
+                 cell.groups = anno.major, 
+                 ref.level = "vis2", 
+                 target.level = "vis3", 
+                 sample.groups = con.major$samples %>% 
+                   names() %>% 
+                   strsplit("_") %>% 
+                   sapply('[[', 3) %>% 
+                   gsub(1, 2, .) %>% 
+                   `names<-`(con.major$samples %>% 
+                               names()),
+                 sample.groups.palette = ggsci::pal_jama()(7))
+
+cao$sample.groups <- con.major$samples %>% 
+  names() %>% 
+  strsplit("_") %>% 
+  sapply('[[', 3) %>% 
+  `names<-`(con.major$samples %>% 
+              names()) %>% 
+  gsub("vis", "Visit ", .)
+
+anno.comb <- anno.major[!names(anno.major) %in% names(anno.aspc)] %>% 
+  {factor(c(., anno.aspc))}
+
+# Get proportions
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(ASPC_DPP4, rownames(.))
+
+tmp <- qread("DPP4_model.qs")
+
+lasso_model <- tmp$lasso_model
+best_lambda <- tmp$lasso_cv$lambda.min
+
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% rownames(coef(lasso_model, s = best_lambda))) %>% 
+  as.matrix() %>% 
+  t()
+
+# Now, train a final model on all data for visualization and gene selection
+cm.test <- cm.bulk.mat %>% 
+  .[rownames(.) %in% names(proportions), ]
+
+lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 40)
+best_lambda <- lasso_cv$lambda.min
+lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)
+
+# Predict on all samples using the final model
+lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)
+r2.all <- cor(proportions, lasso.predict.all)^2
+
+data.frame(Proportions = proportions,
+           Predictions = unname(lasso.predict.all)) %>% 
+  ggplot(aes(Proportions, Predictions)) + 
+  geom_point() +
+  theme_bw() +
+  theme(axis.text = element_text(colour = "black")) +
+  labs(title = paste0("r2 = ", formatC(r2.all, digits = 3)), y = "Predicted proportions (%)", x = "Proportions (%)") +
+  geom_abline(intercept = 0, slope = 1, colour = "firebrick")
+```
+
+# Extended Data Figure 7
+
+## Extended Data Figure 7a,d
 
@@
 
-### Figure S4F
+### Extended Data Figure 7a
 
@@
 
-### Figure S4G
+### Extended Data Figure 7d
 
@@
 
-## Figure S4H
+## Extended Data Figure 7b
+
+```{r, fig.width=6, fig.height=5}
+cts <- c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")
+anno <- getConosCluster(con.integration, "loft_transfer_minor")
+anno.subtype <- anno[anno %in% cts]
+
+# Create sample-wise annotation
+anno.donor <- con.integration$getDatasetPerCell()[rownames(cm.merged.raw)] %>% 
+  .[!grepl("!!", names(.))] %>% 
+  factor()
+
+idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())
+
+anno.donor %<>% .[idx]
+anno.subtype %<>% .[idx] %>% 
+  {`names<-`(as.character(.), names(.))}
+
+anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
+  `names<-`(anno.donor %>% names())
+
+cm.pseudo.tmp <- sccore::collapseCellsByType(cm.merged.raw,
+                                             groups = anno.final, min.cell.count = 1) %>% 
+  t() %>%
+  apply(2, as, "integer")
+
+cm.pseudo <- cm.pseudo.tmp %>% 
+  DESeq2::DESeqDataSetFromMatrix(., 
+                                 colnames(.) %>% 
+                                   strsplit("!!") %>% 
+                                   sget(1) %>%
+                                   data.frame() %>% 
+                                   `dimnames<-`(list(colnames(cm.pseudo.tmp), "group")), 
+                                 design = ~ group) %>% 
+  DESeq2::estimateSizeFactors() %>% 
+  DESeq2::counts(normalized = TRUE)
+
+idx <- cm.pseudo %>% 
+  colnames() %>% 
+  data.frame(id = .) %>% 
+  mutate(type = strsplit(id, "_|!!") %>% 
+           sapply(\(x) if (length(x) == 4) x[2] else x[1]),
+         ct = strsplit(id, "!!") %>% sget(2) %>% 
+           factor(levels = cts)) %>% 
+  mutate(ord = order(ct, type))
+
+genes.order <- read.delim("ASPC_DEG_heatmap_genes.tsv") %$% 
+  setNames(gene, cluster) %>%
+  .[order(names(.))]
+
+x <- cm.pseudo %>% 
+  .[genes.order, ] %>%
+  .[rowSums(.) > 0, idx$ord] %>% 
+  Matrix::t() %>%
+  scale() %>%
+  Matrix::t()
+
+row.order <- names(genes.order)[genes.order %in% rownames(x)]
+
+groups <- colnames(x) %>%
+  `names<-`(colnames(x))
+
+# Limit scale
+x[x > 2] <- 2
+x[x < -1.5] <- -1.5
+
+# Create top annotation
+tannot <- idx %$%
+  .[match(colnames(x), id), ] %>% 
+  select(type) %>% 
+  dplyr::rename(Type = type) %$% 
+  {HeatmapAnnotation(df=.,
+                     border=T,
+                     col=list(Type = setNames(brewer.pal(9, "Greens")[c(3, 6, 9)], unique(Type))), 
+                     show_legend=T)}
+
+# Create color palette
+pal <- colorRampPalette(c('navy','grey95','firebrick'))(1024)
+labeled.gene.subset <- NULL
+
+# Plot
+set.seed(1337)
+
+ha <- ComplexHeatmap::Heatmap(x, 
+                              name='Expression', 
+                              col=pal, 
+                              cluster_columns=FALSE, 
+                              cluster_rows = F,
+                              show_row_names=FALSE, 
+                              show_column_names=FALSE, 
+                              top_annotation=tannot,
+                              left_annotation=NULL,
+                              border=TRUE,
+                              show_column_dend = FALSE, 
+                              show_row_dend = FALSE,
+                              row_split = row.order,
+                              column_split = idx %$% 
+                                .[match(colnames(x), id), ] %>% 
+                                pull(ct) %>% 
+                                factor(levels = cts) %>% 
+                                levels() %>%
+                                {c("C1", rep(.[1], 4), "C2", rep(.[2], 4), "C3", rep(.[3], 4), "C4", rep(.[4], 4))} %>%
+                                factor(., levels = unique(.)),
+                              column_gap = unit(rep(0:1, 4), "mm") %>% .[-length(.)],
+                              border_gp = gpar(col = "black", lwd = 1.5))
+
+ht = draw(ha)
+```
+
+## Extended Data Figure 7c
+
+```{r}
+# We extracted genes from the selected GO term using this snippet
+# genes.go <- clusterProfiler::bitr("GO:0097193", 
+#                                   fromType="GOALL", 
+#                                   toType="SYMBOL", 
+#                                   OrgDb='org.Hs.eg.db')$SYMBOL %>% 
+#   unique()
+
+# Then, we intersected with the DEG genes plotted in our heatmap. The resulting genes are the following:
+genes <- c("CLU", "CRIP1", "GRINA", "HNRNPK", "MMP2", "P4HB", "CXCL12", "TMBIM6", "ARL6IP5")
+
+cts <- c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3")
+anno <- getConosCluster(con.integration, "loft_transfer_minor")
+anno.subtype <- anno[anno %in% cts] %>% 
+  factor()
+cm.merged.sub <- cm.merged.raw %>% 
+  .[rownames(.) %in% names(anno.subtype), ]
+
+cm.clean <- cm.merged.sub[rownames(cm.merged.sub) %in% names(anno.subtype), ]
+lam.scores <- list(LAM.activity = genes)
+seu_obj <- CreateSeuratObject(counts = cm.clean %>% Matrix::t() %>% drop0(),
+                              min.cells = 0, 
+                              min.features = 0)
+m.scores <- UCell::AddModuleScore_UCell(seu_obj, features = lam.scores, ncores = 32)
+```
+
+Prepare
+
+```{r}
+cm.sub <- m.scores$LAM.activity_UCell
+
+# Create sample-wise annotation
+anno.donor <- con.integration$getDatasetPerCell()[names(cm.sub)]
+anno.subtype <- anno.subtype %>% 
+  .[!is.na(.)] %>% 
+  factor()
+
+idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())
+
+anno.donor %<>% .[idx]
+anno.subtype %<>% .[idx] %>% 
+  {`names<-`(as.character(.), names(.))}
+
+anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
+  `names<-`(anno.donor %>% names())
+
+# Prepare plot
+plot.tmp <- cm.sub %>% 
+  data.frame(cid = names(.), value = .) %>% 
+  mutate(donor_anno = anno.final[cid]) %>% 
+  mutate(donor = strsplit(donor_anno, "!!") %>% 
+           sget(1),
+         anno = strsplit(donor_anno, "!!") %>% 
+           sget(2),
+         sex = meta$sex[match(donor, meta$sample)]) %>% 
+  mutate(visit = strsplit(donor, "_") %>% 
+           sapply(\(x) if (length(x) == 3) x[3] else if (length(x) == 2) x[2] else x[1]),
+         type = strsplit(donor, "_") %>% sget(1)) %>% 
+  mutate(sex_visit = paste(sex, visit, sep = "_") %>% 
+           gsub("NA_", "", .) %>% 
+           factor(levels = c("Female_vis1", "Female_vis2", "Female_vis3", "Male_vis1", "Male_vis2", "Male_vis3", "lean", "t0", "t1"))) %>% 
+  group_by(donor_anno, donor, anno, sex, sex_visit, type, visit) %>% 
+  summarize(value = mean(value))
+
+plot.tmp %>% 
+  filter(sex_visit %in% c("lean", "t0", "t1")) %>% 
+  mutate(anno = factor(anno, levels = cts)) %>% 
+  ggplot(aes(anno, value, fill = sex_visit)) +
+  stat_summary(geom = "bar", fun = "mean", position = "dodge") +
+  geom_point(aes(col = type, group = sex_visit), position = position_dodge(width = 0.9)) +
+  theme_bw() + 
+  theme(axis.text = element_text(color = "black"),
+        line = element_blank(),
+        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
+  labs(x = "", y = "Mean UCell module score per sample", fill = "Type", title = "'Intrinsic apoptotic signaling pathway' - cluster 1", col  = "Dataset") +
+  scale_fill_manual(values = brewer.pal(9, "Greens")[c(3, 6, 9)]) +
+  scale_color_manual(values = brewer.pal(9, "Purples")[c(9, 7, 5)])
+```
+
+## Extended Data Figure 7e
 
@@
 
-# Figure S5
+## Extended Data Figure 7f
+
+Here, we're showing how we calculated regulon activity from the output files from Scenic (see `Scenic.ipynb`). This is not run here.
+
+```{r, eval = F}
+reg <- read.table("aspc_adi_macro_reg.csv", sep = ",", header = T, skip = 1) %>% 
+  `colnames<-`(., c(.[1, 1:2], colnames(.)[-c(1,2)])) %>% 
+  dplyr::slice(-1)
+
+tfs <- c("FOS", "FOSB", "JUN", "JUND", "JUNB", "ZEB1", "EBF1", "PPARG", "EBF3", "EBF2")
+
+geneSets <- reg %>% 
+  filter(TF %in% tfs) %$% 
+  split(., TF) %>% 
+  lapply(filter, AUC == max(AUC)) %>% 
+  lapply(pull, TargetGenes) %>%
+  lapply(gsub, pattern = "\\[|\\(|\\'|\b'|\\]|\\)", replacement = "") %>% 
+  lapply(strsplit, ", ") %>% 
+  lget(1) %>% 
+  lapply(\(x) x[seq(1, length(x)-1, 2)])
+
+anno <- getConosCluster(con.integration, "loft_transfer_minor")
+
+anno.sub <- anno %>% 
+  .[. %in% c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3", "Adipocytes")]
+
+cm.sub <- cm.merged[rownames(cm.merged) %in% names(anno.sub), ] %>% 
+  Matrix::t()
+
+auc <- AUCell_run(cm.sub, geneSets)
+
+tmp.regulons <- auc@assays@data@listData$AUC %>%
+  {lapply(split(., rownames(.)), `names<-`, colnames(.))}
+
+qsave(tmp.regulons, "regulons.qs")
+qsave(geneSets, "genesets.qs")
+```
+
+```{r}
+cm <- qread("regulons.qs")[c("FOS", "JUN", "JUND", "JUNB")] %>%
+  as.data.frame()
+
+spc.hinte <- con.integration$getDatasetPerCell() %>% 
+  .[!grepl("!!", names(.))]
+
+anno.sub <- anno %>% 
+  .[. %in% c("ASPC_DPP4", "ASPC_CXCL14", "ASPC_PPARG", "ASPC_EPHA3", "Adipocytes")]
+
+cg <- anno.sub[!anno.sub == "Adipocytes"] %>% 
+  .[match(names(spc.hinte), names(.), nomatch = F)]
+
+spc.final <- spc.hinte[names(cg)] %>% 
+  {setNames(strsplit(as.character(.), "_") %>% sapply(\(x) if (length(x) == 1) x else x[2]), names(.))}
+
+cg.visit <- data.frame(anno = unname(cg), sample = unname(spc.final), row.names = names(cg)) %>% 
+  mutate(anno_sample = paste(anno, sample, sep = "_")) %$% 
+  setNames(anno_sample, rownames(.)) %>% 
+  factor(levels = rev(c("ASPC_DPP4_lean", "ASPC_DPP4_t0", "ASPC_DPP4_t1", "ASPC_CXCL14_lean", "ASPC_CXCL14_t0", "ASPC_CXCL14_t1", "ASPC_PPARG_lean", "ASPC_PPARG_t0", "ASPC_PPARG_t1", "ASPC_EPHA3_lean", "ASPC_EPHA3_t0", "ASPC_EPHA3_t1")))
+
+dotPlot(c("FOS", "JUN", "JUND", "JUNB"), cm, cg.visit, gene.order = rev(c("FOS", "JUN", "JUND", "JUNB")), ) +
+  coord_flip() +
+  guides(size = "none", col = "none") +
+  guides(size = "legend", col = "colorbar") +
+  labs(size = "Fraction of cells in group (%)", col = "Regulon activity (mean AUC)", x = "", y = "") +
+  scale_color_gradient(low = "grey", high = "firebrick")
+```
+
+# Extended Data Figure 8
 
@@
 
-## Figure S5A
+## Extended Data Figure 8a
 
@@
 
-## Figure S5B
+## Extended Data Figure 8b
 
@@
 
-## Figure S5C
+## Extended Data Figure 8c
 
@@
 
-## Figure S5D
+## Extended Data Figure 8d
 
@@
 
-## Figure S5E
+## Extended Data Figure 8e
 
@@
 
-## Figure S5F
+## Extended Data Figure 8f
 
@@
 
-## Figure S5G
+## Extended Data Figure 8g
 
@@
 
-## Figure S5H-I
+## Extended Data Figure 8h-i
 
@@
 
-### Figure S5H
+### Extended Data Figure 8h
 
@@
 
-### Figure S5I
+### Extended Data Figure 8i
 
@@
 
-# Figure S6
-
-## Figure S6A-C
+## Extended Data Figure 8j
+
+First, we integrate adipocytes from both studies. This is not run here
+
+```{r, fig.width=12, fig.height=4, warning=F, eval = F}
+con.adipocytes <- qread("con_adipocytes.qs", nthreads = 10)
+anno.adi <- qread("anno_adipocytes_archetypes.qs")
+
+anno <- getConosCluster(con.integration, "loft_transfer")
+
+anno.hinte <- anno %>%
+  .[!grepl("!!", names(.))] %>%
+  .[. == "Adipocytes"]
+
+samples.processed <- con.integration$samples %>%
+  .[!grepl("vis", names(.))] %>%
+  lapply(conos:::getRawCountMatrix) %>%
+  lapply(\(cm) cm[, colnames(cm) %in% names(anno.hinte)]) %>%
+  lapply(basicP2proc, n.cores = 32, get.largevis = F, get.tsne = F, make.geneknn = F)
+
+con.adipocytes$addSamples(samples.processed)
+con.adipocytes$buildGraph()
+con.adipocytes$findCommunities(resolution = 0.5)
+con.adipocytes$embedGraph()
+anno.archetypes <- con.adipocytes$propagateLabels(anno.adi)
+
+qsave(con.adipocytes, "con_adipocytes_integrated.qs", nthreads = 10)
+qsave(anno.archetypes, "anno_archetypes_integrated.qs")
+```
+
+Load final data
+
+```{r, fig.width=6, fig.height=5}
+anno.archetypes <- qread("anno_archetypes_integrated.qs")
+con.adipocytes <- qread("con_adipocytes_integrated.qs", nthreads = 10)
+
+# We pick top1k cells per subtype based on the uncertainty parameter from the label propagation
+anno.df <- anno.archetypes %$% 
+  data.frame(uncertainty = unname(uncertainty),
+             label = unname(labels),
+             cid = names(labels)) %>% 
+  filter(!grepl("!!", cid)) %$% 
+  split(., label) %>% 
+  lapply(arrange, uncertainty) %>% 
+  lapply(dplyr::slice, seq(1e3)) %>% 
+  bind_rows()
+
+anno.archetypes.hinte <- anno.df %$% 
+  setNames(label, cid) %>% 
+  factor()
+
+anno.subtype <- anno.archetypes.hinte
+
+# Create sample-wise annotation
+anno.donor <- con.integration$getDatasetPerCell()[rownames(cm.merged.raw)] %>% 
+  .[!grepl("!!", names(.))] %>% 
+  factor()
+
+idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())
+
+anno.donor %<>% .[idx]
+anno.subtype %<>% .[idx] %>% 
+  {`names<-`(as.character(.), names(.))}
+
+anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
+  `names<-`(anno.donor %>% names())
+
+cm.pseudo.tmp <- sccore::collapseCellsByType(cm.merged.raw,
+                                             groups = anno.final, min.cell.count = 1) %>% 
+  t() %>%
+  apply(2, as, "integer")
+
+cm.pseudo <- cm.pseudo.tmp %>% 
+  DESeq2::DESeqDataSetFromMatrix(., 
+                                 colnames(.) %>% 
+                                   strsplit("!!") %>% 
+                                   sget(1) %>%
+                                   data.frame() %>% 
+                                   `dimnames<-`(list(colnames(cm.pseudo.tmp), "group")), 
+                                 design = ~ group) %>% 
+  DESeq2::estimateSizeFactors() %>% 
+  DESeq2::counts(normalized = TRUE)
+
+idx <- cm.pseudo %>% 
+  colnames() %>% 
+  data.frame(id = .) %>% 
+  mutate(type = strsplit(id, "_|!!") %>% 
+           sapply(\(x) if (length(x) == 3) x[1] else x[2]),
+         ct = strsplit(id, "!!") %>% sget(2)) %>% 
+  mutate(ord = order(ct, type))
+
+# Order genereted in Figure 5c
+genes.order <- read.delim("Archetypes_DEG_heatmap_genes.tsv") %$% 
+  setNames(gene, cluster %>% factor(labels = seq(4, 1, -1)) %>% as.character()) %>% 
+  .[order(names(.))]
+
+x <- cm.pseudo %>% 
+  .[genes.order, ] %>%
+  .[rowSums(.) > 0, idx$ord] %>%
+  Matrix::t() %>%
+  scale() %>%
+  Matrix::t()
+
+groups <- colnames(x) %>%
+  `names<-`(colnames(x))
+
+# Limit scale
+x[x > 2] <- 2
+x[x < -1] <- -1
+
+# Create top annotation
+tannot <- idx %$%
+  .[match(colnames(x), id), ] %>% 
+  select(type) %>% 
+  dplyr::rename(Type = type) %$% 
+  {HeatmapAnnotation(df=.,
+                     border=T,
+                     col=list(Type = setNames(brewer.pal(9, "Greens")[c(3, 6, 9)], unique(Type))), 
+                     show_legend=T)}
+
+# Create color palette
+pal <- colorRampPalette(c('navy','grey95','firebrick'))(1024)
+labeled.gene.subset <- NULL
+
+# Plot
+set.seed(1337)
+
+ha <- ComplexHeatmap::Heatmap(x, 
+                              name='Expression', 
+                              col=pal, 
+                              cluster_columns=FALSE, 
+                              cluster_rows = F,
+                              show_row_names=FALSE, 
+                              show_column_names=FALSE, 
+                              top_annotation=tannot,
+                              left_annotation=NULL,
+                              border=TRUE,
+                              show_column_dend = FALSE, 
+                              show_row_dend = FALSE,
+                              row_split = names(genes.order)[genes.order %in% rownames(x)],
+                              column_split = anno.archetypes.hinte %>% 
+                                levels() %>%
+                                {c("C1", rep(.[1], 4), "C2", rep(.[2], 4), "C3", rep(.[3], 4))} %>%
+                                factor(., levels = unique(.)),
+                              column_gap = unit(rep(0:1, 3), "mm") %>% .[-length(.)],
+                              border_gp = gpar(col = "black", lwd = 1.5))
+
+ht = draw(ha)
+```
+
+## Extended Data Figure 8k
+
+```{r}
+genes.order <- read.delim("Archetypes_DEG_heatmap_genes.tsv") %$% 
+  setNames(gene, cluster %>% factor(labels = seq(4, 1, -1)) %>% as.character()) %>% 
+  .[order(names(.))] %>%
+  {split(unname(.), names(.))}
+
+cts <- anno.archetypes.hinte %>% 
+  levels()
+
+anno <- getConosCluster(con.integration, "loft_transfer_minor")
+anno.subtype <- anno[names(anno) %in% names(anno.archetypes.hinte)] %>% 
+  factor()
+
+cm.merged.sub <- cm.merged.raw %>% 
+  .[rownames(.) %in% names(anno.subtype), ]
+
+cm.clean <- cm.merged.sub[rownames(cm.merged.sub) %in% names(anno.subtype), ]
+
+cluster.scores <- genes.order %>% 
+  lapply(\(x) {
+    cls.scores <- list(cluster.scores = x)
+    seu_obj <- CreateSeuratObject(counts = cm.clean %>% Matrix::t() %>% drop0(),
+                                  min.cells = 0, 
+                                  min.features = 0)
+    m.scores <- UCell::AddModuleScore_UCell(seu_obj, features = cls.scores, ncores = 32)
+
+    return(m.scores)
+  })
+```
+
+Prepare
+
+```{r}
+cm.sub <- cluster.scores %>% 
+  lget("cluster.scores_UCell") %>% 
+  lapply(\(x) mutate(x, cid = rownames(x))) %>% 
+  bind_rows(.id = "cluster")
+
+# Create sample-wise annotation
+anno.donor <- con.integration$getDatasetPerCell()[unique(cm.sub$cid)]
+anno.subtype <- anno.archetypes.hinte %>% 
+  .[!is.na(.)] %>% 
+  factor()
+
+idx <- intersect(anno.donor %>% names(), anno.subtype %>% names())
+
+anno.donor %<>% .[idx]
+anno.subtype %<>% .[idx] %>% 
+  {`names<-`(as.character(.), names(.))}
+
+anno.final <- paste0(anno.donor,"!!",anno.subtype) %>% 
+  `names<-`(anno.donor %>% names())
+
+# Prepare plot
+plot.tmp <- cm.sub %>% 
+  dplyr::rename(value = cluster.scores_UCell) %>% 
+  mutate(donor_anno = anno.final[cid]) %>% 
+  mutate(donor = strsplit(donor_anno, "!!") %>% 
+           sget(1),
+         anno = strsplit(donor_anno, "!!") %>% 
+           sget(2)) %>% 
+  mutate(visit = strsplit(donor, "_") %>% 
+           sapply(\(x) if (length(x) == 2) x[2] else x[1])) %>%
+  group_by(donor_anno, donor, anno, visit, cluster) %>% 
+  summarize(value = median(value))
+```
+
+```{r, fig.width=7, fig.height=4}
+plot.tmp %>% 
+  filter(visit %in% c("lean", "t0", "t1")) %>% 
+  ggplot(aes(anno, value, fill = visit)) +
+  # geom_boxplot() +
+  facet_grid(~cluster) +
+  stat_summary(geom = "bar", fun = mean, position = "dodge") +
+  geom_point(aes(col = visit), position = position_dodge(width = 0.9)) +
+  theme_bw() + 
+  theme(axis.text = element_text(color = "black"),
+        line = element_blank(),
+        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
+  labs(x = "", y = "Mean UCell module score per sample", fill = "", title = "", col = "") +
+  scale_fill_manual(values = brewer.pal(9, "Greens")[c(3, 6, 9)]) +
+  scale_color_manual(values = brewer.pal(9, "Purples")[c(9, 7, 5)])
+```
+
+## Extended Data Figure 8l
+
+```{r, fig.width=5, fig.height=4}
+gene <- "PLIN5"
+
+plot.dat <- cm.pseudo %>% 
+  t() %>% 
+  .[, colnames(.) %in% gene] %>% 
+  as.data.frame() %>% 
+  tibble::rownames_to_column(var = "sample") %>% 
+  mutate(visit = strsplit(sample, "!!") %>% 
+           sget(1) %>% 
+           strsplit("_") %>% 
+           sapply(\(x) if (length(x) == 1) x[1] else x[2]) %>% 
+           factor(levels = c("lean", "t0", "t1")),
+         anno = strsplit(sample, "!!") %>% 
+           sget(2) %>% 
+           factor(levels = c("DGAT2_archetype", "CLSTN2_archetype", "PRSS23_archetype"))) %>%
+  reshape2::melt(id.vars = c("sample", "visit", "anno"))
+
+plot.dat %>% 
+  ggplot(aes(anno, value, fill = visit)) + 
+  stat_summary(fun = "mean", geom = "bar", position = position_dodge()) +
+  geom_point(aes(col = visit),
+             position = position_dodge(width = 0.9)) +
+  theme_bw() +
+  labs(x = "", y = "Normalized pseudobulk expression", title = gene, fill = "", col = "") +
+  theme(line = element_blank(),
+        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
+        axis.text = element_text(color = "black")) +
+  scale_fill_manual(values = c(brewer.pal(9, "Greens")[c(3,6,9)])) +
+  scale_color_manual(values = c(brewer.pal(9, "Purples")[c(9,7,5)]))
+```
+
+```{r}
+gene <- "ADAMTS10"
+
+plot.dat <- cm.pseudo %>% 
+  t() %>% 
+  .[, colnames(.) %in% gene] %>% 
+  as.data.frame() %>% 
+  tibble::rownames_to_column(var = "sample") %>% 
+  mutate(visit = strsplit(sample, "!!") %>% 
+           sget(1) %>% 
+           strsplit("_") %>% 
+           sapply(\(x) if (length(x) == 1) x[1] else x[2]) %>% 
+           factor(levels = c("lean", "t0", "t1")),
+         anno = strsplit(sample, "!!") %>% 
+           sget(2) %>% 
+           factor(levels = c("DGAT2_archetype", "CLSTN2_archetype", "PRSS23_archetype"))) %>%
+  reshape2::melt(id.vars = c("sample", "visit", "anno"))
+
+plot.dat %>% 
+  ggplot(aes(anno, value, fill = visit)) + 
+  stat_summary(fun = "mean", geom = "bar", position = position_dodge()) +
+  geom_point(aes(col = visit),
+             position = position_dodge(width = 0.9)) +
+  theme_bw() +
+  labs(x = "", y = "Normalized pseudobulk expression", title = gene, fill = "", col = "") +
+  theme(line = element_blank(),
+        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
+        axis.text = element_text(color = "black")) +
+  scale_fill_manual(values = c(brewer.pal(9, "Greens")[c(3,6,9)])) +
+  scale_color_manual(values = c(brewer.pal(9, "Purples")[c(9,7,5)]))
+```
+
+```{r}
+gene <- "MAP3K5"
+
+plot.dat <- cm.pseudo %>% 
+  t() %>% 
+  .[, colnames(.) %in% gene] %>% 
+  as.data.frame() %>% 
+  tibble::rownames_to_column(var = "sample") %>% 
+  mutate(visit = strsplit(sample, "!!") %>% 
+           sget(1) %>% 
+           strsplit("_") %>% 
+           sapply(\(x) if (length(x) == 1) x[1] else x[2]) %>% 
+           factor(levels = c("lean", "t0", "t1")),
+         anno = strsplit(sample, "!!") %>% 
+           sget(2) %>% 
+           factor(levels = c("DGAT2_archetype", "CLSTN2_archetype", "PRSS23_archetype"))) %>%
+  reshape2::melt(id.vars = c("sample", "visit", "anno"))
+
+plot.dat %>% 
+  ggplot(aes(anno, value, fill = visit)) + 
+  stat_summary(fun = "mean", geom = "bar", position = position_dodge()) +
+  geom_point(aes(col = visit),
+             position = position_dodge(width = 0.9)) +
+  theme_bw() +
+  labs(x = "", y = "Normalized pseudobulk expression", title = gene, fill = "", col = "") +
+  theme(line = element_blank(),
+        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
+        axis.text = element_text(color = "black")) +
+  scale_fill_manual(values = c(brewer.pal(9, "Greens")[c(3,6,9)])) +
+  scale_color_manual(values = c(brewer.pal(9, "Purples")[c(9,7,5)]))
+```
+
+# Extended Data Figure 9
+
+## Extended Data Figures 9a-c
 
@@
 
-### Figure S6A,B
+### Extended Data Figures 9a,b
 
@@
 
-### Figure S6C
+### Extended Data Figure 9c
 
@@
 
-## Figure S6D
+## Extended Data Figure 9d
 
@@
 
+## Extended Data Figure 9e
+
+First, we need to integrate samples, calculate new diffusion map, and calculate new trajectory before plotting using the same genes identified for Loft cells only.
+These initial steps are not run here, load data below.
+
+Create new Conos object
+
+```{r, eval = F}
+con.new = qread("con_adipocytes_aspc.qs", nthreads = 10)
+anno <- getConosCluster(con.integration, "loft_transfer") %>% 
+  .[. %in% c("Adipocytes", "FAPs")] %>% 
+  gsub("FAPs", "ASPCs", .)
+
+anno.hinte <- anno %>% 
+  .[!grepl("!!", names(.))]
+
+samples.processed <- con.integration$samples %>% 
+  .[!grepl("vis", names(.))] %>% 
+  lapply(\(sid) {
+    conos:::getRawCountMatrix(sid)
+  }) %>% 
+  lapply(basicP2proc, n.cores = 32, get.largevis = F, get.tsne = F, make.geneknn = F)
+
+con.new$addSamples(samples.processed)
+con.new$buildGraph()
+con.new$findCommunities()
+con.new$embedGraph()
+con.new$embedGraph(method = "UMAP")
+```
+
+Calculate diffusion map
+
+```{r, eval = F}
+data <- con.new
+
+## Convert raw counts to Seurat
+for (name in names(data$samples)) {
+  seu <- CreateSeuratObject(CreateAssayObject(counts = t(data$samples[[name]]$misc$rawCounts)))
+  seu$orig.ident <- substr(name, 0, regexpr("vis", name)-2)
+  seu$visit <- substr(name, regexpr("vis", name), nchar(name))
+  if (name == names(data$samples)[1]) {
+    final <- seu
+  } else {
+    final <- merge(final, seu)
+  }
+}
+
+## Merge annotation
+final <- subset(final, cells = names(anno))
+final <- NormalizeData(final)
+anno.final <- anno[ match(colnames(final), names(anno))]
+final$fine <- anno.final
+final$coarse <- anno.final
+final$dataset <- paste(final$orig.ident, final$visit, sep="_")
+final <- subset(final, cells = names(which(!is.na(final$fine))))
+full <- final
+
+embeddings <- list()
+for (nfeats in c(100)) { 
+  # Recreate original object
+  final <- full
+  # ASPC
+  ASPC <- subset(final, coarse == "ASPCs")
+  VariableFeatures(ASPC) <- NULL
+  obj.list <- SplitObject(ASPC, split.by = "dataset")
+  feats_ASPC <- suppressWarnings(SelectIntegrationFeatures(obj.list, nfeatures = nfeats, verbose = FALSE))
+  # Adipocytes
+  Ad <- subset(final, coarse == "Adipocytes")
+  VariableFeatures(Ad) <- NULL
+  obj.list <- SplitObject(Ad, split.by = "dataset")
+  feats_Ad <- suppressWarnings(SelectIntegrationFeatures(obj.list, nfeatures = nfeats, verbose = FALSE))
+  # Combine
+  feats <- unique(c(feats_ASPC, feats_Ad))
+  ## Embed
+  # PCA
+  VariableFeatures(final) <- feats
+  final <- ScaleData(final)
+  final <- RunPCA(final)
+  for (useHarmony in c(TRUE)) {
+    if (useHarmony) {
+      final <- suppressWarnings(RunHarmony(final, group.by.vars = "dataset", verbose  = FALSE))
+      pcs <- final@reductions$harmony@cell.embeddings
+    } else {
+      pcs <- final@reductions$pca@cell.embeddings
+    }
+    for (dims in c(10)) {
+      dm <- destiny::DiffusionMap(pcs[,1:dims], verbose = FALSE, n_pcs = NA)
+      dm_ev <- dm@eigenvectors
+      colnames(dm_ev) <- paste("DC", c(1:20), sep="_")
+      embeddings[[(length(embeddings) + 1)]] <- dm_ev
+      names(embeddings)[length(embeddings)] <- paste(nfeats, useHarmony, dims, sep="_")
+    }
+  }
+}
+
+# Save embedding
+qsave(embeddings[[1]], "diffusion_embedding.qs")
+```
+
+Create trajectory
+
+```{r, eval = F}
+# Calculate slingshot trajectory
+emb <- qread("diffusion_embedding.qs")
+anno.minor <- getConosCluster(con.integration, "loft_transfer_minor")
+anno.sort <- anno.minor[rownames(emb)] %>%
+  .[!. %in% c("ATM", "cvEC", "Mast cells", "mono/mac", "Pericytes")]
+
+sds_obj <- slingshot(emb[names(anno.sort), 1:2] %>%
+                       `colnames<-`(c("UMAP1","UMAP2")),
+                     anno.sort %>% unname(),
+                     end.clus = "Adipocytes",
+                     start.clus = "ASPC_DPP4")
+
+qsave(sds_obj, "sds_obj_hinte.qs")
+```
+
+Load data
+
+```{r}
+emb <- qread("diffusion_embedding.qs")
+sds_obj <- qread("sds_obj_hinte.qs")
+sds <- as.SlingshotDataSet(sds_obj)
+res <- qread("pseudotime_res_novis2.qs", nthreads = 10) %>% .[!sapply(., is.null)]
+
+# Here, we paste the order of genes so we don't have to calculate it
+pseudoGeneOrder <- c("GPC3", "ITM2B", "CXCL14", "CXCL12", "FOS", "CD81", "B2M", "TMSB10", "SERPING1", "CLU", "CRIP1", "S100A4", "LUM", "ACTB", "S100A10", "LMNA", "C1S", "C3", "CD63", "C1R", "CST3", "IGFBP6", "PLAC9", "AEBP1", "S100A6", "MMP2", "TIMP2", "FYN", "SDC2", "UST", "FLRT2", "MEG8", "SMURF2", "AL110292.1", "PKD2", "MYO9A", "PALLD", "FTO", "MSC-AS1", "LDLRAD4", "TTC3", "MAP3K20", "FBXO11", "DIP2C", "ROCK2", "UPP2", "LINC02456", "PDXDC1", "ADIPOQ-AS1", "IGF1", "PLEKHH2", "C2CD2", "TBL1XR1", "SLC1A3", "C6", "MAP3K5", "RGS6", "LENG8")
+```
+
+Get genes
+
+```{r}
+both <- res %>%
+  sapply(\(gene) {
+    ann <- gene$annova
+    (ann[2, "Pr(>Chisq)"] <= 0.05 & ann[3, "Pr(>Chisq)"] <= 0.05)
+  })
+
+res.both <- res[both]
+
+p.both <- res.both %>% 
+  sapply(\(gene) gene$annova[3, "Pr(>Chisq)"])
+
+tmp <- res.both %>% 
+  lget("annova") %>% 
+  lapply(dplyr::slice, 2:3) %>% 
+  lapply(pull, AIC) %>%
+  .[sapply(., \(x) abs(x[2]) > abs(x[1]))] %>% 
+  bind_rows() %>% 
+  t() %>% 
+  as.data.frame() %>% 
+  setNames(c("reduced", "full")) %>% 
+  mutate(diff = abs(full) - abs(reduced)) %>% 
+  mutate(diff.frac = diff / abs(reduced))
+
+res.filter <- res[
+  tmp %>% 
+    filter(diff.frac >= 0.05) %>% 
+    rownames()
+]
+```
+
+```{r}
+anno <- getConosCluster(con.integration, "loft_transfer") %>% 
+  .[. %in% c("Adipocytes", "FAPs")] %>% 
+  gsub("FAPs", "ASPCs", .)
+
+anno.hinte <- anno %>% 
+  .[!grepl("!!", names(.))] %>% 
+  factor()
+
+pseudotime_hinte <- sds_obj@assays@data@listData$pseudotime[, 1] %>% 
+  .[!is.na(.)] %>% 
+  .[names(anno.hinte)[!anno.hinte %in% c("ATM", "cvEC", "Mast cells", "mono/mac", "Pericytes")]] %>% 
+  .[!is.na(.)]
+
+cm.merged.tmp <- con.integration$getJointCountMatrix() %>% 
+  .[match(names(pseudotime_hinte), rownames(.)), colnames(.) %in% names(res.filter)]
+
+visit = con.integration$getDatasetPerCell()[rownames(cm.merged.tmp)] %>%
+  as.character() %>%
+  .[!grepl("!!", .)] %>% 
+  strsplit("_") %>%
+  sapply(\(x) if (length(x) == 1) x else x[2])
+
+# t1
+cm.merged.vis <- cm.merged.tmp[visit == "t1", ]
+pseudotime.vis <- pseudotime_hinte[rownames(cm.merged.vis)]
+weights.vis <- sds_obj@assays@data@listData$weights %>% .[match(names(pseudotime.vis), rownames(.)), 1] + 1E-7
+
+scFit <- cm.merged.vis %>% 
+  Matrix::t() %>% 
+  tradeSeq::fitGAM(pseudotime = pseudotime.vis, cellWeights = weights.vis, verbose = TRUE)
+
+Smooth3 <- tradeSeq::predictSmooth(scFit, gene = colnames(cm.merged.vis), tidy = FALSE, n=100)
+
+# t0
+cm.merged.vis <- cm.merged.tmp[visit == "t0", ]
+pseudotime.vis <- pseudotime_hinte[rownames(cm.merged.vis)]
+weights.vis <- sds_obj@assays@data@listData$weights %>% .[match(names(pseudotime.vis), rownames(.)), 1] + 1E-7
+
+scFit <- cm.merged.vis %>% 
+  Matrix::t() %>% 
+  tradeSeq::fitGAM(pseudotime = pseudotime.vis, cellWeights = weights.vis, verbose = TRUE)
+
+Smooth1 <- tradeSeq::predictSmooth(scFit, gene = colnames(cm.merged.vis), tidy = FALSE, n=100)
+
+# lean
+cm.merged.vis <- cm.merged.tmp[visit == "lean", ]
+pseudotime.vis <- pseudotime_hinte[rownames(cm.merged.vis)]
+weights.vis <- sds_obj@assays@data@listData$weights %>% .[match(names(pseudotime.vis), rownames(.)), 1] + 1E-7
+
+scFit <- cm.merged.vis %>% 
+  Matrix::t() %>% 
+  tradeSeq::fitGAM(pseudotime = pseudotime.vis, cellWeights = weights.vis, verbose = TRUE)
+
+Smoothl <- tradeSeq::predictSmooth(scFit, gene = colnames(cm.merged.vis), tidy = FALSE, n=100)
+
+# Combine smooth
+Smooth <- cbind(Smooth3, Smooth1, Smoothl)
+
+# Average across replicates and scale
+Smooth <- t(scale(t(Smooth)))
+
+Smooth %<>% 
+  .[pseudoGeneOrder, ]
+
+# Split smooth
+Smooth33 <- Smooth[, 1:100]
+Smooth11 <- Smooth[, 101:200]
+Smoothll <- Smooth[, 201:300]
+
+# t0
+# Seriate the results
+
+col_fun = circlize::colorRamp2(c(-4, 0, 4), c("navy", "white", "firebrick"))
+
+p3 <- Heatmap(Smooth33, 
+              col = col_fun,
+              cluster_columns=F, 
+              cluster_rows=F, 
+              show_column_names = F, 
+              show_row_names = F,
+              show_heatmap_legend = T
+)
+
+# t0
+p1 <- Heatmap(Smooth11, 
+              col = col_fun,
+              cluster_columns=F, 
+              cluster_rows=F, 
+              show_column_names = F, 
+              show_row_names = F,
+              show_heatmap_legend = F
+)
+
+# lean
+gene.highlights <- c("CXCL14", "CXCL12", "CD81", "SERPING1", "CLU", "C1S", "C3", "C1R", "MMP2", "TIMP2", "FTO", "MAP3K20", "PLEKHH2", "TBL1XR1", "C6", "MAP3K5") %>% 
+  {which(rownames(Smoothll) %in% .)}
+
+row_anno <- rowAnnotation(foo = anno_mark(at = gene.highlights, 
+                                          labels = rownames(Smoothll)[gene.highlights],
+                                          side = "left"))
+
+pl <- Heatmap(Smoothll, 
+              col = col_fun,
+              left_annotation = row_anno,
+              cluster_columns=F, 
+              cluster_rows=F, 
+              show_column_names = F, 
+              show_row_names = F,
+              show_heatmap_legend = F
+)
+
+pl
+p1
+p3
+```
+
+## Extended Data Figure 9f
+
+```{r}
+tmp.regulons <- qread("regulons.qs")
+
+pseudotime_all <- sds_obj@assays@data@listData$pseudotime %>% 
+  as.data.frame() %>% 
+  filter(!is.na(Lineage1)) %>% 
+  {setNames(pull(., Lineage1), rownames(.))} %>% 
+  .[names(tmp.regulons[[1]])]
+
+spc <- con.integration$getDatasetPerCell()
+```
+
+```{r}
+tmp.regulons$PPARG %>% 
+  {data.frame(value = unname(.), row.names = names(.))} %>% 
+  mutate(sample = spc[match(rownames(.), names(spc))] %>% unname()) %>%
+  mutate(pseudo = pseudotime_all) %>%
+  mutate(bin = cut(pseudo, breaks = c(0, 0.01, 0.0165, 0.02, 0.026))) %>% 
+  na.omit() %>% 
+  mutate(bin = as.character(bin) %>% 
+           strsplit(",") %>% 
+           sget(2) %>% 
+           gsub("]", "", ., fixed = TRUE)) %>%
+  mutate(visit = as.character(sample) %>% 
+           strsplit("_") %>% 
+           sapply(\(x) if (length(x) == 3) x[3] else if (length(x) == 2) x[2] else x),
+         sex = meta$sex[match(sample, meta$sample)],
+         study = ifelse(grepl("!!", rownames(.)), "Loft", "Hinte")) %>%
+  mutate(sex_visit = paste(sex, visit, sep = "_") %>% 
+           gsub("NA_", "", .) %>% 
+           factor() %>% 
+           factor(levels = c(levels(.)[c(1:3, 5:9, 4)]))) %>%
+  arrange(bin) %>% 
+  mutate(bin = factor(bin, labels = c("Early preadipocytes", "Mid preadipocytes", "Late preadipocytes", "Adipocyte"))) %>% 
+  filter(study == "Hinte") %>%
+  group_by(bin, sex_visit, study, sample) %>%
+  summarize(mean_auc = mean(value)) %>%
+  mutate(sex_visit = factor(sex_visit, levels = c("lean", "t0", "t1"))) %>% 
+  group_by(bin) %>% # Addition
+  mutate(mean_auc = mean_auc - mean(mean_auc[sex_visit == "t0"])) %>% # Addition
+  filter(!sex_visit == "t0") %>% # Addition
+  ggplot(aes(bin, mean_auc)) +
+  stat_summary(aes(fill = sex_visit), fun = "mean", position = position_dodge(width = 0.8), geom = "bar") +
+  geom_point(aes(col = sex_visit), position = position_dodge(width = 0.8)) + 
+  theme_bw() +
+  theme(line = element_blank(),
+        axis.text = element_text(color = "black"),
+        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
+  scale_fill_manual(values = c(brewer.pal(9, "Greys")[c(3,6,9)])) +
+  scale_color_manual(values = c(brewer.pal(9, "Purples")[c(9,7,5)])) +
+  labs(title = "PPARG regulon activity", y = "Change from obese (t0) mean AUC", x = "", fill = "", col = "")
+```
+
+# Extended Data Figure 10
+
+These plots are presented in `Liana.ipynb`.
+
 # Session info

### 2025-07-10 Manuscript_figures.Rmd
Updated for revision

@@
 
-c("PLXDC2","CBLB","TCF4",
-  "GPAM","ACSL1","LPL","DGAT2",
-  "PRSS23","THSD7A","AL139317.5") %>% 
+# c("PLXDC2","CBLB","TCF4",
+#   "GPAM","ACSL1","LPL","DGAT2",
+#   "PRSS23","THSD7A","AL139317.5") %>% 
+c("PRSS23","PDE5A","ADIPOQ", "DGAT2", "GPAM", "LPL", "RORA", "CLSTN2", "CRIM1", "ITGA1") %>% 
   sccore::dotPlot(., 
@@
 
-Load data. We need the `sds_obj` from Figure 6A. Also, for ease of use we provide slingshot pseudotime W/O visit 2 as `pseudotime_novis2.qs`.
+Load data. We need the `sds_obj` from Figure 6a. Also, for ease of use we provide slingshot pseudotime W/O visit 2 as `pseudotime_novis2.qs`.
 
@@
 ```{r, fig.width = 6, fig.height = 4}
-tmp <- con.integration$clusters$loft_transfer$groups %>%
-  renameAnnotation("Endothelial cells", "ECs") %>%
-  renameAnnotation("FAPs", "ASPCs") %>% 
+tmp <- loft.transfer$labels %>%
   renameAnnotation("Lymphoid immune cells", "Lymphoid ICs") %>%
-  renameAnnotation(c("Tissue-resident macrophages", "Myeloid immune cells"), "Myeloid ICs") %>% 
+  renameAnnotation(c("Myeloid immune cells"), "Myeloid ICs") %>% 
   {data.frame(cid = names(.), anno = unname(.))} %>% 
@@
   mutate(anno = factor(anno) %>% 
-           factor(levels = levels(.)[c(1, 2, 5, 8, 3, 7, 6, 4)]) %>% 
-           renameAnnotation("Lymphatic endothelial cells", "LECs"))
+           factor(levels = levels(.)[c(1, 2, 5, 8, 3, 7, 6, 4, 9)]) %>% 
+           renameAnnotation("Lymphatic ECs", "LECs"))
 
@@
 ## Extended Data Figure 3g
+
+We integrated our data with the data from [Hinte *et al.*](https://www.nature.com/articles/s41586-024-08165-7) that were graciously provided by the authors.
+As these data are not public, here we will only show how we plotted the data.
 
@@
 
+```{r, fig.width=4, fig.height=4}
+# Use Cacoa to calculate proportions
+cao <- Cacoa$new(data.object = con.major, 
+                 cell.groups = anno.major, 
+                 ref.level = "vis2", 
+                 target.level = "vis3", 
+                 sample.groups = con.major$samples %>% 
+                   names() %>% 
+                   strsplit("_") %>% 
+                   sapply('[[', 3) %>% 
+                   gsub(1, 2, .) %>% 
+                   `names<-`(con.major$samples %>% 
+                               names()),
+                 sample.groups.palette = ggsci::pal_jama()(7))
+
+cao$sample.groups <- con.major$samples %>% 
+  names() %>% 
+  strsplit("_") %>% 
+  sapply('[[', 3) %>% 
+  `names<-`(con.major$samples %>% 
+              names()) %>% 
+  gsub("vis", "Visit ", .)
+
+anno.comb <- anno.major[!names(anno.major) %in% names(anno.immune)] %>% 
+  {factor(c(., anno.immune))}
+
+# Get proportions
+proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
+  {100 * ./rowSums(.)} %>% 
+  as.data.frame() %$% 
+  setNames(`Early LAM`, rownames(.))
+
+tmp <- qread("eLAM_model.qs")
+
+lasso_model <- tmp$lasso_model
+best_lambda <- tmp$lasso_cv$lambda.min
+
+cm.bulk.mat <- cm.bulk %>% 
+  as.data.frame() %>% 
+  filter(rownames(.) %in% rownames(coef(lasso_model, s = best_lambda))) %>% 
+  as.matrix() %>% 
+  t()
+
+# Now, train a final model on all data for visualization and gene selection
+cm.test <- cm.bulk.mat %>% 
+  .[rownames(.) %in% names(proportions), ]
+
+lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 40)
+best_lambda <- lasso_cv$lambda.min
+lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)
+
+# Predict on all samples using the final model
+lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)
+r2.all <- cor(proportions, lasso.predict.all)^2
+
+p <- data.frame(Proportions = proportions,
+           Predictions = unname(lasso.predict.all)) %>% 
+  ggplot(aes(Proportions, Predictions)) + 
+  geom_point() +
+  theme_bw() +
+  theme(axis.text = element_text(colour = "black")) +
+  labs(title = paste0("r2 = ", formatC(r2.all, digits = 3)), y = "Predicted proportions (%)", x = "Proportions (%)") +
+  geom_abline(intercept = 0, slope = 1, colour = "firebrick")
+
+p
+```
+
+### Extended Data Figure 5h
+
+See Figure 3e for model setup
+
 ```{r}
@@
   as.data.frame() %$% 
-  setNames(`Early LAM`, rownames(.))
-
-tmp <- qread("eLAM_model.qs")
-
-lasso_model <- tmp$lasso_model
-best_lambda <- tmp$lasso_cv$lambda.min
-
-cm.bulk.mat <- cm.bulk %>% 
-  as.data.frame() %>% 
-  filter(rownames(.) %in% rownames(coef(lasso_model, s = best_lambda))) %>% 
-  as.matrix() %>% 
-  t()
-
-# Now, train a final model on all data for visualization and gene selection
-cm.test <- cm.bulk.mat %>% 
-  .[rownames(.) %in% names(proportions), ]
-
-lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 40)
-best_lambda <- lasso_cv$lambda.min
-lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)
-
-# Predict on all samples using the final model
-lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)
-r2.all <- cor(proportions, lasso.predict.all)^2
-
-data.frame(Proportions = proportions,
-           Predictions = unname(lasso.predict.all)) %>% 
-  ggplot(aes(Proportions, Predictions)) + 
-  geom_point() +
-  theme_bw() +
-  theme(axis.text = element_text(colour = "black")) +
-  labs(title = paste0("r2 = ", formatC(r2.all, digits = 3)), y = "Predicted proportions (%)", x = "Proportions (%)") +
-  geom_abline(intercept = 0, slope = 1, colour = "firebrick")
-```
-
-### Extended Data Figure 5h
-
-See Figure 3e for model setup
-
-```{r}
-# Use Cacoa to calculate proportions
-cao <- Cacoa$new(data.object = con.major, 
-                 cell.groups = anno.major, 
-                 ref.level = "vis2", 
-                 target.level = "vis3", 
-                 sample.groups = con.major$samples %>% 
-                   names() %>% 
-                   strsplit("_") %>% 
-                   sapply('[[', 3) %>% 
-                   gsub(1, 2, .) %>% 
-                   `names<-`(con.major$samples %>% 
-                               names()),
-                 sample.groups.palette = ggsci::pal_jama()(7))
-
-cao$sample.groups <- con.major$samples %>% 
-  names() %>% 
-  strsplit("_") %>% 
-  sapply('[[', 3) %>% 
-  `names<-`(con.major$samples %>% 
-              names()) %>% 
-  gsub("vis", "Visit ", .)
-
-anno.comb <- anno.major[!names(anno.major) %in% names(anno.immune)] %>% 
-  {factor(c(., anno.immune))}
-
-# Get proportions
-proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
-  {100 * ./rowSums(.)} %>% 
-  as.data.frame() %$% 
   setNames(LAM, rownames(.))
@@
 ## Extended Data Figure 5j
+
+We integrated our data with the data from [Hinte *et al.*](https://www.nature.com/articles/s41586-024-08165-7) that were graciously provided by the authors.
+As these data are not public, here we will only show how we plotted the data.
 
@@
 ## Extended Data Figure 5m
+
+We integrated our data with the data from [Hinte *et al.*](https://www.nature.com/articles/s41586-024-08165-7) that were graciously provided by the authors.
+As these data are not public, here we will only show how we plotted the data.
 
@@
 con <- qread("con_adipocytes.qs", nthreads = 10)
-
+anno.adipocytes <- qread("anno_adipocytes_archetypes.qs")
 spc <- con$getDatasetPerCell()
@@
 ```{r s5f}
-con$findCommunities(resolution = 1, test.stability = T, name = "l05stab")
+con$findCommunities(resolution = 1, test.stability = T, name = "l1stab")
 
@@
     labs(x = " \n "),
-  con$plotClusterStability("l05stab", "hjc") + 
+  con$plotClusterStability("l1stab", "hjc") + 
     theme_bw() +
@@
 
-```{r s5g}
+```{r s5g, fig.width=6, fig.height=4}
 cm.merged <- con$getJointCountMatrix()
 
-anno.adipocytes <- qread("anno_adipocytes_archetypes.qs")
-
-c("PRSS23","PDE5A","ADIPOQ","DGAT2", "GPAM", "LPL", "DCN", "VIM", "RORA", "CLSTN2", "CRIM1", "ITGA1") %>% 
+c("PRSS23","PDE5A","ADIPOQ", "DGAT2", "GPAM", "LPL", "RORA", "CLSTN2", "CRIM1", "ITGA1") %>% 
   sccore::dotPlot(., 
                   cm.merged, 
-                  anno.adipocytes, 
+                  getConosCluster(con, "leiden0.5"), 
                   gene.order = ., 
@@
 
-First, we integrate adipocytes from both studies. This is not run here
+First, we integrate adipocytes from both studies. This is not run here, and the data are not shared.
 
@@
 
+We integrated our data with the data from [Hinte *et al.*](https://www.nature.com/articles/s41586-024-08165-7) that were graciously provided by the authors.
+As these data are not public, here we will only show how we plotted the data.
+
 ```{r}
@@
 
-Load data. For ease, we also provide pseudotime W/O visit 2 nuclei. We need the `sds_obj` from Figure 6A. Also, for ease of use we provide slingshot pseudotime W/O visit 2 as `pseudotime_novis2.qs`.
+Load data. For ease, we also provide pseudotime W/O visit 2 nuclei. We need the `sds_obj` from Figure 6a. Also, for ease of use we provide slingshot pseudotime W/O visit 2 as `pseudotime_novis2.qs`.
 
@@
 
-We need the `sds_obj` object from Figure 6A.
+We need the `sds_obj` object from Figure 6a.
 
